---
title: "CellRank and Palantir fate workflow"
output: rmarkdown::html_vignette
vignette: >
  %\\VignetteIndexEntry{CellRank and Palantir fate workflow}
  %\\VignetteEngine{knitr::rmarkdown}
---

This workflow keeps the Palantir pseudotime direction and the CellRank fate
model as separate, inspectable result layers. Pseudotime is a relative ordering
and fate probabilities are model-based transition probabilities; neither is a
direct experimental time course.

```{r setup, eval=FALSE}
library(scop)
data(pancreas_sub)
pancreas_sub <- RunStandardWorkflow(pancreas_sub)
```

## Palantir direction and MAGIC layer

The Python reference backend reuses the SCOP environment selected by
`PrepareEnv()`. Existing Python 3.10/3.11 environments receive the compatible
legacy pins, while Python 3.12 environments receive the current pins. MAGIC is
optional and should be used for visualization or trend fitting, never as a
replacement for raw counts in differential testing.

```{r palantir, eval=FALSE}
pancreas_sub <- RunPalantir(
  pancreas_sub,
  group.by = "SubCellType",
  linear_reduction = "PCA",
  nonlinear_reduction = "UMAP",
  early_group = "Ductal",
  terminal_groups = c("Alpha", "Beta", "Delta", "Epsilon"),
  magic_impute = TRUE,
  backend = "python"
)
FeatureDimPlot(
  pancreas_sub,
  features = c("palantir_pseudotime", "palantir_diff_potential")
)
PalantirTrajectoryPlot(pancreas_sub, reduction = "UMAP")
```

## CellRank fate mapping

Use the stored Palantir pseudotime to construct a directed `PseudotimeKernel`.
The explicit Schur size makes the small tutorial run reproducible. The computed
macrostate names are printed before any optional manual terminal-state rerun;
unknown terminal states stop before the Seurat object is mutated.

```{r cellrank, eval=FALSE}
pancreas_sub <- RunCellRank(
  pancreas_sub,
  group.by = "SubCellType",
  kernel_type = "pseudotime",
  time_key = "palantir_pseudotime",
  # `brandts` is portable for small tutorial data; use `krylov` only after
  # installing PETSc/SLEPc in the selected environment.
  schur_method = "brandts",
  schur_n_components = 10L,
  n_macrostates = 8L,
  n_cells_terminal = 10L,
  backend = "python"
)
names(pancreas_sub@tools$CellRank)
pancreas_sub@tools$CellRank$states$macrostate_names
CellRankPlot(pancreas_sub, plot_type = "fate", reduction = "UMAP")
CellRankPlot(pancreas_sub, plot_type = "circular")
```

## Drivers, trends and enrichment

Drivers are statistical associations with fate probabilities. Trend modules are
similar expression-shape clusters, not causal regulatory modules.

```{r trends, eval=FALSE}
pancreas_sub <- RunCellRankTrends(
  pancreas_sub,
  lineage = "Alpha",
  assay = "RNA",
  layer = "MAGIC_imputed_data",
  top_n = 500L,
  heatmap_n = 50L,
  n_points = 200L,
  output_dir = "cellrank_objects"
)
CellRankPlot(pancreas_sub, plot_type = "drivers", lineage = "Alpha")
CellRankPlot(pancreas_sub, plot_type = "trends", lineage = "Alpha")
CellRankPlot(pancreas_sub, plot_type = "clusters", lineage = "Alpha")

pancreas_sub <- RunCellRankEnrichment(
  pancreas_sub,
  lineage = "Alpha",
  species = "Mus_musculus",
  output_dir = "cellrank_objects"
)
pancreas_sub@tools$CellRank$enrichment$Alpha$manifest
```

The complete payload is stored in `srt@tools$Palantir` and
`srt@tools$CellRank`, including the actual backend, versions, cell/lineage
names, transition matrix, fate probabilities, drivers, trend curves, modules,
enrichment tables and run parameters.
