---
title: "Cell annotation workflow"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", error = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 12, fig.height = 8, dpi = 144, out.width = "100%")
```

## What you will do

**Question:** which reference cell types resemble each query cell, and where is
that assignment uncertain? You will transfer labels, compare them with held-out
annotations, inspect marker expression and save the prediction evidence.

The example uses all 1,600 cells of the bundled human `panc8_sub` object. Every
`celseq2` cell is a query; every other technology is reference. This is a
technology-defined holdout, with no random sampling and no shared cells.
It illustrates transfer within one tissue/species, not a validated classifier
for arbitrary tissue. The original package dataset is itself a study subset.

The old human-to-mouse capitalization shortcut is intentionally avoided:
changing letter case is not ortholog mapping. For cross-species work, prepare
and report a versioned ortholog table first, resolving one-to-many mappings.

## Prepare disjoint query and reference inputs

```{r prepare-query-reference}
library(scop)
set.seed(11)
data(panc8_sub)
stopifnot(!anyNA(panc8_sub$tech), !anyNA(panc8_sub$celltype))
query_ids <- colnames(panc8_sub)[panc8_sub$tech == "celseq2"]
reference_ids <- setdiff(colnames(panc8_sub), query_ids)
stopifnot(length(intersect(query_ids, reference_ids)) == 0,
          length(query_ids) + length(reference_ids) == ncol(panc8_sub))
query <- panc8_sub[, query_ids]
reference <- panc8_sub[, reference_ids]
query <- RunStandardWorkflow(query, assay = "RNA", seed = 11, cores = 1,
                             verbose = FALSE)
reference <- RunStandardWorkflow(reference, assay = "RNA", seed = 11, cores = 1,
                                 verbose = FALSE)
query$heldout_label <- query$celltype
shared_features <- intersect(SeuratObject::VariableFeatures(query),
                             SeuratObject::VariableFeatures(reference))
stopifnot(length(shared_features) > 30)
data.frame(input = c("query", "reference"),
           cells = c(ncol(query), ncol(reference)),
           species = "Homo sapiens")
table(reference$celltype)
setdiff(unique(query$heldout_label), unique(reference$celltype))
length(shared_features)
label_colors <- stats::setNames(
  grDevices::hcl.colors(length(unique(reference$celltype)), "Dark 3"),
  sort(unique(as.character(reference$celltype)))
)
```

The final check prints query labels absent from the reference, if any. Such
states cannot be correctly named by closed-reference label transfer. Reference
labels and highly variable genes are model inputs; held-out query labels are
used only for evaluation. Neither UMAP appearance nor prediction scores prove
cell identity.

## Predict from individual reference cells

`ref_collapsing = FALSE` explicitly uses individual reference cells; the
default collapses reference groups to averages. The prefix keeps these labels
separate from the next method and from the original annotations.

```{r single-cell-reference}
query <- RunKNNPredict(
  srt_query = query, srt_ref = reference, ref_group = "celltype",
  query_assay = "RNA", ref_assay = "RNA", features = shared_features,
  ref_collapsing = FALSE, k = 30, filter_lowfreq = 0,
  prefix = "CellKNN", verbose = FALSE
)
stopifnot(identical(colnames(query), query_ids),
          "CellKNN_classification" %in% colnames(query@meta.data))
table(observed = query$heldout_label, predicted = query$CellKNN_classification)
CellDimPlot(query, group.by = c("heldout_label", "CellKNN_classification"),
            reduction = "UMAP", label = FALSE, theme_use = "theme_blank",
            palcolor = label_colors)
```

Read the contingency table row by row. A strong diagonal supports agreement
with the supplied labels; off-diagonal entries identify cells to inspect.
Agreement is descriptive here, because labels come from related pancreas data.
`filter_lowfreq = 0` preserves rare predictions so their uncertainty is visible;
it does not discard rare cells or relabel a failed transfer as correct.

## Compare with bulk-like reference profiles

These are average expression profiles derived from the same reference cells,
not an independent bulk experiment. They test the effect of reference
aggregation. `AverageExpression(..., layer = "data")` averages expression on
its unlogged scale, so we explicitly apply `log1p()` afterward to match the
query RNA `data` layer. Each column is a named reference cell type. `k = 1`
selects the nearest profile; 30 cell neighbors and one group profile are
different models, not a controlled change in only one parameter.

```{r bulk-reference}
profiles <- Seurat::AverageExpression(reference, assays = "RNA", layer = "data",
                                      group.by = "celltype", verbose = FALSE)$RNA
profiles <- log1p(profiles)
query <- RunKNNPredict(
  srt_query = query, bulk_ref = profiles,
  query_assay = "RNA", features = shared_features,
  k = 1, filter_lowfreq = 0, prefix = "ProfileKNN", verbose = FALSE
)
table(cell_reference = query$CellKNN_classification,
      profile_reference = query$ProfileKNN_classification)
names(query@tools)
CellDimPlot(query, group.by = "ProfileKNN_classification", reduction = "UMAP",
            theme_use = "theme_blank", palcolor = label_colors)
comparison <- data.frame(
  method = c("30 individual neighbors", "1 log-scale mean profile"),
  cells = ncol(query),
  matches_supplied_label = c(sum(query$heldout_label == query$CellKNN_classification),
                            sum(query$heldout_label == query$ProfileKNN_classification))
)
comparison$agreement <- comparison$matches_supplied_label / comparison$cells
comparison
```

Both prediction columns remain in the object. Different answers are a reason
to inspect neighbors and markers, not a reason to silently replace one label.
Scores are similarity/vote evidence, not calibrated clinical probabilities.

## Project onto the reference and inspect markers

Projection places query cells in the reference's coordinate system. We request
`projection_method = "knn"` explicitly: this is neighbor-weighted projection,
not a newly learned joint embedding or evidence of successful batch correction.

```{r projection}
ref_umap <- DefaultReduction(reference, pattern = "UMAP")
query <- RunKNNMap(srt_query = query, srt_ref = reference,
                   features = shared_features, ref_umap = ref_umap,
                   projection_method = "knn", verbose = FALSE)
ProjectionPlot(srt_query = query, srt_ref = reference,
                query_group = "CellKNN_classification", ref_group = "celltype",
                query_param = list(palcolor = label_colors, cells.highlight = TRUE),
                ref_param = list(palcolor = label_colors))
markers <- c("INS", "GCG", "SST", "KRT19")
data.frame(marker = markers, present = markers %in% rownames(query))
FeatureDimPlot(query, features = intersect(markers, rownames(query)),
                reduction = "UMAP", ncol = 2)
```

INS, GCG, SST and KRT19 provide marker checks for beta, alpha, delta and ductal
labels, respectively. INS is absent from this feature-filtered package input,
so the receipt records it as unavailable and the plot cannot assess that gene.
Absence from the matrix is different from measured zero expression. Check
multiple markers per state on your own data;
ambient RNA and mixed cells can create misleading single-marker signals.

```{r correlation-heatmap}
ht <- CellCorHeatmap(
  srt_query = query, srt_ref = reference,
  query_group = "heldout_label", ref_group = "celltype",
  features = shared_features, nlabel = 1, label_by = "row", label_size = 8,
  show_row_names = TRUE, show_column_names = TRUE
)
print(ht$plot)
```

This is group-level expression similarity. Similar support for several
reference labels flags ambiguity; these groups should remain under review.

## Save, reload, and check the hand-off

The build uses a temporary directory so repeated rendering cannot overwrite a
previous analysis. For your work, replace it with a new persistent output folder.
Save the object **and** plain-text metadata; a figure alone cannot reproduce the
analysis. No cells are sampled for this export.

```{r export-reload}
output_dir <- file.path(tempdir(), "cell-annotation-workflow")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
result_file <- file.path(output_dir, "analysis.rds")
saveRDS(query, result_file)
reference_file <- file.path(output_dir, "reference.rds")
saveRDS(reference, reference_file)
utils::write.csv(as.matrix(profiles), file.path(output_dir, "log-scale-reference-profiles.csv"))
writeLines(shared_features, file.path(output_dir, "shared-prediction-features.txt"))
utils::write.csv(comparison, file.path(output_dir, "prediction-agreement.csv"), row.names = FALSE)
utils::write.csv(query@meta.data, file.path(output_dir, "cell-metadata.csv"))
reloaded <- readRDS(result_file)
reloaded_reference <- readRDS(reference_file)
stopifnot(identical(colnames(reloaded_reference), reference_ids),
          identical(reloaded_reference@meta.data, reference@meta.data))
stopifnot(identical(colnames(reloaded), colnames(query)),
          identical(reloaded@meta.data, query@meta.data),
          identical(SeuratObject::Reductions(reloaded),
                    SeuratObject::Reductions(query)))
data.frame(file = basename(list.files(output_dir, full.names = TRUE)),
           bytes = file.info(list.files(output_dir, full.names = TRUE))$size)
```

## Use your own data and recover common errors

- Replace the `data(...)` input with `readRDS("your-object.rds")`, or create a
  Seurat object from a gene-by-cell raw-count matrix. Check that cell IDs are
  unique and metadata row names equal the expression column names.
- Set the assay and grouping columns explicitly. Species, gene identifiers and
  reference labels must match your experiment; do not copy pancreas labels to
  unrelated tissue.
- If a reduction or metadata column is missing, inspect
  `SeuratObject::Reductions(object)` and `colnames(object@meta.data)`, then run
  its producer chunk. A plot does not create the result it needs.
- If a dependency is unavailable, install it in the analysis environment,
  restart R and rerun from the input chunk. A skipped or failed method is not
  evidence that it worked. Preserve warnings and the exact error in your report.

## Build and analysis provenance

The receipt records the R seed. Python methods retain their documented backend
seeds and parameters; setting the R seed alone does not control every backend.
Package versions, BLAS libraries and backend changes can alter numerical results. Keep
this receipt with the exported object. It describes this page's actual render.

```{r provenance}
source_revision <- Sys.getenv("GITHUB_SHA")
if (!nzchar(source_revision) && nzchar(Sys.which("git"))) {
  source_revision <- tryCatch(system2("git", c("rev-parse", "HEAD"), stdout = TRUE,
                                      stderr = FALSE)[1], error = function(e) NA_character_)
}
article_source <- knitr::current_input(dir = TRUE)
build_receipt <- data.frame(
  article = "cell-annotation-workflow",
  built_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
  scop = as.character(utils::packageVersion("scop")),
  R = as.character(getRversion()),
  source_revision = source_revision,
  source_has_local_changes = if (nzchar(Sys.which("git")))
    length(system2("git", c("status", "--porcelain"), stdout = TRUE, stderr = FALSE)) > 0 else NA,
  article_md5 = unname(tools::md5sum(article_source)),
  R_seed = 11,
  cells_retained = ncol(query),
  genes_in_active_assay = nrow(query),
  stringsAsFactors = FALSE
)
build_receipt
utils::write.csv(build_receipt, file.path(output_dir, "build-receipt.csv"),
                 row.names = FALSE)
writeLines(capture.output(sessionInfo()), file.path(output_dir, "session-info.txt"))
```

<details>
<summary>Show the complete R session used for this page</summary>

```{r session-info}
sessionInfo()
```

</details>

## What to Report

A compact annotation report should include:

- the disjoint query/reference cells, species, feature set and any ortholog mapping;
- the reference label column and embedding used for projection;
- whether labels came from bulk-like profiles or single-cell neighbors;
- neighbor count and filtering thresholds such as `filter_lowfreq`;
- projection plots, predicted label plots, and marker-gene checks;
- ambiguous groups where several labels have similar support.

Treat annotation as an evidence layer. Keep transferred labels separate from
manual labels until marker genes and biological context support the assignment.
