---
title: "Differential expression and enrichment workflow"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  error = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 12, fig.height = 8,
  dpi = 144,
  out.width = "100%"
)
```

## What you will do

- **Question:** which genes distinguish cell states, and which pathways summarize those differences?
- **Input:** All 1,000 cells in the bundled mouse pancreas object, with `CellType` labels. Cells are observational units for this marker example, not biological replicates.
- **Output:** A complete marker table, over-representation and ranked enrichment results, diagnostic plots and reusable tables.
- **How to run:** start a fresh R session and run the chunks in order. Printed tables and figures below are generated from that run; code-only alternatives are identified explicitly.



This article shows how to organize differential expression, marker visualization,
and pathway enrichment in `scop`. The main distinction is between testing genes,
summarizing gene sets, and interpreting enriched biological terms.

Use this workflow after cell identities and major covariates are stable.
Differential expression is gene-level evidence. Enrichment is a summary of a
gene list or ranked statistic against a named database. Keep those two evidence
layers separate in the text of a report.

Before the first analysis chunk, install the extra packages in the same R
library as SCOP. Use BiocManager for `AnnotationDbi`, `org.Mm.eg.db`, `GO.db`
and `clusterProfiler`, and CRAN for `shadowtext` (used by the network plot).
These setup commands are run once, outside the analysis render:

```r
BiocManager::install(c("AnnotationDbi", "org.Mm.eg.db", "GO.db", "clusterProfiler"))
install.packages("shadowtext")
```

## Prepare the Object

```{r prepare-object}
library(scop)
set.seed(11)

data(pancreas_sub)
input_cells <- colnames(pancreas_sub)
pancreas_sub <- RunStandardWorkflow(pancreas_sub, assay = "RNA",
                                   seed = 11, cores = 1, verbose = FALSE)
stopifnot(identical(colnames(pancreas_sub), input_cells))
data.frame(cells = ncol(pancreas_sub), genes = nrow(pancreas_sub))

table(pancreas_sub$CellType)
```

## Run Differential Expression

`RunDEtest()` writes differential-expression tables under `srt@tools`. The
default single-cell marker workflow is appropriate for exploratory marker
finding. Use sample-aware or pseudobulk methods when biological replicates are
part of the question.

```{r run-detest}
pancreas_sub <- RunDEtest(
  pancreas_sub,
  group.by = "CellType",
  fc.threshold = 1,
  only.pos = FALSE,
  assay = "RNA", layer = "data", test.use = "wilcox",
  max.cells.per.ident = Inf, min.pct = 0.1,
  p.adjust.method = "bonferroni", cores = 1, seed = 11
)

de <- pancreas_sub@tools$DEtest_CellType$AllMarkers_wilcox
stopifnot(nrow(de) > 0, all(c("gene", "group1", "avg_log2FC", "p_val_adj") %in% names(de)))
data.frame(rows = nrow(de), unique_genes = length(unique(de$gene)),
           contrasts = length(unique(de$group1)))
head(de)
```

`fc.threshold = 1` is a fold-change ratio of one (log2 cutoff zero), not a
log2-fold-change cutoff of one. Both directions are retained, with no cell
subsampling. The detection filter is 10% in at least one group and adjusted
p-values use Bonferroni here. An adjusted p-value displayed as zero reflects
numerical underflow, not certainty.

Keep the contrast, grouping column, test method, and threshold in the report.
Those choices define the meaning of each marker table.
For `FindAllMarkers`-style results, `group1` names the group being compared
against the remaining cells unless a specific contrast is stated.

## Plot Differential Results

Use volcano, Manhattan, and ring views for different levels of detail. Volcano
plots are useful for one group or contrast; Manhattan and ring plots are useful
when scanning many groups.

In the first volcano view, the horizontal coordinate is the difference in
detected-cell fractions (`pct.1 - pct.2`), not log-fold change. The vertical
coordinate is signed significance: downregulated genes are drawn below zero;
its magnitude is `-log10(p_val_adj)`. Colour shows log2-fold change. The
enrichment-annotated volcano below instead uses the conventional log2-fold
change horizontal coordinate and unsigned significance.

```{r detest-plots}
DEtestPlot(
  pancreas_sub,
  group.by = "CellType",
  plot_type = "volcano",
  label.size = 2
)

DEtestPlot(
  pancreas_sub,
  group.by = "CellType",
  plot_type = "manhattan",
  label.size = 2
)

DEtestPlot(
  pancreas_sub,
  group.by = "CellType",
  plot_type = "ring",
  label.size = 2
)
```

## Prepare a tested-gene GO background

Install the mouse annotation resources in advance (`org.Mm.eg.db`, `GO.db`
and `AnnotationDbi`), `clusterProfiler` for GSEA and `shadowtext` for network labels. The native ORA backend does
not need clusterProfiler. This chunk reads real, locally installed GO Biological
Process annotations and records their versions. The common background contains
only genes returned by every DE comparison, before significance filtering.
This conservative shared universe avoids treating genes excluded from the tests
as eligible negative results. No database downloads or invented pathway
memberships occur during this step.

```{r prepare-go}
tested_by_group <- split(de$gene, de$group1)
measured_genes <- Reduce(intersect, tested_by_group)
stopifnot(length(measured_genes) > 0)
valid_symbols <- intersect(measured_genes,
                           AnnotationDbi::keys(org.Mm.eg.db::org.Mm.eg.db, keytype = "SYMBOL"))
go_map <- AnnotationDbi::select(org.Mm.eg.db::org.Mm.eg.db,
                                keys = valid_symbols, keytype = "SYMBOL",
                                columns = c("GOALL", "ONTOLOGYALL"))
go_term2gene <- unique(go_map[!is.na(go_map$GOALL) & go_map$ONTOLOGYALL == "BP",
                               c("GOALL", "SYMBOL")])
colnames(go_term2gene) <- c("Term", "symbol")
go_term2name <- AnnotationDbi::select(GO.db::GO.db,
                                      keys = unique(go_term2gene$Term),
                                      keytype = "GOID", columns = "TERM")
go_term2name <- unique(go_term2name[stats::complete.cases(go_term2name), ])
colnames(go_term2name) <- c("Term", "Name")
go_version <- paste("org.Mm.eg.db", utils::packageVersion("org.Mm.eg.db"),
                     "; GO.db", utils::packageVersion("GO.db"))
stopifnot(nrow(go_term2gene) > 0)
data.frame(database = "GO_BP", version = go_version,
           genes_tested_in_every_group = length(measured_genes),
           GO_annotated_background = length(unique(go_term2gene$symbol)),
           terms = length(unique(go_term2gene$Term)))
```

The ORA universe is the union of genes in this restricted GO mapping, not the
whole genome. Genes outside the common tested set or without GO annotation
cannot contribute. Because we supply `TERM2GENE`, SCOP stores the database key
as `custom`; its biological source remains the versioned GO_BP mapping above.
Keep that source information with the exported result.

When enrichment results are available, volcano plots can annotate enriched
terms to connect marker genes with pathway-level interpretation.
The enrichment overlay circles genes that belong to selected enriched terms.
The legend title is the analysis layer (`Enrichment`), and the legend item is
the database, here `custom` for the explicitly supplied GO_BP mapping.

```{r annotated-volcano, fig.width=14, fig.height=9}
pancreas_sub <- RunEnrichment(
  pancreas_sub,
  group.by = "CellType",
  species = "Mus_musculus",
  TERM2GENE = go_term2gene, TERM2NAME = go_term2name,
  DE_threshold = "avg_log2FC > log2(1.5) & p_val_adj < 0.05"
)

DEtestPlot(
  pancreas_sub,
  group.by = "CellType",
  plot_type = "volcano",
  threshold_method = "hyperbolic",
  hyperbola_c = 6,
  annotate_enrichment = TRUE,
  enrich_from = "Enrichment",
  enrich_db = "custom",
  enrich_top_terms = 3,
  enrich_nlabel = 5,
  label.size = 2
)
```

Only five enrichment-associated gene labels per cell group are shown to
keep the panels readable; the full marker and enrichment tables are exported.
The dashed hyperbolic guide is a display selection rule, not an additional
statistical test or the cutoff used to create the ORA input.

## Summarize Markers in Heatmaps

Feature heatmaps are useful after markers have been filtered. Keep the filter
explicit and avoid showing very large marker sets without grouping or splitting.
The complete DE table is retained. Only the ten strongest positive markers per
group are selected for this readable figure; this is a declared display choice,
not a change to the test or its cell population. All cells remain in the plot.

```{r feature-heatmap}
DEGs <- pancreas_sub@tools$DEtest_CellType$AllMarkers_wilcox
DEGs <- DEGs[with(DEGs, avg_log2FC > 1 & p_val_adj < 0.05), ]

display_markers <- do.call(rbind, lapply(split(DEGs, DEGs$group1), function(x) {
  head(x[order(x$avg_log2FC, decreasing = TRUE), ], 10)
}))
display_markers <- display_markers[!duplicated(display_markers$gene), ]
stopifnot(nrow(display_markers) > 0)

ht <- FeatureHeatmap(
  pancreas_sub,
  group.by = "CellType",
  features = display_markers$gene,
  feature_split = display_markers$group1,
  max_cells = Inf,
  exp_legend_title = "Z-score",
  anno_terms = FALSE
)
print(ht$plot)
```

## Run Over-Representation Analysis

`RunEnrichment()` tests whether filtered marker sets are enriched for terms from
the selected database.
`GO_BP` is Gene Ontology Biological Process. The specific enriched terms are the
rows shown by `EnrichmentPlot()`.

```{r ora}
ora <- pancreas_sub@tools$Enrichment_CellType_wilcox$enrichment
head(ora)
data.frame(group = unique(ora$Groups), significant_terms = vapply(
  unique(ora$Groups), function(g) sum(ora$p.adjust[ora$Groups == g] < 0.05,
                                    na.rm = TRUE), integer(1)))
# Reuse the measured result above; plotting does not rerun enrichment.

EnrichmentPlot(
  pancreas_sub,
  db = "custom",
  group.by = "CellType",
  group_use = c("Ductal", "Endocrine"),
  plot_type = "bar"
)
```

Use different plot types for different questions: bar or lollipop plots for
ranked terms, word clouds for broad themes, and networks or enrichmaps for term
relationships.
The exported table includes tested terms that did not pass adjusted p < 0.05;
its row count is not a count of significant pathways. Related GO terms often
share genes, so several similar terms are not independent biological findings.

```{r enrichment-plots}
EnrichmentPlot(
  pancreas_sub,
  db = "custom",
  group.by = "CellType",
  group_use = "Ngn3-low-EP",
  plot_type = "network"
)

EnrichmentPlot(
  pancreas_sub,
  db = "custom",
  group.by = "CellType",
  plot_type = "comparison",
  topTerm = 3
)
```

In the network, numbered circles refer to the GO terms in the legend;
rectangles are genes and edges are annotation membership. This is not an
inferred gene-regulatory network. Its axes are layout coordinates, not
expression or pseudotime. In the comparison plot, dot size is GeneRatio and
colour shows adjusted significance; use the exported table to distinguish
nonsignificant terms from missing results.

## Run GSEA

Use GSEA when ranked gene statistics are more appropriate than a hard marker
cutoff.
ORA depends on a marker cutoff, while GSEA uses a ranked gene list. Use GSEA
when many genes shift modestly and a binary marker threshold would discard
useful signal. This run ranks all finite tested log2-fold changes in both
directions; it does not prefilter genes by significance. Positive NES supports
enrichment toward the positive end of that group-versus-rest ranking, negative
NES toward the other end. Statistical significance and effect direction must
both be read from the real result table. GSEA performs substantial permutation
work and can take longer than DE or plotting.

```{r gsea}
pancreas_sub <- RunGSEA(
  pancreas_sub,
  group.by = "CellType",
  species = "Mus_musculus",
  TERM2GENE = go_term2gene, TERM2NAME = go_term2name,
  DE_threshold = "is.finite(avg_log2FC)"
)

gsea_endocrine <- subset(
  pancreas_sub@tools$GSEA_CellType_wilcox$enrichment,
  Groups == "Endocrine" & Database == "custom"
)
if (nrow(gsea_endocrine) > 0 && any(gsea_endocrine$p.adjust < 0.05, na.rm = TRUE)) {
gsea_id <- gsea_endocrine$ID[which.min(gsea_endocrine$p.adjust)]
print(head(gsea_endocrine[order(gsea_endocrine$p.adjust),
                         c("ID", "Description", "NES", "p.adjust")], 5))

print(GSEAPlot(
  pancreas_sub,
  db = "custom",
  group.by = "CellType",
  group_use = "Endocrine",
  id_use = gsea_id
))

print(GSEAPlot(
  pancreas_sub,
  db = "custom",
  group.by = "CellType",
  group_use = "Endocrine",
  plot_type = "bar",
  direction = "both",
  topTerm = 10
))
} else {
  cat("No Endocrine GO terms passed adjusted p < 0.05; inspect the complete GSEA table and input overlap.\n")
}
```



In the running-score plot, black ticks locate gene-set members in the ranked
gene list; the green curve is the cumulative enrichment score. The lower
panel is the log2-fold-change ranking statistic. The horizontal coordinate is
a gene rank, not cell order or developmental time. Read the signed NES and
adjusted p-value together with the term's contributing genes.

## Save, reload, and check the hand-off

The build uses a temporary directory so repeated rendering cannot overwrite a
previous analysis. For your work, replace it with a new persistent output folder.
Save the object **and** plain-text metadata; a figure alone cannot reproduce the
analysis. No cells are sampled for this export.

```{r export-reload}
output_dir <- file.path(tempdir(), "differential-enrichment-workflow")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
result_file <- file.path(output_dir, "analysis.rds")
saveRDS(pancreas_sub, result_file)
utils::write.csv(de, file.path(output_dir, "all-markers.csv"), row.names = FALSE)
utils::write.csv(go_term2gene, file.path(output_dir, "GO-BP-measured-background.csv"), row.names = FALSE)
writeLines(paste("GO_BP", go_version), file.path(output_dir, "GO-source.txt"))
utils::write.csv(pancreas_sub@tools$Enrichment_CellType_wilcox$enrichment,
                 file.path(output_dir, "ORA.csv"), row.names = FALSE)
utils::write.csv(pancreas_sub@tools$GSEA_CellType_wilcox$enrichment,
                 file.path(output_dir, "GSEA.csv"), row.names = FALSE)
utils::write.csv(pancreas_sub@meta.data, file.path(output_dir, "cell-metadata.csv"))
reloaded <- readRDS(result_file)
stopifnot(identical(colnames(reloaded), colnames(pancreas_sub)),
          identical(reloaded@meta.data, pancreas_sub@meta.data),
          identical(SeuratObject::Reductions(reloaded),
                    SeuratObject::Reductions(pancreas_sub)),
          identical(GetAssayData5(reloaded, assay = "RNA", layer = "counts"),
                    GetAssayData5(pancreas_sub, assay = "RNA", layer = "counts")))
stopifnot(identical(reloaded@tools$DEtest_CellType$AllMarkers_wilcox, de),
          identical(reloaded@tools$Enrichment_CellType_wilcox$enrichment,
                    pancreas_sub@tools$Enrichment_CellType_wilcox$enrichment),
          identical(reloaded@tools$GSEA_CellType_wilcox$enrichment,
                    pancreas_sub@tools$GSEA_CellType_wilcox$enrichment))
data.frame(file = basename(list.files(output_dir, full.names = TRUE)),
           bytes = file.info(list.files(output_dir, full.names = TRUE))$size)
```

## Use your own data and recover common errors

- Replace the `data(...)` input with `readRDS("your-object.rds")`, or create a
  Seurat object from a gene-by-cell raw-count matrix. Check that cell IDs are
  unique and metadata row names equal the expression column names.
- Set the assay and grouping columns explicitly. Species, gene identifiers and
  reference labels must match your experiment; do not copy pancreas labels to
  unrelated tissue.
- If a reduction or metadata column is missing, inspect
  `SeuratObject::Reductions(object)` and `colnames(object@meta.data)`, then run
  its producer chunk. A plot does not create the result it needs.
- If GO returns no terms, inspect gene identifiers, overlap, set-size filters
  and the actual adjusted p-values. Do not loosen thresholds merely to force a
  positive result. A valid empty result and an unavailable database are different.
- If a dependency is unavailable, install it in the analysis environment,
  restart R and rerun from the input chunk. A skipped or failed method is not
  evidence that it worked. Preserve warnings and the exact error in your report.

## Build and analysis provenance

The receipt records the R seed. Python methods retain their documented backend
seeds and parameters; setting the R seed alone does not control every backend.
Package versions, BLAS libraries and backend changes can alter numerical results. Keep
this receipt with the exported object. It describes this page's actual render.

```{r provenance}
source_revision <- Sys.getenv("GITHUB_SHA")
if (!nzchar(source_revision) && nzchar(Sys.which("git"))) {
  source_revision <- tryCatch(system2("git", c("rev-parse", "HEAD"), stdout = TRUE,
                                      stderr = FALSE)[1], error = function(e) NA_character_)
}
article_source <- knitr::current_input(dir = TRUE)
build_receipt <- data.frame(
  article = "differential-enrichment-workflow",
  built_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
  scop = as.character(utils::packageVersion("scop")),
  R = as.character(getRversion()),
  source_revision = source_revision,
  source_has_local_changes = if (nzchar(Sys.which("git")))
    length(system2("git", c("status", "--porcelain"), stdout = TRUE, stderr = FALSE)) > 0 else NA,
  article_md5 = unname(tools::md5sum(article_source)),
  R_seed = 11,
  cells_retained = ncol(pancreas_sub),
  genes_in_active_assay = nrow(pancreas_sub),
  stringsAsFactors = FALSE
)
build_receipt
utils::write.csv(build_receipt, file.path(output_dir, "build-receipt.csv"),
                 row.names = FALSE)
writeLines(capture.output(sessionInfo()), file.path(output_dir, "session-info.txt"))
```

<details>
<summary>Show the complete R session used for this page</summary>

```{r session-info}
sessionInfo()
```

</details>

## What to Report

A compact differential and enrichment report should include:

- the grouping column, contrast, and DE method;
- whether the test is single-cell marker detection, sample-level pseudobulk, or
  bulk-style differential testing;
- thresholds for log fold change, adjusted p value, and marker filtering;
- the gene-set database, species, and enrichment method;
- the exact `srt@tools` result names used for downstream plots;
- pathway interpretations separated from the gene-level DE evidence.

Use enrichment to summarize marker lists or ranked statistics. Do not treat an
enriched term as proof that every gene in the pathway changes coherently.
