---
title: "Integration benchmark workflow"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  error = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 12, fig.height = 8,
  dpi = 144,
  out.width = "100%"
)
```

## What you will do

- **Question:** which method balances batch mixing and biological preservation on these cells?
- **Input:** All 1,600 cells in `panc8_sub`, with `tech` for batch and `celltype` for biological evaluation.
- **Output:** Per-method run status, latent-space metrics, summary plots and exported score tables.
- **How to run:** start a fresh R session and run the chunks in order. Printed tables and figures below are generated from that run; code-only alternatives are identified explicitly.



This article compares batch-correction methods on the same scRNA-seq object
with [RunIntegrationBenchmark()]. The runner scores each **latent** embedding
(not UMAP) with scIB-style metrics, stores tables in
`srt@tools$IntegrationBenchmark`, and returns the Seurat object for plotting
with [IntegrationBenchmarkPlot()].

Use this workflow when you need a reproducible side-by-side report for
`Uncorrected`, `Harmony`, `fastMNN`, or other [RunIntegration()] methods on
the same input.

## Prepare the required Harmony source

SCOP's `RunHarmony2()` checks the official GitHub package origin. A CRAN
installation with the same package name does not satisfy that check and can
trigger an installation attempt in the middle of an analysis. Install the
exact source before starting this page, then restart R:

```r
remotes::install_github(
  "immunogenomics/harmony@df19af23ae0639bd6ea2da63898f973f08c85862",
  upgrade = "never"
)
```

The tutorial's dependency preflight is read-only. It fails with a clear setup
requirement instead of allowing an implicit installation during fitting.

## Load data

```{r load-data}
library(scop)
set.seed(11)
harmony_source <- utils::packageDescription("harmony")
stopifnot(identical(harmony_source$RemoteUsername, "immunogenomics"),
          identical(harmony_source$RemoteSha, "df19af23ae0639bd6ea2da63898f973f08c85862"))
data.frame(package = "harmony", version = harmony_source$Version,
           source = paste(harmony_source$RemoteUsername, harmony_source$RemoteRepo, sep = "/"),
           revision = harmony_source$RemoteSha)

data(panc8_sub)
table(panc8_sub$tech)
table(panc8_sub$celltype)
```

## Run the benchmark

Keep the batch column (`tech`) and a biological label column (`celltype`) fixed
across methods. LISI and clustering metrics are computed on each method's latent
reduction (for example `Harmony` or `Uncorrectedpca`).

```{r run-benchmark}
panc8_sub <- RunIntegrationBenchmark(
  panc8_sub,
  batch = "tech",
  celltype = "celltype",
  methods = c("Uncorrected", "Harmony"),
  nHVF = 2000,
  linear_reduction_dims = 50,
  linear_reduction_dims_use = 1:30,
  perplexity = 30,
  skip_failed = FALSE,
  nonlinear_reduction_dims = 2, seed = 11,
  verbose = FALSE
)
```

The 2,000 variable genes and 30 latent dimensions define the shared model input;
all cells are scored. No cell subsampling is performed. `skip_failed = FALSE`
stops this two-method comparison if either method fails, so one successful
method cannot be mistaken for a complete comparison.

## Print summary tables

Use `thisplot::print_colored_table()` on the stored tables. The summary view
shows scaled overall, bio, and batch scores; the metrics view lists every
metric per method.

```{r print-tables}
bench <- panc8_sub@tools$IntegrationBenchmark
bench$runs
stopifnot(all(bench$runs$status == "success"))

thisplot::print_colored_table(
  bench$summary,
  by = "row",
  palette = "Chinese"
)

metrics <- bench$metrics
metrics <- metrics[
  !grepl("_LISI_mean$", metrics$metric), ,
  drop = FALSE
]
thisplot::print_colored_table(
  metrics,
  by = "col",
  palette = "Paired"
)
```

## Read the scores without overclaiming

Read the `value` and `scaled` columns separately. Raw iLISI increases when
neighborhoods contain more batches; raw cLISI is closest to 1 when a
neighborhood retains a single cell type. The scaled cLISI score reverses that
direction so larger scaled values represent better label preservation. The
printed `direction` field refers to the scaled score, not raw cLISI.

The biological summary combines label separation, connectivity and clustering
agreement. The batch summary combines mixing measures. The overall score uses
60% biological and 40% batch weight by default; it is a chosen trade-off, not
a probability of success. Inspect every metric and the label-by-batch balance
before preferring a method. A high batch score can accompany overcorrection,
and labels unique to one technology cannot be made to mix legitimately.

Missing metrics remain missing rather than becoming a successful result. This
small bundled study subset illustrates comparison on the same cells; it does
not establish a universal ranking of integration methods. To adapt the
benchmark, keep input cells, feature selection, dimensions and evaluation labels
fixed, then report the same raw tables and run statuses for each method.

## Plot all views

`IntegrationBenchmarkPlot()` with `plot_type = "auto"` returns a named list:
`box` (per-cell LISI), `heatmap`, `scatter`, and `umap`.

```{r plot-auto, fig.height=8}
plots <- IntegrationBenchmarkPlot(panc8_sub, verbose = FALSE)
names(plots)
plots$box
plots$scatter
plots$heatmap
plots$umap
```

Individual plot types are also available:

```r
IntegrationBenchmarkPlot(panc8_sub, plot_type = "box")
IntegrationBenchmarkPlot(panc8_sub, plot_type = "umap")
```



## Save, reload, and check the hand-off

The build uses a temporary directory so repeated rendering cannot overwrite a
previous analysis. For your work, replace it with a new persistent output folder.
Save the object **and** plain-text metadata; a figure alone cannot reproduce the
analysis. No cells are sampled for this export.

```{r export-reload}
output_dir <- file.path(tempdir(), "integration-benchmark-workflow")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
result_file <- file.path(output_dir, "analysis.rds")
saveRDS(panc8_sub, result_file)
utils::write.csv(bench$summary, file.path(output_dir, "benchmark-summary.csv"), row.names = FALSE)
utils::write.csv(bench$metrics, file.path(output_dir, "benchmark-metrics.csv"), row.names = FALSE)
utils::write.csv(bench$runs, file.path(output_dir, "benchmark-runs.csv"), row.names = FALSE)
utils::write.csv(panc8_sub@meta.data, file.path(output_dir, "cell-metadata.csv"))
reloaded <- readRDS(result_file)
stopifnot(identical(colnames(reloaded), colnames(panc8_sub)),
          identical(reloaded@meta.data, panc8_sub@meta.data),
          identical(SeuratObject::Reductions(reloaded),
                    SeuratObject::Reductions(panc8_sub)))
data.frame(file = basename(list.files(output_dir, full.names = TRUE)),
           bytes = file.info(list.files(output_dir, full.names = TRUE))$size)
```

## Use your own data and recover common errors

- Replace the `data(...)` input with `readRDS("your-object.rds")`, or create a
  Seurat object from a gene-by-cell raw-count matrix. Check that cell IDs are
  unique and metadata row names equal the expression column names.
- Set the assay and grouping columns explicitly. Species, gene identifiers and
  reference labels must match your experiment; do not copy pancreas labels to
  unrelated tissue.
- If a reduction or metadata column is missing, inspect
  `SeuratObject::Reductions(object)` and `colnames(object@meta.data)`, then run
  its producer chunk. A plot does not create the result it needs.
- If a dependency is unavailable, install it in the analysis environment,
  restart R and rerun from the input chunk. A skipped or failed method is not
  evidence that it worked. Preserve warnings and the exact error in your report.

## Build and analysis provenance

The receipt records the R seed. Python methods retain their documented backend
seeds and parameters; setting the R seed alone does not control every backend.
Package versions, BLAS libraries and backend changes can alter numerical results. Keep
this receipt with the exported object. It describes this page's actual render.

```{r provenance}
source_revision <- Sys.getenv("GITHUB_SHA")
if (!nzchar(source_revision) && nzchar(Sys.which("git"))) {
  source_revision <- tryCatch(system2("git", c("rev-parse", "HEAD"), stdout = TRUE,
                                      stderr = FALSE)[1], error = function(e) NA_character_)
}
article_source <- knitr::current_input(dir = TRUE)
build_receipt <- data.frame(
  article = "integration-benchmark-workflow",
  built_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
  scop = as.character(utils::packageVersion("scop")),
  R = as.character(getRversion()),
  source_revision = source_revision,
  source_has_local_changes = if (nzchar(Sys.which("git")))
    length(system2("git", c("status", "--porcelain"), stdout = TRUE, stderr = FALSE)) > 0 else NA,
  article_md5 = unname(tools::md5sum(article_source)),
  R_seed = 11,
  cells_retained = ncol(panc8_sub),
  genes_in_active_assay = nrow(panc8_sub),
  stringsAsFactors = FALSE
)
build_receipt
utils::write.csv(build_receipt, file.path(output_dir, "build-receipt.csv"),
                 row.names = FALSE)
writeLines(capture.output(sessionInfo()), file.path(output_dir, "session-info.txt"))
```

<details>
<summary>Show the complete R session used for this page</summary>

```{r session-info}
sessionInfo()
```

</details>

## What to report

A compact integration benchmark report should include:

- batch and cell-type columns used for scoring;
- methods compared and shared integration parameters (`nHVF`, latent dimensions);
- the colored summary and metrics tables;
- LISI boxplots, bio-vs-batch scatter, and UMAP panels colored by batch and
  cell type.

Overall score uses scIB weights: `0.6 * bio + 0.4 * batch`. Higher is better
for all scaled metrics shown in the summary table.

A higher batch score alone can reward overcorrection. Compare biological
retention, batch mixing and marker patterns together. Results depend on this
dataset and labels, and do not establish a universally best integration method.
