---
title: "Integration benchmark workflow"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.align = "center",
  dpi = 144,
  out.width = "92%"
)
```

This article compares batch-correction methods on the same scRNA-seq object
with [RunIntegrationBenchmark()]. The runner scores each **latent** embedding
(not UMAP) with scIB-style metrics, stores tables in
`srt@tools$IntegrationBenchmark`, and returns the Seurat object for plotting
with [IntegrationBenchmarkPlot()].

Use this workflow when you need a reproducible side-by-side report for
`Uncorrected`, `Harmony`, `fastMNN`, or other [RunIntegration()] methods on
the same input.

## Load data

```{r load-data}
library(scop)

data(panc8_sub)
table(panc8_sub$tech)
table(panc8_sub$celltype)
```

## Run the benchmark

Keep the batch column (`tech`) and a biological label column (`celltype`) fixed
across methods. LISI and clustering metrics are computed on each method's latent
reduction (for example `Harmony` or `Uncorrectedpca`).

```{r run-benchmark}
panc8_sub <- RunIntegrationBenchmark(
  panc8_sub,
  batch = "tech",
  celltype = "celltype",
  methods = c("Uncorrected", "Harmony"),
  nHVF = 500,
  linear_reduction_dims = 20,
  linear_reduction_dims_use = 1:10,
  perplexity = 10,
  verbose = FALSE
)
```

## Print summary tables

Use `thisplot::print_colored_table()` on the stored tables. The summary view
shows scaled overall, bio, and batch scores; the metrics view lists every
metric per method.

```{r print-tables}
bench <- panc8_sub@tools$IntegrationBenchmark

thisplot::print_colored_table(
  bench$summary,
  by = "row",
  palette = "Chinese"
)

metrics <- bench$metrics
metrics <- metrics[
  !grepl("_LISI_mean$", metrics$metric),
  ,
  drop = FALSE
]
thisplot::print_colored_table(
  metrics,
  by = "col",
  palette = "Paired"
)
```

## Plot all views

`IntegrationBenchmarkPlot()` with `plot_type = "auto"` returns a named list:
`box` (per-cell LISI), `heatmap`, `scatter`, and `umap`.

```{r plot-auto, fig.height=8}
plots <- IntegrationBenchmarkPlot(panc8_sub, verbose = FALSE)
names(plots)
plots$box
plots$scatter
plots$heatmap
```

Individual plot types are also available:

```r
IntegrationBenchmarkPlot(panc8_sub, plot_type = "box")
IntegrationBenchmarkPlot(panc8_sub, plot_type = "umap")
```

## What to report

A compact integration benchmark report should include:

- batch and cell-type columns used for scoring;
- methods compared and shared integration parameters (`nHVF`, latent dimensions);
- the colored summary and metrics tables;
- LISI boxplots, bio-vs-batch scatter, and UMAP panels colored by batch and
  cell type.

Overall score uses scIB weights: `0.6 * bio + 0.4 * batch`. Higher is better
for all scaled metrics shown in the summary table.
