---
title: "Data integration workflow"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  error = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 12, fig.height = 8,
  dpi = 144,
  out.width = "100%"
)
```

## What you will do

- **Question:** does correction reduce technology effects while retaining cell types?
- **Input:** All 1,600 cells and 12,932 supplied genes in `panc8_sub`; `tech` identifies five technologies and `celltype` supplies evaluation labels.
- **Output:** Uncorrected and Harmony reductions on the same cells, marker checks and a reloadable integrated object.
- **How to run:** start a fresh R session and run the chunks in order. Printed tables and figures below are generated from that run; code-only alternatives are identified explicitly.



Use integration when technical structure prevents comparison across samples or
platforms, and retain an unintegrated baseline. If every condition occurs in
only one batch, batch and biology are confounded: no attractive embedding can
resolve that experimental-design problem.

## Prepare the required Harmony source

SCOP's `RunHarmony2()` checks the official GitHub package origin. A CRAN
installation with the same package name does not satisfy that check and can
trigger an installation attempt in the middle of an analysis. Install the
exact source before starting this page, then restart R:

```r
remotes::install_github(
  "immunogenomics/harmony@df19af23ae0639bd6ea2da63898f973f08c85862",
  upgrade = "never"
)
```

The tutorial's dependency preflight is read-only. It fails with a clear setup
requirement instead of allowing an implicit installation during fitting.

## Prepare a Merged Object

The example uses the bundled `panc8_sub` object, a subset of human pancreas
datasets. The `tech` column records the batch or technology label.
The `celltype` column is used as known biology for diagnostics, not as a target
to force during integration.

```{r load-object}
library(scop)
set.seed(11)
harmony_source <- utils::packageDescription("harmony")
stopifnot(identical(harmony_source$RemoteUsername, "immunogenomics"),
          identical(harmony_source$RemoteSha, "df19af23ae0639bd6ea2da63898f973f08c85862"))
data.frame(package = "harmony", version = harmony_source$Version,
           source = paste(harmony_source$RemoteUsername, harmony_source$RemoteRepo, sep = "/"),
           revision = harmony_source$RemoteSha)

data(panc8_sub)
input_cells <- colnames(panc8_sub)
stopifnot(!anyNA(panc8_sub$tech), !anyNA(panc8_sub$celltype))
data.frame(genes = nrow(panc8_sub), cells = ncol(panc8_sub))
table(panc8_sub$tech)
table(panc8_sub$celltype)
```

Always inspect the unintegrated embedding or PCA first. This gives you a
baseline for deciding whether integration removes technical structure without
erasing cell-type structure.

```{r unintegrated-check}
panc8_sub <- RunStandardWorkflow(panc8_sub, assay = "RNA", seed = 11, cores = 1,
                                 verbose = FALSE)

CellDimPlot(
  panc8_sub,
  group.by = c("tech", "celltype"),
  reduction = "UMAP",
  theme_use = "theme_blank"
)
```

## Run an Integration Method

Harmony is a common first pass when the object is already a merged Seurat
object with a batch column.

```{r run-harmony}
panc8_harmony <- RunIntegration(
  srt_merge = panc8_sub,
  batch = "tech",
  integration_methods = "Harmony",
  nonlinear_reduction_dims = 2,
  seed = 11, verbose = FALSE
)

stopifnot(identical(colnames(panc8_harmony), input_cells))
SeuratObject::Reductions(panc8_harmony)
```

Use the method-specific integrated UMAP for plotting. The exact reduction name
depends on the selected method; for Harmony, the examples use `HarmonyUMAP2D`.
When comparing methods, report the reduction name explicitly. A plot labeled
only as UMAP is ambiguous once an object contains uncorrected and integrated
embeddings.

```{r plot-harmony}
CellDimPlot(
  panc8_harmony,
  group.by = c("tech", "celltype"),
  reduction = "HarmonyUMAP2D",
  theme_use = "theme_blank"
)
```

## Choose a Method

Use method choice to match the data and dependency budget:

- `Harmony`, `fastMNN`, `MNN`, `ComBat`, and `Seurat` methods are practical
  R-side choices for many scRNA-seq batch problems.
- `scVI`, `BBKNN`, `Scanorama`, `GLUE`, and `MultiMAP` require Python-side
  dependencies but can be useful for large datasets or multimodal settings.
- `WNN` and cross-modality methods are better suited to multimodal RNA + ATAC
  or paired-modality objects than to ordinary batch correction.
- `Uncorrected` is useful as a baseline and should be kept in reports when
  showing that integration changed the object.



The following alternatives are recipes, not additional measured results in this
page. To compare executed methods, use the [integration benchmark](integration-benchmark-workflow.html).
Other methods follow the same entry-point pattern, but may require additional
runtime dependencies or matrix backends:

```r
panc8_rpca <- RunIntegration(
  srt_merge = panc8_sub,
  batch = "tech",
  integration_methods = "RPCA"
)

panc8_fastmnn <- RunIntegration(
  srt_merge = panc8_sub,
  batch = "tech",
  integration_methods = "fastMNN"
)
```

## Inspect Integration Quality

Do not judge integration from batch mixing alone. Also check known labels,
marker genes, and whether rare cell types remain separated.

```{r inspect-quality}
CellDimPlot(
  panc8_harmony,
  group.by = "tech",
  reduction = "HarmonyUMAP2D",
  theme_use = "theme_blank"
)

CellDimPlot(
  panc8_harmony,
  group.by = "celltype",
  reduction = "HarmonyUMAP2D",
  label = TRUE,
  label_repel = TRUE, label.size = 3,
  theme_use = "theme_blank"
)

FeatureDimPlot(
  panc8_harmony,
  features = c("GCG", "SST", "KRT19"),
  reduction = "HarmonyUMAP2D"
)
```

When comparing methods, keep the input object fixed and compare the same
metadata labels, marker genes, and diagnostic plots across methods. The
unintegrated baseline is useful because it shows what the integration changed.
Good integration mixes batches within the same biological state while keeping
distinct cell types and marker programs separated. Overcorrection is visible
when expected marker-defined groups collapse together.



INS is absent from the supplied feature-filtered object, so this marker
check uses measured GCG, SST and KRT19. Missing input genes must not be
interpreted as measured zero expression.

## Save, reload, and check the hand-off

The build uses a temporary directory so repeated rendering cannot overwrite a
previous analysis. For your work, replace it with a new persistent output folder.
Save the object **and** plain-text metadata; a figure alone cannot reproduce the
analysis. No cells are sampled for this export.

```{r export-reload}
output_dir <- file.path(tempdir(), "integration-workflow")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
result_file <- file.path(output_dir, "analysis.rds")
saveRDS(panc8_harmony, result_file)
utils::write.csv(panc8_harmony@meta.data, file.path(output_dir, "cell-metadata.csv"))
reloaded <- readRDS(result_file)
stopifnot(identical(colnames(reloaded), colnames(panc8_harmony)),
          identical(reloaded@meta.data, panc8_harmony@meta.data),
          identical(SeuratObject::Reductions(reloaded),
                    SeuratObject::Reductions(panc8_harmony)))
data.frame(file = basename(list.files(output_dir, full.names = TRUE)),
           bytes = file.info(list.files(output_dir, full.names = TRUE))$size)
```

## Use your own data and recover common errors

- Replace the `data(...)` input with `readRDS("your-object.rds")`, or create a
  Seurat object from a gene-by-cell raw-count matrix. Check that cell IDs are
  unique and metadata row names equal the expression column names.
- Set the assay and grouping columns explicitly. Species, gene identifiers and
  reference labels must match your experiment; do not copy pancreas labels to
  unrelated tissue.
- If a reduction or metadata column is missing, inspect
  `SeuratObject::Reductions(object)` and `colnames(object@meta.data)`, then run
  its producer chunk. A plot does not create the result it needs.
- If a dependency is unavailable, install it in the analysis environment,
  restart R and rerun from the input chunk. A skipped or failed method is not
  evidence that it worked. Preserve warnings and the exact error in your report.

## Build and analysis provenance

The receipt records the R seed. Python methods retain their documented backend
seeds and parameters; setting the R seed alone does not control every backend.
Package versions, BLAS libraries and backend changes can alter numerical results. Keep
this receipt with the exported object. It describes this page's actual render.

```{r provenance}
source_revision <- Sys.getenv("GITHUB_SHA")
if (!nzchar(source_revision) && nzchar(Sys.which("git"))) {
  source_revision <- tryCatch(system2("git", c("rev-parse", "HEAD"), stdout = TRUE,
                                      stderr = FALSE)[1], error = function(e) NA_character_)
}
article_source <- knitr::current_input(dir = TRUE)
build_receipt <- data.frame(
  article = "integration-workflow",
  built_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
  scop = as.character(utils::packageVersion("scop")),
  R = as.character(getRversion()),
  source_revision = source_revision,
  source_has_local_changes = if (nzchar(Sys.which("git")))
    length(system2("git", c("status", "--porcelain"), stdout = TRUE, stderr = FALSE)) > 0 else NA,
  article_md5 = unname(tools::md5sum(article_source)),
  R_seed = 11,
  cells_retained = ncol(panc8_harmony),
  genes_in_active_assay = nrow(panc8_harmony),
  stringsAsFactors = FALSE
)
build_receipt
utils::write.csv(build_receipt, file.path(output_dir, "build-receipt.csv"),
                 row.names = FALSE)
writeLines(capture.output(sessionInfo()), file.path(output_dir, "session-info.txt"))
```

<details>
<summary>Show the complete R session used for this page</summary>

```{r session-info}
sessionInfo()
```

</details>

## What to Report

A compact integration report should include:

- the batch column and biological annotation columns used for evaluation;
- the integration method and major parameters;
- the integrated reduction names used for plots;
- before and after plots colored by batch and biological labels;
- marker-gene checks for expected cell types;
- any method-specific dependencies or environment setup needed to reproduce the
  run.

Integration is a preprocessing decision, not a biological result by itself. Use
it only when the integrated representation improves downstream interpretation
without hiding real biological differences.
