---
title: "Data integration workflow"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.align = "center",
  dpi = 144,
  out.width = "92%"
)
```

This article shows how to run and inspect dataset integration in `scop`.
Integration should be used when technical batches, platforms, donors, or
experiments obscure the biological structure that should be compared across
datasets.

`integration_scop()` provides a common entry point for several integration
methods. The key practical decisions are the batch column, the integration
method, the reduction names to inspect, and whether the integrated structure
preserves known biology.

Integration is not a required step for every dataset. Use it when technical
structure prevents comparison across samples or platforms. Keep an unintegrated
baseline so the report can show what changed after correction.

## Prepare a Merged Object

The example uses the bundled `panc8_sub` object, a subset of human pancreas
datasets. The `tech` column records the batch or technology label.
The `celltype` column is used as known biology for diagnostics, not as a target
to force during integration.

```{r load-object}
library(scop)

data(panc8_sub)
table(panc8_sub$tech)
table(panc8_sub$celltype)
```

Always inspect the unintegrated embedding or PCA first. This gives you a
baseline for deciding whether integration removes technical structure without
erasing cell-type structure.

```{r unintegrated-check}
panc8_sub <- standard_scop(panc8_sub, verbose = FALSE)

CellDimPlot(
  panc8_sub,
  group.by = c("tech", "celltype"),
  reduction = "UMAP",
  theme_use = "theme_blank"
)
```

## Run an Integration Method

Harmony is a common first pass when the object is already a merged Seurat
object with a batch column.

```{r run-harmony}
panc8_harmony <- integration_scop(
  srt_merge = panc8_sub,
  batch = "tech",
  integration_method = "Harmony"
)

SeuratObject::Reductions(panc8_harmony)
```

Use the method-specific integrated UMAP for plotting. The exact reduction name
depends on the selected method; for Harmony, the examples use `HarmonyUMAP`.
When comparing methods, report the reduction name explicitly. A plot labeled
only as UMAP is ambiguous once an object contains uncorrected and integrated
embeddings.

```{r plot-harmony}
CellDimPlot(
  panc8_harmony,
  group.by = c("tech", "celltype"),
  reduction = "HarmonyUMAP",
  theme_use = "theme_blank"
)
```

## Choose a Method

Use method choice to match the data and dependency budget:

- `Harmony`, `fastMNN`, `MNN`, `ComBat`, and `Seurat` methods are practical
  R-side choices for many scRNA-seq batch problems.
- `scVI`, `BBKNN`, `Scanorama`, `GLUE`, and `MultiMAP` require Python-side
  dependencies but can be useful for large datasets or multimodal settings.
- `WNN` and cross-modality methods are better suited to multimodal RNA + ATAC
  or paired-modality objects than to ordinary batch correction.
- `Uncorrected` is useful as a baseline and should be kept in reports when
  showing that integration changed the object.

```{r method-examples}
panc8_uncorrected <- integration_scop(
  srt_merge = panc8_harmony,
  batch = "tech",
  integration_method = "Uncorrected",
  verbose = FALSE
)
```

Other methods follow the same entry-point pattern, but may require additional
runtime dependencies or matrix backends:

```r
panc8_rpca <- integration_scop(
  srt_merge = panc8_sub,
  batch = "tech",
  integration_method = "RPCA"
)

panc8_fastmnn <- integration_scop(
  srt_merge = panc8_sub,
  batch = "tech",
  integration_method = "fastMNN"
)
```

## Inspect Integration Quality

Do not judge integration from batch mixing alone. Also check known labels,
marker genes, and whether rare cell types remain separated.

```{r inspect-quality}
CellDimPlot(
  panc8_harmony,
  group.by = "tech",
  reduction = "HarmonyUMAP",
  theme_use = "theme_blank"
)

CellDimPlot(
  panc8_harmony,
  group.by = "celltype",
  reduction = "HarmonyUMAP",
  label = TRUE,
  theme_use = "theme_blank"
)

FeatureDimPlot(
  panc8_harmony,
  features = c("INS", "GCG", "SST", "KRT19"),
  reduction = "HarmonyUMAP"
)
```

When comparing methods, keep the input object fixed and compare the same
metadata labels, marker genes, and diagnostic plots across methods. The
unintegrated baseline is useful because it shows what the integration changed.
Good integration mixes batches within the same biological state while keeping
distinct cell types and marker programs separated. Overcorrection is visible
when expected marker-defined groups collapse together.

## What to Report

A compact integration report should include:

- the batch column and biological annotation columns used for evaluation;
- the integration method and major parameters;
- the integrated reduction names used for plots;
- before and after plots colored by batch and biological labels;
- marker-gene checks for expected cell types;
- any method-specific dependencies or environment setup needed to reproduce the
  run.

Integration is a preprocessing decision, not a biological result by itself. Use
it only when the integrated representation improves downstream interpretation
without hiding real biological differences.
