---
title: "Reproducing pancreatic endocrinogenesis"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 144,
  out.width = "92%"
)

print_scop_plot <- function(x) {
  if (is.list(x) && "plot" %in% names(x)) {
    print(x$plot)
  } else {
    print(x)
  }
}
```

This article reproduces the main analytical logic of a pancreatic
endocrinogenesis study using the bundled `pancreas_sub` object. The biological
reference is the mouse embryonic pancreas single-cell study by
Bastidas-Ponce et al. (2019, *Development*, DOI: 10.1242/dev.173849).

The website displays this end-to-end workflow without evaluating its code
chunks because trajectory, velocity, and regulatory-network steps create and
use external R and Python environments. Run the chunks interactively in order
when reproducing the analysis locally.

The goal is a compact, runnable reproduction rather than a one-to-one rebuild
of every original figure. The workflow keeps the same evidence chain: identify
ductal, endocrine progenitor, pre-endocrine, and endocrine states; order them
along developmental trajectories; find dynamic genes; then connect downstream
transcription factors, regulatory network structure, marker genes, and pathway
terms.

## Reproduction Scope

The default dataset is a small annotated subset, so it is suitable for local
article rendering and interactive exploration. It can support the following
questions:

- whether ductal, Ngn3-positive endocrine progenitor, pre-endocrine, alpha,
  beta, delta, and epsilon states form an ordered manifold;
- whether known endocrine-development genes vary along this manifold;
- whether dynamic genes and TF activity recover expected regulators such as
  `Neurog3`, `Fev`, `Neurod1`, `Pdx1`, `Arx`, `Pax4`, `Mafb`, `Rfx6`, `Isl1`,
  and `Insm1`;
- whether differential genes and enriched terms agree with endocrine
  differentiation and hormone-producing cell states.

## Load and Process the Dataset

Start from the package object and run the standard single-cell preprocessing
workflow. This creates normalized expression, variable features, scaled data,
PCA, neighbor graph, clustering, and UMAP coordinates.

```{r preprocess}
library(scop)

data(pancreas_sub)

pancreas_sub <- standard_scop(
  pancreas_sub,
  assay = "RNA",
  nHVF = 2000,
  linear_reduction_dims = 50,
  cluster_resolution = 0.6,
  verbose = FALSE
)

pancreas_sub <- RunCellQC(pancreas_sub)

data.frame(
  genes = nrow(pancreas_sub),
  cells = ncol(pancreas_sub),
  cell_states = length(unique(pancreas_sub$SubCellType))
)

head(sort(table(pancreas_sub$SubCellType), decreasing = TRUE), 8)
```

## Check Cell States and Markers

The first reproduction step is to confirm that the annotated cell states and
canonical marker genes match the expected developmental progression. Ductal
markers support the root state; `Neurog3` marks endocrine progenitors; `Fev`
and `Neurod1` mark pre-endocrine cells; hormone genes separate terminal
endocrine states.

```{r cell-state-overview, fig.width=9, fig.height=6}
CellDimPlot(
  pancreas_sub,
  group.by = c("CellType", "SubCellType"),
  reduction = "UMAP",
  label = TRUE
)
```

```{r marker-feature-plot, fig.width=10, fig.height=7}
pancreas_markers <- c(
  "Sox9", "Anxa2",
  "Neurog3", "Hes6",
  "Fev", "Neurod1",
  "Rbp4", "Pyy",
  "Ins1", "Gcg", "Sst", "Ghrl"
)

FeatureDimPlot(
  pancreas_sub,
  features = pancreas_markers,
  reduction = "UMAP",
  ncol = 4
)
```

```{r marker-heatmap, fig.width=8, fig.height=7}
ht_markers <- GroupHeatmap(
  pancreas_sub,
  features = pancreas_markers,
  group.by = c("CellType", "SubCellType"),
  heatmap_palette = "Spectral",
  show_row_names = TRUE,
  add_dot = TRUE,
  add_reticle = TRUE
)
print(ht_markers$plot)
```

## Reconstruct Developmental Trajectories

The endocrine differentiation branch is directional, so the trajectory needs an
explicit start state. Here the ductal population is used as the root, with alpha
and beta states as two terminal endpoints.

```{r slingshot-trajectory, fig.width=8, fig.height=6}
pancreas_sub <- RunSlingshot(
  pancreas_sub,
  group.by = "SubCellType",
  reduction = "UMAP",
  start = "Ductal",
  end = c("Alpha", "Beta")
)

CellDimPlot(
  pancreas_sub,
  group.by = "SubCellType",
  lineages = c("Lineage1", "Lineage2"),
  reduction = "UMAP",
  label = TRUE
)
```

```{r pseudotime-feature-plot, fig.width=8, fig.height=4}
FeatureDimPlot(
  pancreas_sub,
  features = c("Lineage1", "Lineage2"),
  reduction = "UMAP"
)
```

Monocle3 provides an independent graph-based pseudotime layer. The root cell is
fixed to a ductal cell so the article renders without interactive selection.

```{r monocle3-trajectory, fig.width=9, fig.height=5}
ductal_root <- colnames(pancreas_sub)[pancreas_sub$SubCellType == "Ductal"][1]

pancreas_sub <- RunMonocle3(
  pancreas_sub,
  group.by = "SubCellType",
  use_partition = FALSE,
  root_cells = ductal_root
)

FeatureDimPlot(
  pancreas_sub,
  features = "Monocle3_Pseudotime",
  reduction = "UMAP"
)
```

## Add RNA Velocity

RNA velocity adds a local direction field on the same embedding. This is useful
for checking whether the inferred flow points from ductal and progenitor states
toward endocrine states.

```{r velocity, fig.width=8, fig.height=6}
pancreas_sub <- RunSCVELO(
  pancreas_sub,
  group.by = "SubCellType",
  linear_reduction = "PCA",
  nonlinear_reduction = "UMAP",
  verbose = FALSE
)

VelocityPlot(
  pancreas_sub,
  group.by = "SubCellType",
  reduction = "UMAP",
  plot_type = "stream"
)
```

## Detect Dynamic Genes Along Lineages

Dynamic genes connect the trajectory geometry to interpretable gene programs.
The heatmap is split along the two lineages and annotated with pathway terms and
feature classes.

```{r dynamic-features, fig.width=9, fig.height=8}
pancreas_sub <- RunDynamicFeatures(
  pancreas_sub,
  lineages = c("Lineage1", "Lineage2"),
  n_candidates = 200
)

pancreas_sub <- AnnotateFeatures(
  pancreas_sub,
  species = "Mus_musculus",
  db = c("TF", "CSPA")
)

ht_dynamic <- DynamicHeatmap(
  pancreas_sub,
  lineages = c("Lineage1", "Lineage2"),
  use_fitted = TRUE,
  n_split = 6,
  reverse_ht = "Lineage1",
  species = "Mus_musculus",
  db = "GO_BP",
  anno_terms = TRUE,
  anno_keys = FALSE,
  anno_features = FALSE,
  heatmap_palette = "viridis",
  cell_annotation = "SubCellType",
  feature_annotation = c("TF", "CSPA")
)
print(ht_dynamic$plot)
```

## Infer TF Links and Regulatory Activity

The reproduction then moves from dynamic genes to regulatory hypotheses. Use a
focused set of pancreatic-development TFs as candidate regulators and combine
them with marker genes and highly variable genes as targets. This keeps the
example small while preserving the key regulators expected in endocrine
differentiation.

```{r tf-and-target-sets}
pancreas_tfs <- c(
  "Sox9", "Neurog3", "Neurod1", "Fev", "Pdx1", "Nkx2-2", "Pax4",
  "Arx", "Mafb", "Mafa", "Rfx6", "Isl1", "Insm1"
)
pancreas_tfs <- intersect(pancreas_tfs, rownames(pancreas_sub))

pancreas_targets <- unique(c(
  pancreas_markers,
  pancreas_tfs,
  head(SeuratObject::VariableFeatures(pancreas_sub), 250)
))
pancreas_targets <- intersect(pancreas_targets, rownames(pancreas_sub))

data.frame(
  candidate_tfs = length(pancreas_tfs),
  candidate_targets = length(pancreas_targets),
  tf_examples = paste(head(pancreas_tfs, 8), collapse = ", ")
)
```

Run GRNBoost2 first to inspect direct TF-target edges. The table is useful for
checking which TFs are connected to hormone genes or progenitor-state markers
before regulon scoring.

```{r grnboost2}
grn_edges <- RunGRNBoost2(
  pancreas_sub,
  regulators = pancreas_tfs,
  targets = pancreas_targets,
  backend = "cpp",
  max_edges_per_target = 3,
  n_rounds = 500,
  cores = 1,
  seed = 1234,
  verbose = FALSE
)

head(grn_edges[order(grn_edges$importance, decreasing = TRUE), ], 10)
```

SCENIC prunes the regulatory links with cisTarget and scores regulon activity
per cell. The reference files are cached by `scop`; the first run may download
the mouse cisTarget resources.

```{r scenic-run}
scenic_dir <- file.path(tempdir(), "scop_pancreas_scenic")

pancreas_sub <- RunSCENIC(
  pancreas_sub,
  species = "Mus_musculus",
  backend = "cpp",
  regulators = pancreas_tfs,
  targets = pancreas_targets,
  work_dir = scenic_dir,
  min_regulon_size = 3,
  max_regulon_targets = 30,
  n_rounds = 500,
  cores = 1,
  seed = 1234,
  force = TRUE,
  verbose = FALSE
)

data.frame(
  scenic_entries = paste(names(pancreas_sub@tools$SCENIC), collapse = ", "),
  regulons = length(pancreas_sub@tools$SCENIC$regulon_list)
)
```

Rank regulons by cell-state specificity and keep the top regulons for heatmaps
and network plots.

```{r scenic-rss-rank, fig.width=9, fig.height=6}
scenic_rss <- SCENICPlot(
  pancreas_sub,
  group.by = "SubCellType",
  plot_type = "rss_rank",
  top_n = 6,
  highlight_tf = pancreas_tfs,
  label_size = 2.5
)
print_scop_plot(scenic_rss)
```

```{r scenic-top-regulons}
top_regulons <- unique(scenic_rss$top_table$regulon)
top_tfs <- unique(scenic_rss$top_table$TF)

top_regulons <- head(top_regulons, 10)
top_tfs <- head(top_tfs, 4)

data.frame(
  top_regulons = paste(top_regulons, collapse = ", "),
  top_tfs = paste(top_tfs, collapse = ", ")
)
```

```{r scenic-heatmaps, fig.width=8, fig.height=7}
scenic_rss_heatmap <- SCENICPlot(
  pancreas_sub,
  group.by = "SubCellType",
  plot_type = "rss_heatmap",
  features = top_regulons,
  rss_scale = TRUE,
  heatmap_order = "group",
  heatmap_show_row_names = TRUE,
  heatmap_cluster_columns = FALSE
)
print_scop_plot(scenic_rss_heatmap)

scenic_activity_heatmap <- SCENICPlot(
  pancreas_sub,
  group.by = "SubCellType",
  plot_type = "activity_heatmap",
  features = top_regulons,
  activity_scale = TRUE,
  heatmap_order = "group",
  heatmap_show_row_names = TRUE,
  heatmap_cluster_columns = FALSE
)
print_scop_plot(scenic_activity_heatmap)
```

```{r scenic-network, fig.width=9, fig.height=7}
scenic_network <- SCENICPlot(
  pancreas_sub,
  group.by = "SubCellType",
  plot_type = "network",
  network_tf = top_tfs,
  max_targets = 12,
  label_nodes = "tfs"
)
print_scop_plot(scenic_network)
```

```{r scenic-target-bar, fig.width=9, fig.height=5}
scenic_targets <- SCENICPlot(
  pancreas_sub,
  group.by = "SubCellType",
  plot_type = "target_bar",
  features = head(top_regulons, 4),
  max_targets = 12
)
print_scop_plot(scenic_targets)
```

## Differential Genes and Pathway Terms

Marker and enrichment analysis provide a second downstream layer. These results
should be interpreted together with the trajectory and TF analyses: dynamic
genes show when programs change, while DE and enrichment show which programs
separate annotated states.

```{r differential-and-enrichment, fig.width=10, fig.height=7}
pancreas_sub <- RunDEtest(
  pancreas_sub,
  group.by = "CellType",
  fc.threshold = 1,
  only.pos = FALSE
)

pancreas_sub <- RunEnrichment(
  pancreas_sub,
  group.by = "CellType",
  db = "GO_BP",
  species = "Mus_musculus",
  DE_threshold = "avg_log2FC > log2(1.5) & p_val_adj < 0.05",
  cores = 2
)

go_bp_overlay_values <- setNames(
  rep("#CFE8CC", 2),
  c(paste("Enrichment", "GO_BP", sep = " | "), "GO_BP")
)

de_volcano_plots <- DEtestPlot(
  pancreas_sub,
  group.by = "CellType",
  plot_type = "volcano",
  threshold_method = "hyperbolic",
  hyperbola_c = 6,
  annotate_enrichment = TRUE,
  enrich_from = "Enrichment",
  enrich_db = "GO_BP",
  enrich_top_terms = 2,
  enrich_nlabel = 10,
  label.size = 2,
  combine = FALSE
)

de_volcano_plots <- lapply(
  de_volcano_plots,
  function(p) {
    p +
      ggplot2::scale_fill_manual(
        name = "Enrichment",
        values = go_bp_overlay_values,
        labels = function(x) sub("^.* \\|\\s*", "", x)
      ) +
      ggplot2::guides(
        fill = ggplot2::guide_legend(override.aes = list(size = 2))
      ) +
      ggplot2::theme(legend.box = "vertical")
  }
)

patchwork::wrap_plots(de_volcano_plots, ncol = 2, guides = "collect") &
  ggplot2::theme(legend.position = "right")
```

Here `GO_BP` denotes the Gene Ontology Biological Process database used for the
enrichment overlay. The circled genes are genes from the top enriched GO-BP
terms selected for each cell-state comparison; `GO_BP` is the database label,
not one specific pathway term. The specific enriched terms are summarized in the
bar plot below.

```{r enrichment-plot, fig.width=8, fig.height=5}
EnrichmentPlot(
  pancreas_sub,
  group.by = "CellType",
  group_use = c("Ductal", "Endocrine"),
  plot_type = "bar"
)
```

## Export the Reproduction Object

The final object contains the upstream preprocessing state, trajectory metadata,
velocity output, dynamic genes, SCENIC regulon activity, differential genes, and
enrichment results. Export it for interactive review in SCExplorer.

```{r export-scexplorer}
scexplorer_dir <- file.path(tempdir(), "SCExplorer_pancreas_reproduction")

PrepareSCExplorer(
  list(pancreas_reproduction = pancreas_sub),
  base_dir = scexplorer_dir
)

head(list.files(scexplorer_dir), 10)
```

## Reproduction Summary

This compact reproduction should be read as a linked evidence trail:

- marker genes validate the major developmental states;
- Slingshot, Monocle3, and velocity provide directional support for the
  endocrine differentiation path;
- dynamic genes recover stage-specific programs along the pseudotime axis;
- GRNBoost2 and SCENIC prioritize TF-target links and regulon activity;
- DE and GO enrichment summarize the state-level transcriptional programs;
- SCExplorer keeps the full object available for interactive inspection.
