---
title: "Single-cell main workflow"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.align = "center",
  dpi = 144,
  out.width = "92%"
)
```

This article shows the recommended first-pass workflow for a single-cell Seurat
object in `scop`. It starts from a count matrix stored in a Seurat object, runs
the standard preprocessing workflow, checks cell-level quality, and creates the
basic embedding, feature, and group-level plots used before choosing more
specialized downstream methods.

Use this workflow as the default entry point when the goal is to inspect a new
single-cell dataset. After this pass, branch into integration, annotation,
trajectory, differential expression, enrichment, or interactive exploration.

This article deliberately stays at the first-pass level. It explains the object
state that downstream articles assume: which assay is active, which reductions
exist, which metadata columns can be used for grouping, and where `scop` stores
method-specific outputs.

## Load and Inspect the Object

The examples use the bundled mouse pancreas subset. It contains RNA, spliced,
and unspliced assays, so the same object can later be used for trajectory and
RNA-velocity examples.

```{r load-object}
library(scop)

data(pancreas_sub)
pancreas_sub
```

Before running a full workflow, check the active assay, available assays,
metadata columns, and whether a previous dimensional reduction already exists.
These checks prevent a common documentation mistake: plotting a reduction or
metadata column left over from a previous run without stating where it came
from.

```{r inspect-object}
SeuratObject::DefaultAssay(pancreas_sub)
SeuratObject::Assays(pancreas_sub)
head(pancreas_sub@meta.data)
SeuratObject::Reductions(pancreas_sub)
```

## Run the Standard Workflow

`standard_scop()` runs the common preprocessing path: normalization, variable
feature selection, scaling, dimensional reduction, neighbor graph construction,
clustering, and UMAP. The result stays in the same Seurat object.

```{r standard-workflow}
pancreas_sub <- standard_scop(
  pancreas_sub,
  assay = "RNA",
  nHVF = 2000,
  linear_reduction_dims = 50,
  cluster_resolution = 0.8,
  verbose = FALSE
)

pancreas_sub
SeuratObject::Reductions(pancreas_sub)
```

The stable contract is the Seurat object itself: normalized expression layers
are stored in the assay, reductions are stored in `@reductions`, metadata
columns are added to `@meta.data`, and method-specific summaries are stored
under `@tools` when a wrapper has additional results.

For this object, `CellType` is a coarse biological label and `SubCellType`
contains finer developmental states. Those labels are useful for examples, but
they should still be checked with marker genes before being used for biological
claims.

## Check Cell Quality

Run quality control before interpreting clusters. `RunCellQC()` combines common
cell-level flags, writes summary columns to metadata, and keeps details under
the object tools slot.

```{r cell-qc}
pancreas_sub <- RunCellQC(pancreas_sub)

table(pancreas_sub$CellQC)
names(pancreas_sub@tools)
```

Plot the final QC label first, then inspect the individual failure categories
when too many cells fail.

```{r qc-plots}
CellDimPlot(
  pancreas_sub,
  group.by = "CellQC",
  reduction = "UMAP"
)

CellStatPlot(
  pancreas_sub,
  stat.by = c(
    "db_qc", "decontX_qc", "outlier_qc",
    "umi_qc", "gene_qc", "mito_qc",
    "ribo_qc", "ribo_mito_ratio_qc", "species_qc"
  ),
  plot_type = "upset",
  stat_level = "Fail"
)
```

## Plot Cell Groups and Features

Use `CellDimPlot()` to inspect discrete annotations and cluster structure.
`FeatureDimPlot()` is the companion for gene expression or metadata-derived
numeric features.

```{r embedding-plots}
CellDimPlot(
  pancreas_sub,
  group.by = c("CellType", "SubCellType"),
  reduction = "UMAP",
  label = TRUE
)

FeatureDimPlot(
  pancreas_sub,
  features = c("Sox9", "Neurog3", "Fev", "Rbp4"),
  reduction = "UMAP"
)
```

When several marker genes define related states, `compare_features = TRUE`
helps identify the dominant feature per cell.
In the pancreas example, `Ins1`, `Gcg`, `Sst`, and `Ghrl` are hormone genes
used to distinguish beta, alpha, delta, and epsilon endocrine states.

```{r compare-features}
FeatureDimPlot(
  pancreas_sub,
  features = c("Ins1", "Gcg", "Sst", "Ghrl"),
  compare_features = TRUE,
  label = TRUE,
  label_insitu = TRUE,
  reduction = "UMAP"
)
```

## Summarize Marker Patterns by Group

Use grouped heatmaps after the embedding looks coherent. This is usually more
readable than inspecting many feature plots one by one.
Here the dot layer summarizes the fraction of cells expressing each marker, and
the heatmap color summarizes average expression by group. A marker should be
interpreted from both signals: strong average expression in very few cells is
different from broad expression across a group.

```{r group-heatmap}
ht <- GroupHeatmap(
  pancreas_sub,
  features = c(
    "Sox9", "Anxa2",
    "Neurog3", "Hes6",
    "Fev", "Neurod1",
    "Rbp4", "Pyy",
    "Ins1", "Gcg", "Sst", "Ghrl"
  ),
  group.by = c("CellType", "SubCellType"),
  heatmap_palette = "Spectral",
  cell_annotation = c("Phase", "G2M_score", "Cdh2"),
  show_row_names = TRUE,
  row_names_side = "left",
  add_dot = TRUE,
  add_reticle = TRUE
)
print(ht$plot)
```

## What to Report

A compact first-pass report should include:

- the object source, assay, and expression layer used;
- the cell and feature counts before and after filtering;
- the preprocessing parameters that affect interpretation, especially variable
  features, PCs, clustering resolution, and UMAP reduction name;
- QC pass or fail summaries and the dominant failure categories;
- the annotation column used for plots and any marker genes used to validate it;
- the output locations used downstream: assay layers, reductions, metadata, and
  `srt@tools`.

Keep this article as the short entry point. Use the specialized articles for
batch integration, annotation transfer, trajectories, differential expression,
pathway analysis, and deployment.
