---
title: "Single-cell main workflow"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  error = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 12, fig.height = 8,
  dpi = 144,
  out.width = "100%"
)
```

## What you will do

- **Question:** can I trust the input and recognize its main cell states?
- **Input:** All 1,000 cells and 15,998 supplied genes in the bundled mouse pancreas object. The package dataset is already a subset of the study; this article does not sample it again.
- **Output:** QC flags, normalized RNA, PCA/UMAP, clusters, marker plots and a reloadable object.
- **How to run:** start a fresh R session and run the chunks in order. Printed tables and figures below are generated from that run; code-only alternatives are identified explicitly.



Start here before integration, annotation, trajectories or pathway analysis.
The counts and QC tables establish what was measured; marker plots check the
meaning of labels; the saved object is the input for the specialist articles.

## Load and Inspect the Object

The examples use the bundled mouse pancreas subset. It contains RNA, spliced,
and unspliced assays, so the same object can later be used for trajectory and
RNA-velocity examples.

```{r load-object}
library(scop)
set.seed(11)

data(pancreas_sub)
pancreas_sub
input_cells <- colnames(pancreas_sub)
counts <- GetAssayData5(pancreas_sub, assay = "RNA", layer = "counts")
stopifnot(identical(colnames(counts), rownames(pancreas_sub@meta.data)),
          !anyDuplicated(colnames(counts)), all(counts@x >= 0))
data.frame(genes = nrow(counts), cells = ncol(counts),
           mitochondrial_genes = sum(grepl("^mt-", rownames(counts))))
```

Before running a full workflow, check the active assay, available assays,
metadata columns, and whether a previous dimensional reduction already exists.
These checks prevent a common documentation mistake: plotting a reduction or
metadata column left over from a previous run without stating where it came
from.

```{r inspect-object}
SeuratObject::DefaultAssay(pancreas_sub)
SeuratObject::Assays(pancreas_sub)
head(pancreas_sub@meta.data)
SeuratObject::Reductions(pancreas_sub)
```

## Run the Standard Workflow

`RunStandardWorkflow()` runs the common preprocessing path: normalization, variable
feature selection, scaling, dimensional reduction, neighbor graph construction,
clustering, and UMAP. The result stays in the same Seurat object.

```{r standard-workflow}
pancreas_sub <- RunStandardWorkflow(
  pancreas_sub,
  assay = "RNA",
  nHVF = 2000,
  linear_reduction_dims = 50,
  cluster_resolution = 0.8,
  seed = 11,
  cores = 1,
  verbose = FALSE
)

pancreas_sub
SeuratObject::Reductions(pancreas_sub)
```

The stable contract is the Seurat object itself: normalized expression layers
are stored in the assay, reductions are stored in `@reductions`, metadata
columns are added to `@meta.data`, and method-specific summaries are stored
under `@tools` when a wrapper has additional results.

For this object, `CellType` is a coarse biological label and `SubCellType`
contains finer developmental states. Those labels are useful for examples, but
they should still be checked with marker genes before being used for biological
claims.

## Check Cell Quality

Run quality control before interpreting clusters. This first pass measures
UMIs, detected genes and mitochondrial fraction on the supplied RNA counts.
It annotates every cell, without silently removing failures. The object is
already feature-filtered: these totals are not original-library QC.
Doublet and ambient-RNA models require separate dependencies and are not part
of this basic receipt. For production data, run QC on the unfiltered count
assay, justify thresholds and rerun preprocessing after any filtering.

```{r cell-qc}
pancreas_sub <- RunCellQC(
  pancreas_sub, assay = "RNA",
  qc_metrics = c("umi", "gene", "mito"),
  UMI_threshold = 3000, gene_threshold = 1000, mito_threshold = 20,
  return_filtered = FALSE, verbose = FALSE
)
stopifnot(identical(colnames(pancreas_sub), input_cells))

table(pancreas_sub$CellQC)
names(pancreas_sub@tools)
data.frame(
  criterion = c("umi_qc", "gene_qc", "mito_qc"),
  failures = vapply(pancreas_sub@meta.data[c("umi_qc", "gene_qc", "mito_qc")],
                    function(x) sum(x == "Fail", na.rm = TRUE), integer(1))
)
```

Plot the final QC label first, then inspect the individual failure categories
when too many cells fail.

```{r qc-plots}
CellDimPlot(
  pancreas_sub,
  group.by = "CellQC",
  reduction = "UMAP"
)

qc_failures <- data.frame(
  metric = c("UMI", "Genes", "Mitochondrial fraction"),
  failures = vapply(pancreas_sub@meta.data[c("umi_qc", "gene_qc", "mito_qc")],
                    function(x) sum(x == "Fail", na.rm = TRUE), integer(1))
)
ggplot2::ggplot(qc_failures, ggplot2::aes(metric, failures)) +
  ggplot2::geom_col() +
  ggplot2::geom_text(ggplot2::aes(label = failures), vjust = -0.5) +
  ggplot2::scale_y_continuous(limits = c(0, max(1, qc_failures$failures) * 1.1),
                            breaks = scales::breaks_pretty(n = 4)) +
  ggplot2::labs(x = NULL, y = "Cells flagged",
                subtitle = "Zero means no cell failed that criterion") +
  ggplot2::theme_minimal(base_size = 12)
```

## Plot Cell Groups and Features

Use `CellDimPlot()` to inspect discrete annotations and cluster structure.
`FeatureDimPlot()` is the companion for gene expression or metadata-derived
numeric features.

```{r embedding-plots}
CellDimPlot(
  pancreas_sub,
  group.by = c("CellType", "SubCellType"),
  reduction = "UMAP",
  label = FALSE, theme_use = "theme_blank"
)

FeatureDimPlot(
  pancreas_sub,
  features = c("Sox9", "Neurog3", "Fev", "Rbp4"),
  reduction = "UMAP"
)
```

When several marker genes define related states, `compare_features = TRUE`
blends feature-specific colours to show overlapping expression. It does not
identify the highest-abundance gene or assign a cell type.
In the pancreas example, `Ins1`, `Gcg`, `Sst`, and `Ghrl` are hormone genes
used to distinguish beta, alpha, delta, and epsilon endocrine states.

```{r compare-features}
FeatureDimPlot(
  pancreas_sub,
  features = c("Ins1", "Gcg", "Sst", "Ghrl"),
  compare_features = TRUE,
  label = TRUE,
  reduction = "UMAP"
)
```

## Summarize Marker Patterns by Group

Use grouped heatmaps after the embedding looks coherent. This is usually more
readable than inspecting many feature plots one by one.
Here dot size summarizes the fraction of cells expressing each marker. Colour
shows row-standardized group-average expression: compare groups within a gene,
not absolute expression between genes. A marker should be
interpreted from both signals: strong average expression in very few cells is
different from broad expression across a group.

```{r group-heatmap}
ht <- GroupHeatmap(
  pancreas_sub,
  features = c(
    "Sox9", "Anxa2",
    "Neurog3", "Hes6",
    "Fev", "Neurod1",
    "Rbp4", "Pyy",
    "Ins1", "Gcg", "Sst", "Ghrl"
  ),
  group.by = c("CellType", "SubCellType"),
  heatmap_palette = "Spectral",
  cell_annotation = c("Phase", "G2M_score", "Cdh2"),
  show_row_names = TRUE, nlabel = 0,
  row_names_side = "left",
  add_dot = TRUE,
  add_reticle = TRUE
)
print(ht$plot)
```



## Save, reload, and check the hand-off

The build uses a temporary directory so repeated rendering cannot overwrite a
previous analysis. For your work, replace it with a new persistent output folder.
Save the object **and** plain-text metadata; a figure alone cannot reproduce the
analysis. No cells are sampled for this export.

```{r export-reload}
output_dir <- file.path(tempdir(), "single-cell-main-workflow")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
result_file <- file.path(output_dir, "analysis.rds")
saveRDS(pancreas_sub, result_file)
utils::write.csv(pancreas_sub@meta.data, file.path(output_dir, "cell-metadata.csv"))
reloaded <- readRDS(result_file)
stopifnot(identical(colnames(reloaded), colnames(pancreas_sub)),
          identical(reloaded@meta.data, pancreas_sub@meta.data),
          identical(SeuratObject::Reductions(reloaded),
                    SeuratObject::Reductions(pancreas_sub)))
data.frame(file = basename(list.files(output_dir, full.names = TRUE)),
           bytes = file.info(list.files(output_dir, full.names = TRUE))$size)
```

## Use your own data and recover common errors

- Replace the `data(...)` input with `readRDS("your-object.rds")`, or create a
  Seurat object from a gene-by-cell raw-count matrix. Check that cell IDs are
  unique and metadata row names equal the expression column names.
- Set the assay and grouping columns explicitly. Species, gene identifiers and
  reference labels must match your experiment; do not copy pancreas labels to
  unrelated tissue.
- If a reduction or metadata column is missing, inspect
  `SeuratObject::Reductions(object)` and `colnames(object@meta.data)`, then run
  its producer chunk. A plot does not create the result it needs.
- If a dependency is unavailable, install it in the analysis environment,
  restart R and rerun from the input chunk. A skipped or failed method is not
  evidence that it worked. Preserve warnings and the exact error in your report.

## Build and analysis provenance

The receipt records the R seed. Python methods retain their documented backend
seeds and parameters; setting the R seed alone does not control every backend.
Package versions, BLAS libraries and backend changes can alter numerical results. Keep
this receipt with the exported object. It describes this page's actual render.

```{r provenance}
source_revision <- Sys.getenv("GITHUB_SHA")
if (!nzchar(source_revision) && nzchar(Sys.which("git"))) {
  source_revision <- tryCatch(system2("git", c("rev-parse", "HEAD"), stdout = TRUE,
                                      stderr = FALSE)[1], error = function(e) NA_character_)
}
article_source <- knitr::current_input(dir = TRUE)
build_receipt <- data.frame(
  article = "single-cell-main-workflow",
  built_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
  scop = as.character(utils::packageVersion("scop")),
  R = as.character(getRversion()),
  source_revision = source_revision,
  source_has_local_changes = if (nzchar(Sys.which("git")))
    length(system2("git", c("status", "--porcelain"), stdout = TRUE, stderr = FALSE)) > 0 else NA,
  article_md5 = unname(tools::md5sum(article_source)),
  R_seed = 11,
  cells_retained = ncol(pancreas_sub),
  genes_in_active_assay = nrow(pancreas_sub),
  stringsAsFactors = FALSE
)
build_receipt
utils::write.csv(build_receipt, file.path(output_dir, "build-receipt.csv"),
                 row.names = FALSE)
writeLines(capture.output(sessionInfo()), file.path(output_dir, "session-info.txt"))
```

<details>
<summary>Show the complete R session used for this page</summary>

```{r session-info}
sessionInfo()
```

</details>

## What to Report

A compact first-pass report should include:

- the object source, assay, and expression layer used;
- the cell and feature counts before and after filtering;
- the preprocessing parameters that affect interpretation, especially variable
  features, PCs, clustering resolution, and UMAP reduction name;
- QC pass or fail summaries and the dominant failure categories;
- the annotation column used for plots and any marker genes used to validate it;
- the output locations used downstream: assay layers, reductions, metadata, and
  `srt@tools`.

Keep this article as the short entry point. Use the specialized articles for
batch integration, annotation transfer, trajectories, differential expression,
pathway analysis, and deployment.
