---
title: "Spatial framework bridges"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This article documents SCOP's lightweight bridge to Giotto. The bridge moves a
single spatial image between Seurat and Giotto; it does not wrap Giotto's
analysis or plotting workflow and it does not merge slices implicitly.

## Choose the interface

| Goal | Interface | Returned object | Plotting |
|---|---|---|---|
| Move one Seurat image into Giotto | `srt_to_giotto()` | Native Giotto object | Giotto's native plotting API |
| Move a native Giotto object into Seurat | `giotto_to_srt()` | New Seurat object | scop or Seurat plotting |

## Giotto version boundary

Current Giotto Suite releases split analysis functions and object/conversion
helpers across the `Giotto` and `GiottoClass` namespaces. scop resolves these
optional functions at runtime and does not add Giotto to Imports, Depends, or
Suggests.

The forward bridge selects `seuratToGiottoV5()` for Seurat 5 and
`seuratToGiottoV4()` for Seurat 4. The reverse bridge selects
`giottoToSeuratV5()` for Seurat 5 and `giottoToSeuratV4()` for Seurat 4.
These converters are expected from a current GiottoClass installation. If an
older Giotto installation does not provide them, update Giotto before using
the bridges.

## Single-image conversion

The bridges process exactly one image at a time:

- a single image is selected automatically;
- a multi-image object requires an explicit `image`;
- an unknown image name is rejected;
- slices are never merged implicitly.

```{r giotto-roundtrip}
data(visium_human_pancreas_sub)

spatial <- Seurat::NormalizeData(
  visium_human_pancreas_sub,
  assay = "Spatial",
  verbose = FALSE
)

gobject <- srt_to_giotto(spatial, image = "slice1")
spatial_roundtrip <- giotto_to_srt(gobject)
```

After conversion, inspect the returned object with the target framework's own
accessors. SCOP retains responsibility for Seurat-side analysis and plotting;
Giotto retains responsibility for Giotto-side analysis and plotting.
