---
title: "Spatial domain benchmark workflow"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.align = "center",
  dpi = 144,
  out.width = "92%"
)
```

This article compares spatial domain clustering methods with
[RunSpatialBenchmark()]. Each method runs in an isolated R process; results
include ARI, NMI, purity, runtime, and peak memory against a gold-standard
label column. Visualize the result with [SpatialBenchmarkPlot()].

Supported producers include `BayesSpace`, `BANKSY`, and `SmoothClust`. This
workflow is separate from scRNA batch integration benchmarking
([RunIntegrationBenchmark()]).

## Load data and define a gold standard

The bundled Visium pancreas subset is used here. For demonstration, a simple
cyclic domain label is assigned as the gold standard. Replace this with manual
annotations or reference domains for real analyses.

```{r load-data}
library(scop)

data(visium_human_pancreas_sub)
spatial <- visium_human_pancreas_sub

spatial$gold_domain <- factor(
  paste0("domain_", (seq_len(ncol(spatial)) - 1) %% 3 + 1)
)
table(spatial$gold_domain)
```

## Run the benchmark

Methods share the same spot set. Tune per-method arguments through
`method_params` when backends need different PCA counts or layers.

```{r run-benchmark}
bench <- RunSpatialBenchmark(
  spatial,
  gold_standard = "gold_domain",
  method_params = list(
    BayesSpace = list(n.PCs = 5, n.HVGs = 200),
    BANKSY = list(layer = "counts"),
    SmoothClust = list(layer = "counts", min_spots = 1)
  ),
  verbose = FALSE
)

bench$summary
```

## Inspect the result object

`RunSpatialBenchmark()` returns a `spatial_benchmark_result` list:

- `$summary`: quality and resource metrics per method;
- `$predictions`: aligned domain labels per spot;
- `$metrics`: long metric table for custom plots.

```{r inspect-result}
str(bench$summary, max.level = 1)
head(bench$predictions)
```

## Plot benchmark views

`SpatialBenchmarkPlot()` defaults to a publication-oriented overview that
combines clustering agreement with runtime and peak-memory efficiency.

```{r plot-overview, fig.height=6}
SpatialBenchmarkPlot(data = bench, plot_type = "overview")
```

Other views:

```r
SpatialBenchmarkPlot(data = bench, plot_type = "quality")
SpatialBenchmarkPlot(data = bench, plot_type = "efficiency")
SpatialBenchmarkPlot(data = bench, plot_type = "heatmap")
```

## What to report

A spatial benchmark report should include:

- the gold-standard column and number of domains;
- methods compared and any `method_params` overrides;
- `$summary` with ARI, NMI, purity, runtime, and peak memory;
- the overview or quality plot;
- failed or unavailable methods (status column) when backends are missing.

Agreement metrics are higher when better; runtime and peak memory are lower
when better.
