---
title: "Spatial framework bridges"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center",
                      dpi = 144, out.width = "92%")
```

Use this page when an analysis needs a different data container. You will
convert a real Visium dataset to SpatialExperiment, verify what survived,
convert back to Seurat, and save a reloadable result. A separate Giotto section
checks the installed backend before attempting the same task. Conversion does
not run either framework's biological analysis.

Start with the [main spatial workflow](spatial-main-workflow.html) for analysis,
or the [platform page](spatial-platform-workflows.html) for vendor imports and
assay/image choices.

## Inputs and preservation boundaries

The executable input is the complete bundled `visium_human_pancreas_sub`
object: 5,000 genes and all 1,986 packaged tissue spots from human PanIN sample
GSM8058244, study [GSE254829](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE254829).
The package has already selected genes shared with `panc8_sub`; this article
performs no additional feature or spot subsampling. `coda_label` is supplied
microanatomical annotation, not a predicted cell type.

| Field | Seurat ↔ SpatialExperiment | Seurat ↔ Giotto |
|---|---|---|
| Expression | One selected assay/layer becomes `scop_input`; reverse conversion places it in Seurat `counts` | The official GiottoClass converter controls layer and assay naming |
| IDs and metadata | Expression columns and coordinate rows are aligned by ID; `include_meta = TRUE` exports metadata | Check returned IDs and metadata against the selected input; do not assume names are unchanged |
| Coordinates | Raw coordinates by default; an explicit legacy display export stores an invertible transform | Official converter's coordinate convention; validate before distance-based analysis |
| Other Seurat content | Does not round-trip images, reductions, graphs, tools, or all assays | Retention follows the installed GiottoClass converter |
| Analysis | No target-framework analysis is run | No Giotto analysis or Python environment is initialized |

The reverse SpatialExperiment bridge treats its selected expression assay as
counts. Therefore this round-trip exports raw counts. Do not export a
normalized layer and then describe the returned `counts` layer as raw data.

```{r bridge-input}
library(scop)
stopifnot(requireNamespace("SpatialExperiment", quietly = TRUE))
data(visium_human_pancreas_sub)
spatial <- visium_human_pancreas_sub
assay_name <- "Spatial"
image_name <- "slice1"
counts_before <- GetAssayData5(spatial, assay = assay_name, layer = "counts")
raw_before <- SpatialCoordinates(spatial, image = image_name, space = "raw")

stopifnot(
  identical(dim(counts_before), c(5000L, 1986L)),
  !anyDuplicated(colnames(counts_before)),
  setequal(colnames(counts_before), raw_before$data$cell_id)
)
raw_xy <- as.matrix(raw_before$data[colnames(counts_before), c("x", "y")])
input_receipt <- data.frame(
  source = "GSE254829 / GSM8058244 / complete bundled object",
  features = nrow(counts_before), observations = ncol(counts_before),
  assay = assay_name, layer = "counts", image = image_name,
  coordinate_space = "raw", coordinate_units = "full-resolution pixels"
)
input_receipt
raw_before$source
```

The dimension assertion protects this particular example. Replace it with your
own expected dimensions when adapting the code. The ID check is general: stop
and resolve any mismatch instead of assigning coordinates by row position.

## Round-trip through SpatialExperiment

```{r spe-roundtrip}
spe <- srt_to_spe(
  spatial, assay = assay_name, layer = "counts", image = image_name,
  include_meta = TRUE, coordinate_space = "raw"
)
spatial_roundtrip <- spe_to_srt(spe, assay = assay_name, layer = "scop_input")
counts_after <- GetAssayData5(spatial_roundtrip, assay = assay_name, layer = "counts")

# Coordinate columns are checked separately because reverse conversion writes x/y.
metadata_columns <- setdiff(colnames(spatial@meta.data), c("x", "y"))
spe_checks <- c(
  observation_order = identical(colnames(spatial_roundtrip), colnames(counts_before)),
  feature_order = identical(rownames(counts_after), rownames(counts_before)),
  counts = isTRUE(all.equal(counts_after, counts_before)),
  metadata = isTRUE(all.equal(
    spatial_roundtrip@meta.data[, metadata_columns, drop = FALSE],
    spatial@meta.data[, metadata_columns, drop = FALSE]
  )),
  raw_coordinates = isTRUE(all.equal(
    as.matrix(spatial_roundtrip@meta.data[, c("x", "y")]), raw_xy,
    check.attributes = FALSE
  )),
  provenance = identical(
    spatial_roundtrip@misc$spatial_coordinate_import,
    S4Vectors::metadata(spe)$scop_spatial_coordinates
  )
)
data.frame(check = names(spe_checks), passed = unname(spe_checks))
stopifnot(all(spe_checks))
SummarizedExperiment::assayNames(spe)
SeuratObject::Images(spatial_roundtrip)
```

Every displayed check must be `TRUE`. An empty image list is expected: the
lightweight conversion keeps coordinates in metadata but does not reconstruct
the raster. The selected count matrix and original metadata are checked,
not just the number of spots. Existing Seurat analysis results are outside
this bridge's contract; keep the original object if you need them.

```{r spe-map, fig.height=4.5}
stopifnot("GCG" %in% rownames(counts_after))
SpatialSpotPlot(
  spatial_roundtrip, image = NULL, coord.cols = c("x", "y"),
  features = "GCG", assay = assay_name, layer = "counts",
  overlay_image = FALSE, palette = "YlGnBu", legend.title = "GCG counts"
)
```

This is the measured GCG count map after conversion, drawn from preserved raw
coordinates without a tissue raster. A spatially localized signal is not by
itself a cell-type assignment, and the count range is not normalized expression.

### Verify a legacy display export can be inverted

Use raw coordinates for new distance-sensitive work. This compatibility check
is useful when an older workflow explicitly expects display-scaled and
vertically flipped coordinates.

```{r spe-display-roundtrip}
spe_display <- srt_to_spe(
  spatial, assay = assay_name, layer = "counts", image = image_name,
  coordinate_space = "legacy_display"
)
display_provenance <- S4Vectors::metadata(spe_display)$scop_spatial_coordinates
display_provenance
from_display <- spe_to_srt(spe_display, assay = assay_name, layer = "scop_input")
display_error <- max(abs(
  as.matrix(from_display@meta.data[, c("x", "y")]) - raw_xy
))
data.frame(max_raw_coordinate_error = display_error, tolerance = 1e-8)
stopifnot(is.finite(display_error), display_error < 1e-8)
```

The error is measured in original pixel units and should be zero or numerical
round-off. Preserve the saved transform when passing this object to another
person. External SpatialExperiment objects without SCOP provenance are treated
as raw coordinates in their supplied units; verify those units yourself.

## Optional: round-trip through Giotto

SCOP chooses the official `GiottoClass` v4/v5 converters for the installed
Seurat major version. A single image can be resolved automatically; multiple
images require `image =`. Normalize before export because the official reverse
converter needs a nonempty normalized layer. This does not replace raw counts.

The following code is evaluated: it records backend availability, runs the
conversion if compatible converters are installed, and reports any error.
It never installs a backend during rendering. A missing backend is reported
as `not_run`, and a failed preservation check as `failed`, never `completed`.
Install GiottoClass separately following the
[official Giotto instructions](https://github.com/giotto-suite/GiottoClass)
and rerun this section to validate your exact package versions.

```{r giotto-roundtrip}
seurat_major <- as.integer(strsplit(as.character(packageVersion("Seurat")), "\\.")[[1]][1])
converter_suffix <- if (seurat_major >= 5L) "V5" else "V4"
converters <- paste0(c("seuratToGiotto", "giottoToSeurat"), converter_suffix)
giotto_available <- requireNamespace("GiottoClass", quietly = TRUE)
giotto_compatible <- giotto_available && all(vapply(
  converters, exists, logical(1), envir = asNamespace("GiottoClass"), inherits = FALSE
))
giotto_result <- NULL
giotto_receipt <- data.frame(
  backend = "GiottoClass",
  version = if (giotto_available) as.character(packageVersion("GiottoClass")) else NA_character_,
  status = "not_run",
  reason = if (!giotto_available) "GiottoClass is not installed" else
    if (!giotto_compatible) "Required converters are absent" else "Ready to attempt conversion"
)

if (giotto_compatible) {
  giotto_result <- tryCatch({
    giotto_input <- Seurat::NormalizeData(spatial, assay = assay_name, verbose = FALSE)
    native <- srt_to_giotto(giotto_input, image = image_name)
    restored <- giotto_to_srt(native)
    restored_assay <- SeuratObject::DefaultAssay(restored)
    restored_counts <- GetAssayData5(restored, assay = restored_assay, layer = "counts")
    stopifnot(setequal(colnames(restored_counts), colnames(counts_before)),
              setequal(rownames(restored_counts), rownames(counts_before)))
    restored_counts <- restored_counts[rownames(counts_before), colnames(counts_before)]
    restored_xy <- SpatialCoordinates(restored, space = "raw")$data
    checks <- c(
      counts = isTRUE(all.equal(restored_counts, counts_before)),
      coda_annotation = "coda_label" %in% colnames(restored@meta.data) &&
        identical(as.character(restored@meta.data[colnames(counts_before), "coda_label"]),
                  as.character(spatial$coda_label)),
      coordinates = isTRUE(all.equal(
        as.matrix(restored_xy[colnames(counts_before), c("x", "y")]), raw_xy,
        check.attributes = FALSE
      ))
    )
    print(data.frame(check = names(checks), passed = unname(checks)))
    stopifnot(all(checks))
    list(native = native, restored = restored, assay = restored_assay, checks = checks,
         metadata_columns_retained = intersect(colnames(spatial@meta.data), colnames(restored@meta.data)),
         metadata_columns_missing = setdiff(colnames(spatial@meta.data), colnames(restored@meta.data)))
  }, error = function(e) e)
  if (inherits(giotto_result, "error")) {
    giotto_receipt$status <- "failed"
    giotto_receipt$reason <- conditionMessage(giotto_result)
    giotto_result <- NULL
  } else {
    giotto_receipt$status <- "completed"
    giotto_receipt$reason <- "Conversion and count, annotation, and coordinate checks passed"
    print(giotto_result[c("metadata_columns_retained", "metadata_columns_missing")])
  }
}
giotto_receipt
```

A successful conversion receipt verifies those named checks only. Read the
metadata retention lists before relying on other columns. If this section
reports `failed`, resolve its message and rerun before using the converted
object for analysis. Even `completed` does not mean that Giotto's own analysis
or plotting pipeline has run.

## Save, reload, and recheck

Save both the portable object and the receipt. The example uses a temporary
directory so a build does not overwrite your work. For a persistent analysis,
set `SCOP_BRIDGE_OUTPUT_DIR` to a new project directory before running it.

```{r save-bridge}
output_dir <- Sys.getenv("SCOP_BRIDGE_OUTPUT_DIR", file.path(tempdir(), "scop-bridges"))
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
bridge_file <- file.path(output_dir, "panin-spatialexperiment-roundtrip.rds")
bridge_bundle <- list(
  spatial_experiment = spe, seurat = spatial_roundtrip,
  input = input_receipt, checks = spe_checks, giotto_receipt = giotto_receipt,
  giotto = giotto_result, session = utils::sessionInfo()
)
saveRDS(bridge_bundle, bridge_file)
reloaded <- readRDS(bridge_file)
reload_checks <- c(
  file_exists = file.exists(bridge_file),
  observations = identical(colnames(reloaded$seurat), colnames(spatial)),
  counts = isTRUE(all.equal(
    GetAssayData5(reloaded$seurat, assay = assay_name, layer = "counts"), counts_before
  )),
  coordinates = identical(SpatialExperiment::spatialCoords(reloaded$spatial_experiment),
                         SpatialExperiment::spatialCoords(spe)),
  receipt = identical(reloaded$checks, spe_checks)
)
data.frame(check = names(reload_checks), passed = unname(reload_checks))
stopifnot(all(reload_checks))
```

The saved `seurat` object can be passed directly to the earlier plot call.
Reloading or plotting does not rerun conversion, normalization, or a model.

## Adapt to your data and recover from errors

1. Replace the `data()` input with your imported Seurat object and choose one
   image with `SeuratObject::Images()`. Select its matching assay and inspect
   `SeuratObject::Layers()` before choosing `counts`.
2. Keep the entire selected image unless exclusion has a scientific/QC reason.
   With a multi-image source, derive expected IDs from the selected image and
   compare the exported object with that same image's expression columns.
3. Replace the fixed example dimension check and `GCG` with your expected
   dimensions and an actual feature. Keep the ID, count, and coordinate checks.
4. Record platform, units, sample, assay/layer, image, input selection, package
   versions, and whether each bridge completed. Keep the original Seurat object
   for images and analysis fields that do not cross the bridge.

| Symptom | Check and recovery |
|---|---|
| SpatialExperiment is missing | Install it with Bioconductor in your analysis environment, restart R, and rerun the input section |
| Multiple spatial images | Supply one valid `image =`; export each sample separately rather than merging unrelated coordinate systems |
| IDs do not match | Compare `colnames()` and coordinate row names; align by verified IDs, never rename by row position |
| An assay/layer is missing | Inspect `Assays()`/`Layers()`; restore the intended raw count source instead of relabeling normalized data |
| Display import needs a scale or height | Restore the original SCOP provenance or export raw coordinates again; do not invent a transform |
| Giotto conversion fails on normalized data | Run `NormalizeData()` on the input count assay and check compatible converter exports and versions |
| No image after the SPE round-trip | Expected: use metadata coordinates with `image = NULL`, or retain the original Seurat object when a raster is required |

```{r bridge-session}
data.frame(
  package = c("scop", "Seurat", "SeuratObject", "SpatialExperiment"),
  version = vapply(c("scop", "Seurat", "SeuratObject", "SpatialExperiment"),
                   function(p) as.character(utils::packageVersion(p)), character(1))
)
```

## Build provenance

This receipt belongs to the actual rendered source and retained observations.
Save it beside the exported object when adapting the workflow.

```{r build-provenance}
source_revision <- Sys.getenv("GITHUB_SHA")
if (!nzchar(source_revision) && nzchar(Sys.which("git"))) {
  source_revision <- system2("git", c("rev-parse", "HEAD"), stdout = TRUE,
                             stderr = FALSE)[1]
}
build_receipt <- data.frame(
  article = "spatial-framework-bridges", built_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
  source_revision = source_revision,
  source_has_local_changes = if (nzchar(Sys.which("git")))
    length(system2("git", c("status", "--porcelain"), stdout = TRUE,
                   stderr = FALSE)) > 0 else NA,
  article_md5 = unname(tools::md5sum(knitr::current_input(dir = TRUE))),
  R = as.character(getRversion()), scop = as.character(utils::packageVersion("scop")),
  retained_observations = ncol(spatial_roundtrip), genes = nrow(spatial_roundtrip)
)
build_receipt
utils::write.csv(build_receipt, file.path(output_dir, "build-receipt.csv"), row.names = FALSE)
writeLines(capture.output(sessionInfo()), file.path(output_dir, "session-info.txt"))
```
