---
title: "Spatial transcriptomics main workflow"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.align = "center",
  dpi = 144,
  out.width = "92%"
)
```

This is the short, result-first path for a spatial Seurat object: inspect the
input, run a small baseline workflow, inspect the stored result, and plot it.
The bundled `visium_human_pancreas_sub` object is a real Visium spot-level
dataset used for the executable examples below. It is not a single-cell
measurement: one spot can contain transcripts from several cells.

Use [Spatial imports and standard workflow methods](spatial-platform-workflows.html)
for Visium/Visium HD/Xenium input choices, assays, units, and backend-specific
parameters. Use [Spatial framework bridges](spatial-framework-bridges.html)
for Seurat, SpatialExperiment, and Giotto conversion boundaries.

## Check the input before analysis

```{r load-spatial}
library(scop)

data(visium_human_pancreas_sub)
spatial <- visium_human_pancreas_sub
SeuratObject::DefaultAssay(spatial) <- "Spatial"

SeuratObject::Images(spatial)
head(spatial@meta.data[, c("x", "y", "coda_label")])
raw_coordinates <- SpatialCoordinates(
  spatial,
  image = "slice1",
  space = "raw"
)
raw_coordinates$source
```

The first checks should answer four questions: which assay contains the counts,
which image is selected, which columns identify the observations, and whether
the coordinate IDs match the expression columns. SCOP uses raw acquisition
coordinates for distance-sensitive analysis. `image.scale` is a display choice
for a selected raster; it does not change distances or rewrite Seurat scale
factors.

For Visium, the observations are spots on an acquisition grid. For Xenium or
other segmented assays, the observations are cells and the right visualization
is [SpatialCellPlot()], not a spot-composition interpretation. A spot label is
therefore not automatically a cell type.

## Draw an input map

```{r input-map, fig.height=4.5}
SpatialSpotPlot(
  spatial,
  group.by = "coda_label",
  image = "slice1",
  palcolor = c(
    acini = "#E69F00",
    collagen = "#56B4E9",
    fat = "#009E73",
    islets = "#CC79A7",
    `normal epithelium` = "#0072B2",
    panin = "#D55E00",
    `smooth muscle` = "#999999"
  )
)
```

The map is an input and metadata check, not a result from a clustering or
deconvolution model. Named palettes are matched to category names, so changing
the row order or subsetting spots does not silently change the biological color
meaning. `SpatialSpotPlot()` keeps equal coordinate scaling by default.

## Run the baseline spatial workflow

Keep the first run small and interpretable. This real Visium example performs
spot QC and native Moran spatial-variable-feature scoring. It does not run an
optional deconvolution or domain backend.

```{r standard-spatial}
spatial <- RunStandardWorkflow(
  spatial,
  workflow = "spatial",
  assay = "Spatial",
  image = "slice1",
  coord.cols = c("x", "y"),
  do_spot_qc = TRUE,
  do_spatial_variable_features = TRUE,
  spatial_variable_features_params = list(
    method = "moran",
    nfeatures = 50,
    nperm = 0
  ),
  do_spatial_cluster = FALSE,
  do_deconvolution = FALSE,
  verbose = FALSE
)
```

## Inspect stored results before plotting

```{r inspect-results}
workflow <- spatial@tools$run_standard_spatial_workflow
workflow$status
workflow$stages[, c("stage", "requested", "status", "actual_method", "reason")]
spatial@tools$SpatialVariableFeatures$summary
head(spatial@tools$SpatialVariableFeatures$summary$top_features)
```

The stage table is the execution receipt. `completed`, `skipped`, and `failed`
are different states; an unavailable optional backend is not relabelled as a
successful result. A spatially variable feature is a candidate spatial pattern,
not automatically a domain marker or a cell-type marker.

## Plot QC and spatial features

```{r plot-results, fig.height=4.5}
SpatialSpotPlot(spatial, group.by = "SpotQC", image = "slice1")
SpatialVariableFeaturePlot(
  spatial,
  plot_type = "summary",
  nfeatures = 4
)
SpatialSpotPlot(
  spatial,
  features = head(spatial@tools$SpatialVariableFeatures$summary$top_features, 2),
  assay = "Spatial",
  layer = "counts",
  image = "slice1"
)
```

`combine = FALSE` returns named plots for independent editing, including
long-format point maps and dominant deconvolution maps. Long-format `plot.data`
accepts explicit cutoffs and quantiles; its default range remains the full data
range. Plotting reads stored results or the explicitly selected assay/layer;
it does not rerun QC or spatial-feature detection. A saved Seurat
object can be reloaded and plotted again as long as the stored result and the
selected spatial context remain complete.

## Proportion and empty-panel display

Composition plots require a spot-by-cell-type result. The following compact
fixture deliberately creates a display-only matrix from real spot IDs so that
the layout contract is executable without installing an optional deconvolution
backend. It is not a biological inference and must not be reported as one.

```{r composition-layout, fig.width=10, fig.height=4.5}
display_object <- spatial[, unique(round(seq(1, ncol(spatial), length.out = 120)))]
fraction <- seq(0, 1, length.out = ncol(display_object))
demo_proportions <- cbind(component_A = fraction, component_B = 1 - fraction,
                         unmeasured = NA_real_)
rownames(demo_proportions) <- colnames(display_object)
display_object@tools$display_only_proportions <- list(proportions = demo_proportions)

SpatialDeconvolutionPlot(
  display_object,
  tool_name = "display_only_proportions",
  combine = TRUE,
  overlay_image = FALSE,
  ncol = 3,
  palette = "YlGnBu",
  legend.position = "bottom",
  legend.direction = "horizontal"
) + patchwork::plot_annotation(title = "Display-only matrix example")
```

In a real analysis, replace that fixture with the stored output of
`RunRCTD()`, `RunSPOTlight()`, or another supported producer and keep the
producer's provenance. The common matrix renderer keeps numeric values and
colors identical between `combine = TRUE` and `combine = FALSE`; all-missing
panels remain informative empty panels. q05 abundance from cell2location is a
different quantity from a proportion and is not forced onto a 0--1 scale.

For a proportion pie, use the explicit numeric matrix with
`SpatialSpotPlot(..., plot_type = "pie")` when `scatterpie` is installed. Pie
fractions are relative to the selected components; they are not cell counts.

## Multiple samples and optional backends

`split.by` is useful for a quick faceted view of metadata:

```{r sample-facets, fig.height=4.5}
spatial$display_sample <- rep(c("sample-A", "sample-B"), length.out = ncol(spatial))
SpatialSpotPlot(
  spatial,
  group.by = "coda_label",
  split.by = "display_sample",
  overlay_image = FALSE
)
```

This facet does not register images or create a cross-sample inference. A
multi-image analysis must select one covering image per sample; ambiguous
coverage errors instead of dropping a sample. The platform page contains the
`RunSpatialIntegration()` example with an explicit named image map and labels
the PRECAST dependency as an execution condition.

Domain clustering, deconvolution, neighborhood, communication, and framework
conversion are separate questions. Add one only after the baseline map and
stage receipt are sensible, then report the observation unit, assay, image,
coordinate columns, method, and result status.

## Save, reload, and redraw

Save the analysis object, not only its figure. This example uses a temporary
file; choose a persistent filename when keeping your own results. The real
analysis object is separate from the display-only fixture above.

```{r save-reload, fig.height=4.5}
result_file <- tempfile(fileext = ".rds")
saveRDS(spatial, result_file)
reloaded <- readRDS(result_file)
stopifnot(identical(colnames(reloaded), colnames(spatial)))
stopifnot(identical(reloaded@tools$SpatialVariableFeatures,
                    spatial@tools$SpatialVariableFeatures))
SpatialSpotPlot(reloaded, group.by = "SpotQC", image = "slice1")
unlink(result_file)
```

The redraw does not rerun QC, normalization, or spatial-feature detection.

## What to report

At minimum record:

- the source object and platform, assay/layer, selected image, coordinate
  columns, and coordinate space;
- the number of spots or cells retained at each stage;
- QC status and the selected spatial-variable-feature method;
- domain or composition methods only when they were actually run;
- backend availability and failed/skipped stages, rather than presenting
  unexecuted code as a result.
