---
title: "Spatial imports and standard workflow methods"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center",
                      dpi = 144, out.width = "92%")
```

This page helps you choose the right input for a spatial method. First run the
complete bundled Visium example: validate IDs and coordinates, calculate
limited-assay QC, normalize without changing counts, inspect a real expression
map, and save/reload the result. Then adapt the import section to an existing
Visium, Visium HD, or Xenium vendor output directory.

The [main spatial workflow](spatial-main-workflow.html) provides the analysis
path. The [framework bridges](spatial-framework-bridges.html) page verifies
conversion to SpatialExperiment and checks the optional Giotto backend.

## Run a complete input check

`visium_human_pancreas_sub` comes from sample GSM8058244 (PanIN-LG2), study
[GSE254829](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE254829).
It contains all 1,986 packaged tissue spots and 5,000 genes shared with the
bundled pancreas reference. The package has already selected those genes;
this page does not reduce the object further. `coda_label` is supplied
microanatomical annotation, not a predicted single-cell type.

```{r platform-input}
library(scop)
data(visium_human_pancreas_sub)
visium <- visium_human_pancreas_sub
assay_name <- "Spatial"
image_name <- "slice1"
counts <- GetAssayData5(visium, assay = assay_name, layer = "counts")
raw_coordinates <- SpatialCoordinates(visium, image = image_name, space = "raw")

stopifnot(
  identical(dim(counts), c(5000L, 1986L)),
  !anyDuplicated(colnames(counts)),
  identical(rownames(visium@meta.data), colnames(visium)),
  setequal(colnames(counts), raw_coordinates$data$cell_id),
  all(is.finite(as.matrix(raw_coordinates$data[, c("x", "y")]))),
  all(is.finite(counts@x)), all(counts@x >= 0),
  all(counts@x == floor(counts@x))
)
input_receipt <- data.frame(
  source = "GSE254829 / GSM8058244 / complete bundled object",
  observations = ncol(counts), features = nrow(counts),
  observation_unit = "spot", assay = assay_name, layer = "counts",
  image = image_name, coordinate_space = "raw", coordinate_units = "pixel",
  total_selected_gene_counts = sum(counts)
)
input_receipt
raw_coordinates$source
SeuratObject::Layers(visium[[assay_name]])
```

The count total describes only the packaged genes. The ID checks prevent an
expression column from being paired with another spot's coordinates. The
fixed dimension check is specific to this example; replace it with your own
expected dimensions. Keep the general ID, finite-coordinate, and raw-count
checks when changing datasets.

### Label QC problems without discarding observations

A gene-filtered assay cannot reproduce whole-transcriptome library complexity
or mitochondrial percentages. Here, UMI and gene thresholds are an executable
example on the retained genes, not validated tissue-QC cutoffs. For real QC,
import the complete raw expression assay and choose thresholds after examining
its distributions and tissue image.

```{r platform-qc}
qc_parameters <- list(qc_metrics = c("umi", "gene"), UMI_threshold = 500,
                      gene_threshold = 200, return_filtered = FALSE)
visium_qc <- do.call(RunSpotQC, c(
  list(object = visium, assay = assay_name, verbose = FALSE), qc_parameters
))
qc_summary <- as.data.frame(table(visium_qc$SpotQC, useNA = "ifany"))
colnames(qc_summary) <- c("status", "spots")
qc_summary
summary(visium_qc@meta.data[, c("nCount_Spatial", "nFeature_Spatial")])
stopifnot(
  identical(colnames(visium_qc), colnames(visium)),
  isTRUE(all.equal(GetAssayData5(visium_qc, assay = assay_name, layer = "counts"), counts)),
  sum(qc_summary$spots) == ncol(visium)
)
```

`Pass` means both requested thresholds passed; `Fail` means at least one
failed. An `NA` result is unresolved, not a pass. All observations remain in
this example because `return_filtered = FALSE`, so you can inspect failures
before deciding whether to exclude them.

```{r platform-qc-map, fig.height=4.5}
SpatialSpotPlot(visium_qc, group.by = "SpotQC", image = image_name)
```

### Normalize and inspect a measured feature

```{r platform-normalize}
visium_normalized <- Seurat::NormalizeData(
  visium_qc, assay = assay_name, normalization.method = "LogNormalize",
  scale.factor = 10000, verbose = FALSE
)
stopifnot(isTRUE(all.equal(
  GetAssayData5(visium_normalized, assay = assay_name, layer = "counts"), counts
)))
SeuratObject::Layers(visium_normalized[[assay_name]])
feature <- "GCG"
stopifnot(feature %in% rownames(counts))
feature_values <- GetAssayData5(visium_normalized, assay = assay_name, layer = "data")[feature, ]
data.frame(feature = feature, spots_with_counts = sum(counts[feature, ] > 0),
           minimum_log_expression = min(feature_values),
           maximum_log_expression = max(feature_values))
```

```{r platform-expression-map, fig.height=4.5}
SpatialSpotPlot(visium_normalized, features = feature, assay = assay_name,
                layer = "data", image = image_name, palette = "YlGnBu",
                legend.title = "Log-normalized GCG")
```

This map shows measured GCG expression after library-size scaling and `log1p`
transformation over the retained genes. It is not a deconvolution result or a
cell-type call. Recompute normalization from a complete assay for production
analysis; values normalized over different gene universes need not be comparable.
The `counts` equality check confirms that the raw source was preserved.

### Save, reload, and verify

```{r platform-save-reload}
output_dir <- Sys.getenv("SCOP_PLATFORM_OUTPUT_DIR", file.path(tempdir(), "scop-platform"))
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
result_file <- file.path(output_dir, "panin-input-qc-normalized.rds")
saveRDS(list(object = visium_normalized, input = input_receipt,
             qc_parameters = qc_parameters, qc_summary = qc_summary,
             coordinates = raw_coordinates, session = utils::sessionInfo()), result_file)
reloaded <- readRDS(result_file)
reload_checks <- c(
  observations = identical(colnames(reloaded$object), colnames(visium)),
  counts = isTRUE(all.equal(
    GetAssayData5(reloaded$object, assay = assay_name, layer = "counts"), counts)),
  qc_labels = identical(reloaded$object$SpotQC, visium_normalized$SpotQC),
  coordinates = identical(
    SpatialCoordinates(reloaded$object, image = image_name, space = "raw"), raw_coordinates)
)
data.frame(check = names(reload_checks), passed = unname(reload_checks))
stopifnot(all(reload_checks))
```

The build writes to a temporary directory. Set `SCOP_PLATFORM_OUTPUT_DIR` to a
new persistent project directory before running this code on your data. Use
`reloaded$object` in either plotting call above; plotting does not rerun QC or
normalization. Save the input receipt alongside figures so their units and
selected assay remain recoverable.

## Platform contracts at a glance

| Input | Observation | Typical assay/image | Coordinate meaning | Natural SCOP entry |
|---|---|---|---|---|
| Visium | spot | `Spatial` / `slice1` | full-resolution image pixels; display scale is separate | `SpatialSpotPlot()`, spot QC, Moran/Geary, domain clustering, spot deconvolution |
| Visium HD | bin | `Spatial.008um` or `Spatial.016um` / matching image | full-resolution pixels for the selected bin resolution | `SpatialSpotPlot()`, resolution-aware BayesSpace, spatial-variable features |
| Xenium | segmented cell | vendor assay / `fov` | vendor segmentation coordinates in microns | `SpatialCellPlot()`, cell-level context or communication |

The exact assay and image names are read from the loaded object; do not copy a
name from a different resolution or sample. Check the selected context before
calling a method:

```{r platform-context}
SpatialCoordinates(visium, image = image_name, space = "raw")$source
SeuratObject::DefaultAssay(visium)
SeuratObject::Images(visium)
```

SCOP keeps spot IDs/cell IDs aligned with expression columns. If several images
cover one sample, analysis and plotting require an explicit selection or an
explicit named sample-to-image map. A display `lowres`/`hires` scale never
changes raw distances and does not turn pixels into microns.

## Import existing vendor output

Set one or more of `SCOP_VISIUM_DIR`, `SCOP_VISIUM_HD_DIR`, and
`SCOP_XENIUM_DIR` to existing, unpacked vendor output directories before
running this section. Visium needs the expression matrix plus the `spatial/`
outputs; HD needs the selected binned outputs and matching images; Xenium
needs expression and existing cell segmentation. Do not point at an archive.
No path is configured by default, so the table explicitly reports `not_run`.

```{r import-vendor-data}
vendor_paths <- c(
  visium = Sys.getenv("SCOP_VISIUM_DIR"),
  visium_hd = Sys.getenv("SCOP_VISIUM_HD_DIR"),
  xenium = Sys.getenv("SCOP_XENIUM_DIR")
)
vendor_objects <- list()
import_receipts <- lapply(names(vendor_paths), function(technology) {
  path <- vendor_paths[[technology]]
  status <- "not_run"
  reason <- "No vendor directory configured"
  observations <- NA_integer_
  if (nzchar(path)) {
    result <- tryCatch({
      if (!dir.exists(path)) stop("Configured vendor directory does not exist")
      ReadSpatialData(data.dir = path, technology = technology,
                      bin.size = c(8, 16), sample_id = paste0(technology, "_sample1"))
    }, error = function(e) e)
    if (inherits(result, "error")) {
      status <- "failed"
      reason <- conditionMessage(result)
    } else {
      vendor_objects[[technology]] <<- result
      status <- "completed"
      reason <- "Vendor import and SCOP context validation completed"
      observations <- ncol(result)
    }
  }
  data.frame(technology = technology, status = status, observations = observations,
             reason = reason)
})
import_receipts <- do.call(rbind, import_receipts)
import_receipts
```

Only `completed` means an import ran. For example, after a successful Visium
import, use `visium <- vendor_objects$visium`, inspect its images/assays, and
repeat the input checks with its actual dimensions. If your matrix filename
is nonstandard, pass Seurat's `filename =` argument through `ReadSpatialData()`;
the GSE254829 source archive uses `filename = "filtered_feature_bc_matrix"`.
For HD, inspect every image and its matching assay separately: one resolution
is not interchangeable with another.

`ReadSpatialData()` delegates file parsing to Seurat and records import
provenance under `object@misc$scop_spatial_input`. It validates that each
returned image, assay, observation set, resolution, and coordinate-unit label
agree. It does not normalize counts, segment cells, or infer cell types.

Public data sources for testing the input formats are available from the
[10x Genomics Visium datasets](https://www.10xgenomics.com/datasets), the
[10x Visium HD human-pancreas example](https://www.10xgenomics.com/datasets/visium-hd-cytassist-11mm-human-pancreas),
the [10x Visium HD developer example data](https://www.10xgenomics.com/support/software/space-ranger/latest/resources/visium-hd-example-data),
and [10x Xenium Explorer demo datasets](https://www.10xgenomics.com/support/software/xenium-explorer/latest/resources/xenium-explorer-demos).
Download and unpack them outside the package; do not hard-code a machine path
in a reproducible article.

## Optional method recipes: check availability first

The complete input/QC/normalization path above is evaluated. The following
backend-specific recipes are not executed by this page: several require
separate installations, another sample, a matched reference, or a costly model
fit. This table measures installation status and explicitly distinguishes it
from execution. A package being installed does not prove that its model ran.

```{r platform-backend-status}
backend_packages <- c("SpaNorm", "Banksy", "BayesSpace", "PRECAST", "spacexr")
backend_status <- data.frame(
  package = backend_packages,
  installed = vapply(backend_packages, requireNamespace, logical(1), quietly = TRUE),
  version = vapply(backend_packages, function(p) {
    if (requireNamespace(p, quietly = TRUE)) as.character(packageVersion(p)) else NA_character_
  }, character(1)),
  execution = "not_run: optional recipe on this page"
)
backend_status
```

The main workflow contains executable spatial-feature and reference-based
examples. Use its execution receipts to establish which methods actually ran.

### Visium and Visium HD: keep counts and normalized data separate

Raw counts are the input for count-dependent QC and methods that model sampling
depth. Normalized/log-transformed data are appropriate for expression-level
plots and methods whose contract asks for a normalized layer. `SpaNorm` is a
separate normalization stage; it does not replace the original count source.

```{r spanorm-template, eval=FALSE}
stopifnot(requireNamespace("SpaNorm", quietly = TRUE),
          requireNamespace("Banksy", quietly = TRUE))
normalized <- RunStandardWorkflow(
  visium,
  workflow = "spatial",
  assay = "Spatial",
  image = "slice1",
  normalization_method = "SpaNorm",
  spot_qc_params = qc_parameters,
  spanorm_params = list(new_assay = "SpaNormAnalysis"),
  do_spatial_cluster = TRUE,
  spatial_cluster_method = "BANKSY",
  do_deconvolution = FALSE
)

normalized@tools$run_standard_spatial_workflow$stages

# A separate count-dependent method; q is a required choice, not an inferred truth.
stopifnot(requireNamespace("BayesSpace", quietly = TRUE))
bayesspace <- RunBayesSpace(
  normalized,
  assay = "Spatial",  # this assay must retain the original counts
  image = "slice1",
  q = 4,
  store_sce = FALSE
)
table(bayesspace$BayesSpace_cluster)
```

The workflow attaches matching original counts when needed for QC and uses the
separate normalized assay for downstream expression preprocessing. It rejects
an explicit `do_normalization = FALSE` conflict with `SpaNorm`. Do not relabel
normalized values as counts. `RunBayesSpace()` has no `layer` argument: its
selected assay must contain the intended counts. Inspect the domain map and
compare justified choices of `q`; four domains in this recipe is a parameter,
not a biological conclusion.

### Standard methods by platform

For a Visium spot or HD bin, a compact method selection is:

```{r standard-methods, eval=FALSE}
baseline <- RunStandardWorkflow(
  visium,
  workflow = "spatial",
  assay = "Spatial",
  image = "slice1",
  do_spot_qc = TRUE,
  spot_qc_params = qc_parameters,
  do_spatial_variable_features = TRUE,
  spatial_variable_features_params = list(method = "moran", nfeatures = nrow(visium[["Spatial"]])),
  do_spatial_cluster = FALSE,
  do_deconvolution = FALSE
)

baseline@tools$run_standard_spatial_workflow$stages
stopifnot(requireNamespace("Banksy", quietly = TRUE))
banksy <- RunStandardWorkflow(
  visium,
  workflow = "spatial",
  assay = "Spatial",
  image = "slice1",
  do_spot_qc = FALSE,
  do_spatial_variable_features = FALSE,
  do_spatial_cluster = TRUE,
  spatial_cluster_method = "BANKSY",
  spatial_cluster_params = list(lambda = 0.2, resolution = 0.6),
  do_deconvolution = FALSE
)

banksy@tools$run_standard_spatial_workflow$stages
table(banksy$BANKSY_cluster)
SpatialSpotPlot(banksy, group.by = "BANKSY_cluster", image = "slice1")
```

`BayesSpace` is the spot/bin array branch; `spatial_q` or BayesSpace's `q`
controls the number of domains. `BANKSY` uses its own resolution parameters.
`SmoothClust` accepts a requested `n_clusters` and its smoothing parameters.
Segmented cell observations are not silently sent through spot deconvolution or
BayesSpace; use cell-level methods and plots instead.

### Multi-sample PRECAST integration

Integration needs a sample label and a covering image per sample. With a merged
object, the image map names are the merged object's image keys. With list input,
the map names refer to the original list element names before Seurat renames
duplicate image keys.

Use two independently acquired samples saved as Seurat RDS files. Set the two
environment variables below to those files; do not split or duplicate this
single bundled section to manufacture an integration result. Inspect each
object and adjust the assay/image map before running.

```{r precast-template, eval=FALSE}
stopifnot(requireNamespace("PRECAST", quietly = TRUE))
sample_files <- c(S1 = Sys.getenv("SCOP_SPATIAL_S1_RDS"),
                  S2 = Sys.getenv("SCOP_SPATIAL_S2_RDS"))
stopifnot(all(nzchar(sample_files)), all(file.exists(sample_files)),
          length(unique(normalizePath(sample_files))) == 2L)
visium_S1 <- readRDS(sample_files[["S1"]])
visium_S2 <- readRDS(sample_files[["S2"]])
stopifnot(inherits(visium_S1, "Seurat"), inherits(visium_S2, "Seurat"))
lapply(list(S1 = visium_S1, S2 = visium_S2), SeuratObject::Images)

integrated <- RunSpatialIntegration(
  object = list(S1 = visium_S1, S2 = visium_S2),
  method = "PRECAST",
  assay = "Spatial",
  layer = "counts",
  image = c(S1 = "slice1", S2 = "slice1"),
  store_object = FALSE
)

integrated@tools$SpatialIntegration$summary
SeuratObject::Images(integrated)
SpatialIntegrationPlot(integrated, plot_type = "spatial")
saveRDS(integrated, "precast-integration.rds")
```

The plotting call resolves the images in the merged object. Inspect
`SeuratObject::Images(integrated)` before supplying an explicit plotting map:
duplicate input names such as `slice1` may have been renamed during merging.
Do not reuse the input-list map as a merged-object map.

The PRECAST package is an execution-time optional dependency. The template is
not a successful run unless PRECAST is installed and the input objects exist.
`store_object = FALSE` omits only the complete native PRECAST object; domains,
embedding, aligned coordinates, parameters, source information, and standard
plots remain stored. Keep the default `TRUE` when a later native PRECAST
analysis needs the full object.

PRECAST receives a fresh backend object built from the selected `assay` and
`layer`, which must contain finite, non-negative integer counts; it normalizes
these counts internally. The default adjacency uses six nearest neighbors in
the selected coordinate space (`adj_params = list(type = "fixed_number", number = 6)`).
SCOP constructs this graph with its shared exact KNN implementation and adapts
it to PRECAST's column-neighbor matrix convention. Each sample must retain more
spots than the requested neighbor count. PRECAST's fixed-distance Visium/ST mode expects array indices and is
rejected for SCOP's analysis coordinates. Empty or invalid graphs stop before
training. Effective backend arguments and per-sample nonzero edge counts are
saved under the method bundle's `parameters`.

For SpatialCellChat, verify the physical calibration before interpreting
distance-dependent communication. The current wrapper has a legacy ordinary
Visium fallback of `65 / spot_diameter_fullres` and `tol = 32.5`; these are
implementation assumptions, not the current platform specification.
[10x specifies a 55-micron diameter for ordinary Visium spots](https://kb.10xgenomics.com/s/article/360035487812-What-is-the-size-of-the-spots-on-the-Visium-Gene-Expression-Slide).
For confirmed 55-micron spots with full-resolution pixel coordinates, supply
`ratio = 55 / spot_diameter_fullres` explicitly and choose a justified `tol`
in microns. Do not apply a full-resolution ratio to resized-image coordinates.
Visium HD needs its own calibrated `ratio` and tolerance for the chosen bin
or cell size; an ordinary spot-diameter conversion does not apply. Imported HD
provenance and assay names such as `Spatial.008um` inform automatic technology
detection; an explicit `technology = "visium_hd"` is needed when that evidence
is absent.

## Deconvolution and cell2location storage choices

Spot deconvolution is reference-dependent and estimates composition per spot;
it does not turn a spot into a segmented cell. `RunRCTD()`, `RunSPOTlight()`,
and `RunCell2location()` preserve their own result semantics. Use
`SpatialDeconvolutionPlot()` or `Cell2locationPlot()` after checking the stored
result. Cell2location q05 abundance and normalized proportions are distinct;
the abundance view uses its native range rather than a forced 0--1 scale.
Cell2location plotting matrices retain the original spot IDs. Spots excluded
from modeling, such as zero-count observations, have `NA` values rather than
zero abundance. The saved `cells` field identifies modeled spots, and
`input_summary$dropped_spots` records filtered observations. Historical compact
results are expanded only when those records explain all missing rows; missing
predictions or unknown IDs are not silently accepted.
An explicit `lower_cutoff`, `upper_cutoff`, or quantile control overrides the
default matrix-wide display range. With a one-sided override, each panel keeps
the remaining bound computed by `SpatialSpotPlot()`; specify both cutoffs when
you want identical explicit limits across panels. `combine` changes only the
return form, including for the dominant point map.

For a complete reference-backed RCTD run, use the
[main workflow](spatial-main-workflow.html). Choose a reference with compatible
gene identifiers and biologically relevant labels, and retain its count assay.
The stored model result, not an invented component matrix, is the input to
`SpatialDeconvolutionPlot()`. For a saved cell2location result, verify that
`input_summary`, modeled `cells`, and the selected abundance/proportion matrix
are present before calling `Cell2locationPlot()`.

## Common errors and how to recover

| Symptom | Check and recovery |
|---|---|
| Vendor directory does not exist or a matrix is missing | Unpack the archive; point at the vendor output directory; check the actual matrix filename and loader arguments |
| More than one image or HD resolution | List images and matching assays; select the intended pair explicitly, one resolution at a time |
| Image IDs are absent from the assay | Compare IDs and sample/resolution provenance; choose the matching assay instead of silently taking an intersection |
| QC labels are unexpectedly good or bad | Check the gene universe and count layer; inspect distributions and tissue; do not apply whole-transcriptome thresholds to a selected-gene assay |
| SpaNorm conflicts with normalization settings | Remove `do_normalization = FALSE` when requesting SpaNorm and preserve the original count assay |
| A backend is installed but no result is present | Read the actual workflow stage status/reason; package availability is not a completed analysis |
| PRECAST adjacency is empty or invalid | Verify raw coordinates, sample/image coverage, and more than six spots per sample for the default six-neighbor graph |
| Spots have missing composition values | Check the producer's modeled cells and dropped-spot record; `NA` means unmodeled/missing, not zero abundance |

## Validation and analysis receipts

Acceptance runs use public slide data outside the package: developer fixtures
check format and computation rather than biological performance on production
slides, and public slide data are not redistributed with scop.

For a new dataset, preserve the following in the analysis receipt:

- platform and observation unit (`spot`, `bin`, or `cell`);
- assay and layer, selected image, coordinate columns, coordinate space, and
  physical units when known;
- raw-count versus normalized-expression input decisions;
- sample-to-image map for every sample;
- stage status and backend versions for methods that actually ran.

```{r platform-session}
data.frame(
  package = c("scop", "Seurat", "SeuratObject", "Matrix"),
  version = vapply(c("scop", "Seurat", "SeuratObject", "Matrix"),
                   function(p) as.character(utils::packageVersion(p)), character(1))
)
```

## Build provenance

This receipt belongs to the actual rendered source and retained observations.
Save it beside the exported object when adapting the workflow.

```{r build-provenance}
source_revision <- Sys.getenv("GITHUB_SHA")
if (!nzchar(source_revision) && nzchar(Sys.which("git"))) {
  source_revision <- system2("git", c("rev-parse", "HEAD"), stdout = TRUE,
                             stderr = FALSE)[1]
}
build_receipt <- data.frame(
  article = "spatial-platform-workflows", built_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
  source_revision = source_revision,
  source_has_local_changes = if (nzchar(Sys.which("git")))
    length(system2("git", c("status", "--porcelain"), stdout = TRUE,
                   stderr = FALSE)) > 0 else NA,
  article_md5 = unname(tools::md5sum(knitr::current_input(dir = TRUE))),
  R = as.character(getRversion()), scop = as.character(utils::packageVersion("scop")),
  retained_observations = ncol(visium_normalized), genes = nrow(visium_normalized)
)
build_receipt
utils::write.csv(build_receipt, file.path(output_dir, "build-receipt.csv"), row.names = FALSE)
writeLines(capture.output(sessionInfo()), file.path(output_dir, "session-info.txt"))
```
