---
title: "Spatial framework bridges"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

scop exchanges native objects with the Giotto and SPATA2 frameworks through
dedicated bridge functions. Each bridge processes exactly one image at a time
and never wraps the target framework's analysis or plotting API.

## Choose the interface

| Goal | Interface | Returned object | Plotting |
|---|---|---|---|
| Move one Seurat image into Giotto | `srt_to_giotto()` | Native Giotto object | Giotto's native plotting API |
| Move a native Giotto object into Seurat | `giotto_to_srt()` | New Seurat object | scop or Seurat plotting |
| Move between Seurat and SPATA2 | `srt_to_spata2()`, `spata2_to_srt()` | Native SPATA2 or Seurat object | Framework-native plotting |

## Giotto version boundary

Current Giotto Suite releases split analysis functions and object/conversion
helpers across the `Giotto` and `GiottoClass` namespaces. scop resolves these
optional functions at runtime and does not add Giotto to Imports, Depends, or
Suggests.

The forward bridge selects `seuratToGiottoV5()` for Seurat 5 and
`seuratToGiottoV4()` for Seurat 4. The reverse bridge selects
`giottoToSeuratV5()` for Seurat 5 and `giottoToSeuratV4()` for Seurat 4.
These converters are expected from a current GiottoClass installation. If an
older Giotto installation does not provide them, update Giotto before using
the bridges.

## GitHub documentation build contract

Giotto, GiottoClass, and SPATA2 remain runtime-optional backends. The examples
in this article are deliberately shown but not evaluated during the GitHub
pkgdown build, so publishing the documentation does not select or install a
particular framework generation. Reference pages are still parsed and indexed,
and dependency checks run when a user calls a bridge.

On pull requests, the GitHub workflow validates the complete reference index.
After a merge to `main`, changes that include source, reference, or pkgdown
configuration files trigger a full site build; an article-only change can use
the narrower article build. The scheduled full build remains the compatibility
check against the then-current dependency ecosystem.

Ordinary package and pkgdown checks remain independent of a user's Giotto
installation. A Giotto compatibility failure must be classified as a scop
regression, an upstream Giotto API regression, or a runner failure.

## Single-image conversion

The bridges process exactly one image at a time:

- a single image is selected automatically;
- a multi-image object requires an explicit `image`;
- an unknown image name is rejected;
- slices are never merged implicitly.

```{r}
gobject <- srt_to_giotto(spatial, image = "slice1")
spatial_roundtrip <- giotto_to_srt(gobject)
```

## SPATA2 bridge

SPATA2 receives the selected image explicitly through its `img_name`
interface. The bridge does not run trajectories, spatial screening, or SPATA2
plotting.

SPATA2 3.1.4 exports `asSeurat()` as a development stub rather than a complete
converter. scop still calls the official function first; when it does not
return a Seurat object, the compatibility fallback transfers counts, metadata,
and raw coordinates into a new Seurat object. It does not attempt to reproduce
SPATA2 analysis slots or framework plots.

```{r}
spata_object <- srt_to_spata2(spatial, image = "slice1")
spatial_roundtrip <- spata2_to_srt(spata_object)
```

Use SPATA2's native API after conversion. scop only owns image validation,
single-image isolation, dependency checks, and object exchange.
