---
title: "Spatial domain benchmark workflow"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.align = "center",
  dpi = 144,
  out.width = "92%"
)
```

This article compares spatial domain clustering methods with
[RunSpatialBenchmark()]. Each method runs in an isolated R process; results
include ARI, NMI, purity, runtime, and peak memory against a gold-standard
label column. Visualize the result with [SpatialBenchmarkPlot()].

Supported producers include `BayesSpace`, `BANKSY`, and `SmoothClust`. This
workflow is separate from scRNA batch integration benchmarking
([RunIntegrationBenchmark()]).

## Load data and define a gold standard

This workflow uses a small simulated spatial object with three regions and
region-specific expression signals. It demonstrates the benchmark interface;
replace the simulated labels with tissue annotations for real analyses.

```{r load-data}
library(scop)

set.seed(42)
n_spots <- 36L
n_genes <- 60L
grid <- expand.grid(x = 1:6, y = 1:6)
spatial_domain <- factor(ifelse(grid$x <= 2, "left", ifelse(grid$x <= 4, "middle", "right")))
counts <- matrix(rpois(n_genes * n_spots, 2),
  nrow = n_genes,
  dimnames = list(paste0("Gene", seq_len(n_genes)), paste0("spot", seq_len(n_spots)))
)
signal <- split(seq_len(n_genes), rep(c("left", "middle", "right"), each = 20))
for (label in names(signal)) {
  counts[signal[[label]], spatial_domain == label] <-
    counts[signal[[label]], spatial_domain == label] + rpois(sum(spatial_domain == label) * 20, 5)
}
spatial <- SeuratObject::CreateSeuratObject(counts)
spatial$x <- grid$x
spatial$y <- grid$y
spatial$gold_domain <- spatial_domain
table(spatial$gold_domain)
```

## Run the benchmark

Methods share the same spot set. Tune per-method arguments through
`method_params` when backends need different PCA counts or layers.

```{r run-benchmark}
bench <- RunSpatialBenchmark(
  spatial,
  gold_standard = "gold_domain",
  method_params = list(
    BayesSpace = list(n.PCs = 5, n.HVGs = 20, coord.cols = c("x", "y")),
    BANKSY = list(layer = "counts", coord.cols = c("x", "y")),
    SmoothClust = list(layer = "counts", min_spots = 1, coord.cols = c("x", "y"))
  ),
  verbose = FALSE
)

bench$summary
```

## Inspect the result object

`RunSpatialBenchmark()` returns a `spatial_benchmark_result` list:

- `$summary`: quality and resource metrics per method;
- `$predictions`: aligned domain labels per spot;
- `$metrics`: long metric table for custom plots.

```{r inspect-result}
str(bench$summary, max.level = 1)
head(bench$predictions)
```

## Plot benchmark views

`SpatialBenchmarkPlot()` defaults to a publication-oriented overview that
combines clustering agreement with runtime and peak-memory efficiency.

```{r plot-overview, fig.height=6}
SpatialBenchmarkPlot(data = bench, plot_type = "overview")
```

Other views:

```r
SpatialBenchmarkPlot(data = bench, plot_type = "quality")
SpatialBenchmarkPlot(data = bench, plot_type = "efficiency")
SpatialBenchmarkPlot(data = bench, plot_type = "heatmap")
```

## What to report

A spatial benchmark report should include:

- the gold-standard column and number of domains;
- methods compared and any `method_params` overrides;
- `$summary` with ARI, NMI, purity, runtime, and peak memory;
- the overview or quality plot;
- failed or unavailable methods (status column) when backends are missing.

Agreement metrics are higher when better; runtime and peak memory are lower
when better.
