---
title: "Spatial framework bridges"
---

This page describes two separate bridges from the [main spatial workflow](spatial-main-workflow.html):
Seurat↔SpatialExperiment and Seurat↔Giotto. A bridge moves a data container;
it does not claim that the target framework's analysis or plotting backend ran.
The [platform page](spatial-platform-workflows.html) defines assay, image, and
platform-specific input choices.

## What each bridge preserves

| Field | Seurat ↔ SpatialExperiment | Seurat ↔ Giotto |
|---|---|---|
| Expression | One selected Seurat `assay`/`layer` becomes the `scop_input` assay; it is restored as the requested Seurat assay | The selected image subset is passed to the official GiottoClass converter; native Giotto layers follow that converter's rules |
| Observation IDs and order | SpatialExperiment column names and coordinate row names must match exactly; reverse conversion reorders to assay-column order | The official converter owns target IDs; convert one selected image and check the returned object before analysis |
| Metadata | `include_meta = TRUE` copies Seurat metadata to `colData`; reverse conversion copies `colData` to Seurat metadata | Metadata retention follows `GiottoClass`; the bridge does not invent or merge sample labels |
| Coordinates | Raw coordinates are the default; an explicit display export records its transform and reverse conversion inverts it | One selected image's spatial coordinates are sent through the official converter; no extra coordinate system is invented by SCOP |
| Image/raster information | Raster image slots are not represented by the lightweight SpatialExperiment conversion | One image is selected and subset; multiple images/slices are never silently merged |
| Analysis and plots | No SpatialExperiment analysis is run by the bridge | No Giotto analysis or plot is run by the bridge |

The common invariant is identity: expression columns, metadata rows, and
coordinate rows must refer to the same spots/cells in the same intended order.
The bridge keeps provenance where SCOP can invert it, but it cannot preserve
target-framework fields that have no counterpart.

## Seurat and SpatialExperiment round-trip

`srt_to_spe()` exports one selected assay layer and coordinates. It defaults to
raw acquisition coordinates, which is the safe choice for downstream distances.
`coordinate_space = "legacy_display"` is an explicit compatibility export; the
source and affine transform are stored in
`metadata(spe)$scop_spatial_coordinates`.

```{r spe-roundtrip, eval=FALSE}
library(scop)
data(visium_human_pancreas_sub)

spatial <- Seurat::NormalizeData(
  visium_human_pancreas_sub,
  assay = "Spatial",
  verbose = FALSE
)
spe <- srt_to_spe(
  spatial,
  assay = "Spatial",
  layer = "counts",
  image = "slice1",
  coordinate_space = "raw"
)
spatial_roundtrip <- spe_to_srt(spe, assay = "Spatial", layer = "scop_input")

stopifnot(identical(colnames(spatial_roundtrip), colnames(spe)))
head(spatial_roundtrip@meta.data)
SpatialSpotPlot(spatial_roundtrip, image = NULL, features = rownames(spatial_roundtrip)[1])
```

The `stopifnot()` check is intentionally about IDs, not byte-for-byte equality
of every Seurat slot. The conversion is lightweight: it does not promise to
round-trip Seurat images, reductions, graphs, tools, or all assays. External
SpatialExperiment objects without SCOP coordinate provenance are interpreted
as raw coordinates in their supplied units; verify their units before using a
distance-sensitive method.

## Seurat and Giotto round-trip

The Giotto bridge is a one-image adapter. A single-image object can be resolved
automatically; a multi-image object requires `image =`. SCOP selects the
official `GiottoClass` v4/v5 converter for the installed Seurat major version,
reuses an installed optional package, and does not initialize Giotto's optional
Python environment.

```{r giotto-roundtrip, eval=FALSE}
library(scop)
data(visium_human_pancreas_sub)

spatial <- Seurat::NormalizeData(
  visium_human_pancreas_sub,
  assay = "Spatial",
  verbose = FALSE
)
giotto <- srt_to_giotto(spatial, image = "slice1")
spatial_from_giotto <- giotto_to_srt(giotto)

SeuratObject::Images(spatial_from_giotto)
SpatialSpotPlot(spatial_from_giotto, features = rownames(spatial_from_giotto)[1])
```

This block runs only when a compatible GiottoClass installation is available;
`eval=FALSE` is deliberate so a documentation build does not install or
download an external backend. A live round-trip must compare the expression
features, observation IDs, metadata columns, and coordinate values that the
installed converter claims to preserve. It must not be presented as proof that
Giotto's own analysis or plotting pipeline ran.

## Choosing a bridge

Use SpatialExperiment when a Bioconductor container and explicit assay/column
metadata are the interchange target. Use Giotto when the native Giotto object
and its official converter are the target. For either route, select one image,
record the coordinate space and units, check IDs immediately after conversion,
and re-check the target object's own accessors before starting downstream
analysis.
