---
title: "Spatial transcriptomics main workflow"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.align = "center",
  dpi = 144,
  out.width = "92%"
)
```

This article shows the recommended first-pass SCOP workflow for a spatial
transcriptomics object. It uses the bundled Visium pancreas subset and keeps the
main path short: inspect the object, run the spatial workflow, check summaries,
plot spatial features, and branch into optional backends only when the question
requires them.

It is not a catalogue of every spatial backend. Use method pages for backend
specific parameters and dependency notes.

Read the workflow as an evidence ladder: first confirm coordinates and image
orientation, then compute QC and spatially variable features, then add domains,
deconvolution, or neighborhood models only when they answer the biological
question.

## Load a Spatial Object

Start with a Seurat object that has expression, metadata, and spatial
coordinates. The bundled `visium_human_pancreas_sub` object has a `Spatial`
assay, a `slice1` image, and metadata coordinates.

```{r load-spatial}
library(scop)

data(visium_human_pancreas_sub)
spatial <- visium_human_pancreas_sub

SeuratObject::DefaultAssay(spatial) <- "Spatial"
SeuratObject::Images(spatial)
head(spatial@meta.data[, c("x", "y", "coda_label")])
```

Before running methods, make one direct map. This checks whether coordinates,
image orientation, labels, and the default spatial theme are sensible.
If this first map is rotated, mirrored, or shifted, downstream spatial domains
and neighborhood results will be difficult to interpret even if the model runs.
SCOP reads full-resolution raw coordinates for analysis and applies the selected
Seurat scale factor exactly once for display. The default is the low-resolution
raster. If the image slot contains `tissue_hires_image.png`, pass
`image.scale = "hires"`; do not rewrite `@scale.factors`.
Seurat's standard `VisiumV2` loader stores source `imagerow`/`imagecol`
positionally as centroid `x`/`y`; SCOP restores horizontal image-column and
vertical image-row order. Generic FOV centroid tables with explicit `x`/`y`
retain their order. Legacy Visium pixel coordinates are normalized only when
explicit image row/column names are available, before the selected display
transform is applied.

```{r inspect-spatial}
SpatialSpotPlot(
  spatial,
  group.by = "coda_label"
)

SpatialSpotPlot(
  spatial,
  features = rownames(spatial)[1:2],
  assay = "Spatial",
  layer = "counts"
)

# For an object loaded with a hires raster:
# SpatialSpotPlot(spatial, group.by = "coda_label", image.scale = "hires")
```

## Run the Basic Spatial Workflow

`RunStandardWorkflow(workflow = "spatial")` is the shortest entry point for a
first-pass spatial analysis. Keep optional domain clustering and deconvolution
off at first. This makes the first run fast and gives you QC and spatial
variable features before choosing heavier backends.

```{r standard-spatial}
spatial <- RunStandardWorkflow(
  spatial,
  workflow = "spatial",
  assay = "Spatial",
  image = "slice1",
  coord.cols = c("x", "y"),
  do_spot_qc = TRUE,
  do_spatial_variable_features = TRUE,
  spatial_variable_features_params = list(
    method = "moran",
    nfeatures = 50
  ),
  do_spatial_cluster = FALSE,
  do_deconvolution = FALSE
)
```

After the run, look at the stable result locations before making more plots.
Most spatial wrappers write method parameters and summaries under `srt@tools`.
Spatial variable features are genes whose expression varies with spatial
position according to the selected method. They are candidates for spatial
patterning, not automatically domain markers.

```{r inspect-results}
names(spatial@tools)
spatial@tools$SpatialVariableFeatures$summary
head(spatial@tools$SpatialVariableFeatures$summary$top_features)
```

## Plot the First Results

The default spatial maps hide axes and crop to the observed spots. Use
`show_axes = TRUE` only when debugging coordinates.
`SpotQC` summarizes spot-level quality-control status. Spatial feature surfaces
should be interpreted together with raw spot maps, because smoothing can make
sparse signal look broader than it is.

```{r plot-results}
SpatialSpotPlot(
  spatial,
  group.by = "SpotQC"
)

SpatialVariableFeaturePlot(
  spatial,
  plot_type = "summary"
)

SpatialVariableFeaturePlot(
  spatial,
  plot_type = "surface",
  nfeatures = 4
)
```

## Add Spatial Domains When Needed

Run a domain method only after the baseline object looks correct. BayesSpace is
a common Visium choice. BANKSY and SmoothClust are useful alternatives when
their assumptions match the data and optional dependencies are available.
Domain labels are unsupervised spatial clusters. They are useful for segmenting
tissue regions, but they need marker genes, histology, or reference labels
before being named biologically.

```r
spatial_bayes <- RunStandardWorkflow(
  spatial,
  workflow = "spatial",
  assay = "Spatial",
  image = "slice1",
  coord.cols = c("x", "y"),
  do_spot_qc = FALSE,
  do_spatial_variable_features = FALSE,
  do_spatial_cluster = TRUE,
  spatial_cluster_method = "BayesSpace",
  spatial_q = 3
)

SpatialSpotPlot(
  spatial_bayes,
  group.by = "BayesSpace_cluster"
)
```

For other domain backends, keep the output contract in mind: cluster labels
should land in metadata and method details should stay under `srt@tools`.

```r
spatial <- RunBANKSY(
  spatial,
  assay = "Spatial",
  layer = "counts",
  coord.cols = c("x", "y"),
  cluster_colname = "BANKSY_cluster"
)
SpatialSpotPlot(spatial, group.by = "BANKSY_cluster")
```

## Add Cell Composition Only With a Reference

For spot-level data, deconvolution depends more on reference quality than on the
wrapper choice. Use a matched single-cell reference, then inspect dominant cell
types and maximum proportions before interpreting spatial biology.

```r
spatial <- RunSpatialDWLS(
  spatial,
  reference = reference,
  reference_label = "celltype",
  assay = "Spatial",
  reference_assay = "RNA",
  coord.cols = c("x", "y"),
  prefix = "SpatialDWLS"
)

spatial@tools$SpatialDWLS$summary

SpatialSpotPlot(
  spatial,
  group.by = "SpatialDWLS_dominant_type"
)

SpatialSpotPlot(
  spatial,
  group.by = "SpatialDWLS_dominant_type",
  plot_type = "pie"
)
```

`RCTD`, `SPOTlight`, `CARD`, and `STdeconvolve` follow the same practical rule:
first check the result summary, then visualize dominant labels, maximum
proportions, and selected proportions.
Deconvolution estimates cell-type composition per spot. It is constrained by
the reference, so missing reference cell types can be misassigned to the closest
available label.

## Add Neighborhood or Context Models Last

Neighborhood methods answer a different question from domain clustering. Use
them after labels are stable, either from metadata, clustering, or
deconvolution.

```{r spatial-neighborhood}
spatial <- RunSpatialNeighborhood(
  spatial,
  group.by = "coda_label",
  coord.cols = c("x", "y"),
  k = 6
)

spatial@tools$SpatialNeighborhood$summary

SpatialNeighborhoodPlot(spatial, plot_type = "heatmap")
SpatialNeighborhoodPlot(spatial, plot_type = "stat", top_n = 12)
```

With `method = NULL`, the package computes observed KNN or radius summaries when
`split.by` is absent and preserves the historical spicyR route when `split.by`
is supplied. For new differential neighborhood analyses, request spicyR
explicitly and provide the condition column; SCOP errors before execution if
that design is incomplete.

```r
spatial <- RunSpatialNeighborhood(
  spatial,
  group.by = "coda_label",
  method = "spicyR",
  split.by = "condition",
  sample.by = "sample"
)
```

```r
SpatialNeighborhoodPlot(spatial, plot_type = "network", top_n = 12)
```

Use `RunStatialKontextual()` when you have explicit `from`, `to`, and `parent`
cell populations. Use `RunMistyR()` when the question is feature-level local or
broader spatial context rather than pairwise label enrichment.
Both are independent producers rather than `RunSpatialNeighborhood()` method
choices. HoodscanR remains unimplemented.
Neighborhood enrichment is label-level spatial association. Context models are
feature-level spatial association. Keep those interpretations separate in
reports.

Spatial results follow the same structure as single-cell analyses in SCOP:
each producer stores its output as a plain key in `srt@tools`, for example
`spatial@tools$MistyR` or `spatial@tools$SpatialNeighborhood`, with
method-specific fields at the top level plus `parameters` and `summary`.
Read results directly from `srt@tools[[tool_name]]` and pass them to the
dedicated plot functions. Custom `tool_name` values are honored exactly as
written.

```r
coords <- SpatialCoordinates(spatial, image = "slice1", space = "raw")
network_result <- spatial@tools$SpatialNetwork
network_graph <- GetSpatialGraph(
  res = network_result,
  format = "sparse",
  value = "weight"
)
names(spatial@tools)
```

Distance-sensitive methods default to full-resolution raw acquisition
coordinates. `coordinate_space = "legacy_display"` remains an explicit
compatibility option only. Plot functions map results to display coordinates by
cell or spot ID and `image.scale` never changes raw distances or neighborhoods.
Objects with multiple spatial images must always select one explicitly with
`image`; analysis, plotting, and framework conversion never silently use the
first image. Coordinate-dependent results created before contract v2 are not
silently migrated; rerun their producer. Metadata passed directly to
`SpatialSpotPlot()` cannot be traced to its producer, so users upgrading an old
analysis should rerun that analysis before plotting its metadata.

```r
spatial <- RunMistyR(
  spatial,
  assay = "Spatial",
  layer = "data",
  features = head(spatial@tools$SpatialVariableFeatures$summary$top_features, 20),
  coord.cols = c("x", "y"),
  views = "para"
)
spatial@tools$MistyR$summary
MistyRPlot(spatial, type = "improvements")
```

For label-level contextual scores, use
`StatialKontextualPlot(spatial)` after `RunStatialKontextual()`. These dedicated
plots summarize stored backend output and never rerun the analysis.

## Add Spatially Constrained Communication Deliberately

`RunCellChat()` and `RunSpatialCellChat()` answer different questions. The
first estimates expression-supported communication without a physical spatial
constraint. The second uses micron-scale distances and stores its group-level
table under the distinct method name `"SpatialCellChat"`, so the two results
can coexist in `spatial@tools$CCC` without overwriting one another.

Choose the analysis level from the observation represented by each column:

- `"cell"` for segmented cells such as Xenium or CosMx;
- `"spot"` for spot/domain communication, which must not be described as
  direct cell-type communication;
- `"composition"` for Visium spots with a spot-by-cell-type proportion matrix.

Distance parameters are always interpreted in microns. Micron coordinates use
`ratio = 1`. For Visium full-image pixel coordinates, `RunSpatialCellChat()`
can derive the conversion ratio as `65 / spot_diameter_fullres` when the
selected image contains exactly one trusted full-resolution spot diameter;
otherwise provide `ratio` explicitly. Never substitute a low-resolution display
scale. This article uses one slice; multi-sample objects are run independently
and require an explicit `sample.by` and named image mapping.

```{r spatial-cellchat, eval=FALSE}
spatial <- RunSpatialCellChat(
  spatial,
  group.by = "coda_label",
  image = "slice1",
  technology = "visium",
  analysis.level = "spot",
  coordinate.unit = "pixel",
  tol = 32.5,
  species = "Homo_sapiens",
  database = "protein",
  store.object = "minimal"
)

CCCNetworkPlot(spatial, method = "SpatialCellChat", plot_type = "circle")
CCCNetworkPlot(
  spatial,
  method = "SpatialCellChat",
  condition = "default",
  sample = "slice1",
  plot_type = "spatial",
  signaling = "SPP1",
  top_n = 50
)
CCCHeatmap(spatial, method = "SpatialCellChat", plot_type = "bubble")
SpatialCellChatPlot(spatial, plot_type = "incoming")
result <- spatial@tools$SpatialCellChat
```

`CCCNetworkPlot(plot_type = "spatial")` is the common stored-result renderer
for `SpatialCellChat`, `SpaTalk`, and `COMMOT`. It overlays all stored cells or
spots, draws group-centroid nodes with abundance-scaled sizes, and draws
sender-coloured directed edges whose widths represent the filtered communication
score. For `analysis.level = "composition"`, the default spot layer is a
proportion pie (`composition_display = "pie"`); use `composition_display =
"dominant"` only when a single representative type is acceptable. This plot
does not rerun an optional backend. A result created before the current spatial
plot payload was stored must be rerun rather than reconstructed from mutable
metadata.

`store.object = "minimal"` stores normalized interaction, pathway, network,
coordinate, and diagnostic results but not the large native S4 object or a
materialized cell-by-cell-by-LR edge table. Use `"full"` only when native
SpatialCellChat downstream analysis is required, then retrieve it with
`GetCCCObject(method = "SpatialCellChat")`.

Visium spots form a hexagonal acquisition grid, but six-neighbor topology is
not a replacement for the Euclidean diffusion model. Contact-dependent
signaling is therefore rejected in spot mode. P-values from spatial
permutation and non-spatial CellChat analyses also have different null models
and should not be compared as if they were interchangeable.

SpatialDM addresses a different question: which ligand-receptor pairs are
spatially associated, and which spots contain local interaction evidence. It
does not infer sender-receiver cell-type edges, so its Moran's R must not be
reported as communication probability or transport mass. The official
workflow is global selection followed by local spot selection. Use normalized
or log-transformed expression for `layer`, raw counts for `counts.layer`, and
set `l` or `eff_dist` in the raw coordinate unit of the selected image.

```{r spatialdm, eval=FALSE}
spatial <- RunSpatialDM(
  spatial,
  species = "human",
  image = "slice1",
  layer = "data",
  counts.layer = "counts",
  l = 1.2,
  cutoff = 0.2,
  global.threshold = 0.1,
  local.threshold = 0.1
)

SpatialDMPlot(spatial, plot_type = "weights", spot = "spot_001")
SpatialDMPlot(spatial, plot_type = "global", highlight = "SPP1_CD44")
SpatialDMPlot(spatial, plot_type = "local", pair = "SPP1_CD44")
GetSpatialDMResult(spatial, type = "global")
```

`SpatialDMPlot()` follows the information content of the official melanoma
tutorial while returning SCOP `ggplot`/`patchwork` objects and using SCOP
themes and palettes. The `weights` view should be checked before interpreting
global or local results because `l`, `cutoff`, and coordinate scale determine
the effective interaction range.

## What to Report

A compact first-pass spatial report should include:

- the source object, assay, image, and coordinate columns;
- spot or cell counts after filtering;
- top spatial variable features and the method used;
- the domain method, cluster count, and domain sizes if clustering was run;
- deconvolution dominant labels and maximum proportions if composition was run;
- neighborhood or context summaries only after labels are stable.

Keep the main workflow short. Add backend-specific sections only when they
answer a concrete biological question for the current dataset.
