---
title: "Trajectory and dynamic feature workflow"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.align = "center",
  dpi = 144,
  out.width = "92%"
)
```

This article shows how to organize trajectory, RNA-velocity, and dynamic-feature
analyses in `scop`. These methods answer related but distinct questions:
connectivity between cell states, pseudotime ordering, directional flow, and
features that vary along inferred lineages.

Use trajectory methods only after the basic embedding and annotation are stable.
Most trajectory errors are interpretation errors caused by unstable labels,
missing root states, or incompatible input assays.

The methods below should be read as complementary evidence layers, not as
interchangeable trajectory labels. PAGA describes group-level connectivity,
Slingshot and Monocle3 assign pseudotime, RNA velocity estimates local
direction, and dynamic-feature models test genes against the chosen ordering.

## Prepare the Object

The bundled pancreas object contains spliced and unspliced assays, making it a
convenient example for both ordinary trajectory inference and RNA velocity.

```{r prepare-object}
library(scop)

data(pancreas_sub)
pancreas_sub <- standard_scop(pancreas_sub, verbose = FALSE)

SeuratObject::Assays(pancreas_sub)
table(pancreas_sub$SubCellType)
```

## RNA Velocity

RNA velocity requires spliced and unspliced matrices. Generate those matrices
outside `scop` with tools such as velocyto, bustools, or alevin-fry, then store
them as assays in the Seurat object.

```{r scvelo}
pancreas_sub <- RunSCVELO(
  pancreas_sub,
  group.by = "SubCellType",
  linear_reduction = "PCA",
  nonlinear_reduction = "UMAP"
)

VelocityPlot(
  pancreas_sub,
  reduction = "UMAP",
  group.by = "SubCellType"
)

VelocityPlot(
  pancreas_sub,
  reduction = "UMAP",
  plot_type = "stream"
)
```

Interpret velocity as a local directional estimate. It should agree with marker
genes, developmental knowledge, and trajectory structure before being used as
evidence for cell-state transitions.
Velocity arrows or streams do not define cell types by themselves. They are most
useful when their direction agrees with known early and terminal marker genes.

## PAGA Connectivity

PAGA summarizes coarse connectivity between annotated groups. It is useful for
checking whether a graph-based topology matches known or expected transitions.

```{r paga}
pancreas_sub <- RunPAGA(
  pancreas_sub,
  group.by = "SubCellType",
  linear_reduction = "PCA",
  nonlinear_reduction = "UMAP"
)

PAGAPlot(
  pancreas_sub,
  reduction = "UMAP",
  label = TRUE,
  label_insitu = TRUE,
  label_repel = TRUE,
  edge_size = c(0.5, 1),
  edge_color = "black"
)
```

Use PAGA to reason about group-level graph structure. Use pseudotime methods
when the question requires an ordering of individual cells.

## Slingshot Lineages

Slingshot is a practical choice when the embedding has clear branching and root
or terminal groups can be specified.

```{r slingshot}
pancreas_sub <- RunSlingshot(
  pancreas_sub,
  group.by = "SubCellType",
  reduction = "UMAP",
  start = "Ductal",
  end = c("Alpha", "Beta")
)

CellDimPlot(
  pancreas_sub,
  group.by = "SubCellType",
  lineages = paste0("Lineage", 1:2),
  reduction = "UMAP"
)

FeatureDimPlot(
  pancreas_sub,
  features = paste0("Lineage", 1:2),
  reduction = "UMAP",
  theme_use = "theme_blank"
)
```

The lineage pseudotime columns can be used directly by dynamic-feature methods
and plotting functions.
`Lineage1` and `Lineage2` are numeric pseudotime columns written to metadata by
the Slingshot wrapper. They are method-specific coordinates, not absolute time.

## Monocle3 Trajectories

Monocle3 learns a trajectory graph and stores trajectory layers that can be
added to ordinary `scop` plots.

```{r monocle3}
pancreas_sub <- RunMonocle3(
  pancreas_sub,
  group.by = "SubCellType",
  use_partition = FALSE,
  root_cells = colnames(pancreas_sub)[pancreas_sub$SubCellType == "Ductal"][1]
)

trajectory <- pancreas_sub@tools$Monocle3$trajectory
milestones <- pancreas_sub@tools$Monocle3$milestones

CellDimPlot(
  pancreas_sub,
  group.by = "SubCellType",
  reduction = "UMAP",
  label = TRUE
) +
  trajectory +
  milestones

FeatureDimPlot(
  pancreas_sub,
  features = "Monocle3_Pseudotime",
  reduction = "UMAP"
) +
  trajectory
```

Check partitions, clusters, roots, and pseudotime direction before interpreting
branch-specific changes.
`Monocle3_Pseudotime` is another numeric ordering derived from the Monocle3
principal graph. It can agree with Slingshot but does not have to use the same
scale or branch assignment.

## Dynamic Features

After a lineage has been selected, use dynamic-feature analysis to find genes
that vary along one or more lineages.

```{r dynamic-features}
pancreas_sub <- RunDynamicFeatures(
  pancreas_sub,
  lineages = c("Lineage1", "Lineage2"),
  n_candidates = 200
)

pancreas_sub <- AnnotateFeatures(
  pancreas_sub,
  species = "Mus_musculus",
  db = c("TF", "CSPA")
)
```

Use `DynamicHeatmap()` for module-like changes and `DynamicPlot()` for a small
set of known features.
In this heatmap, `GO_BP` means the Gene Ontology Biological Process database
used to annotate dynamic gene modules. The term annotation summarizes module
themes; it is not a claim that every gene in the module belongs to one pathway.

```{r dynamic-plots}
ht <- DynamicHeatmap(
  pancreas_sub,
  lineages = c("Lineage1", "Lineage2"),
  use_fitted = TRUE,
  n_split = 6,
  reverse_ht = "Lineage1",
  species = "Mus_musculus",
  db = "GO_BP",
  anno_terms = TRUE,
  anno_keys = FALSE,
  anno_features = FALSE,
  heatmap_palette = "viridis",
  cell_annotation = "SubCellType",
  feature_annotation = c("TF", "CSPA")
)
print(ht$plot)

DynamicPlot(
  pancreas_sub,
  lineages = c("Lineage1", "Lineage2"),
  group.by = "SubCellType",
  features = c("Plk1", "Hes1", "Neurod2", "Ghrl", "Gcg", "Ins2")
)
```

## What to Report

A compact trajectory report should include:

- the object, embedding, and annotation used to define cell states;
- whether spliced and unspliced assays were available for velocity;
- the selected method and root or terminal groups;
- pseudotime or lineage columns written to metadata;
- graph, velocity, and pseudotime plots with the same annotation labels;
- dynamic genes and any functional annotations used to interpret them.

Do not mix method outputs as if they measure the same quantity. Report
connectivity, directionality, pseudotime, and dynamic expression as separate
evidence layers.
