---
title: "Cell-cell communication workflow"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.align = "center",
  dpi = 144,
  out.width = "92%"
)
```

This article shows a practical first-pass workflow for cell-cell communication
analysis in `scop`: prepare a cell-type annotation, run ligand-receptor
inference, inspect sender-receiver summaries, then focus on specific ligands,
receptors, pathways, or downstream targets only after the global result looks
stable.

The executable examples below use the bundled `pancreas_sub` data and run the
R-side CellChat and LIANA wrappers directly. CellphoneDB is intentionally not
used here because it requires a Python environment and a CellPhoneDB database.

`scop` keeps the workflow unified in three layers:

| Layer | Entry point | Purpose |
| --- | --- | --- |
| Result tables | `RunCCC()` output | Store standardized `long_table`, `pair_table`, and method status in `srt@tools$CCC`. |
| Heatmap layer | `CCCHeatmap()` | Matrix, dot, tile, sample, pathway, and differential heatmap views. |
| Statistical layer | `CCCStatPlot()` | Ranked interactions, distributions, pathway summaries, and comparison statistics. |
| Network layer | `CCCNetworkPlot()` | Sender-receiver networks, pathway networks, embedding networks, and focused ligand-receptor flow. |

The plots in this article summarize inferred ligand-receptor evidence. A high
score means the selected method found strong expression-based support for a
sender ligand, receiver receptor, and sender-receiver context. It should not be
read as a direct measurement of secreted protein amount, physical distance, or
causal signaling.

## Prepare the Object

Start with a normalized Seurat object and a metadata column that represents the
cell identities used as sender and receiver groups. For the bundled pancreas
example, `CellType` is the coarse annotation and `SubCellType` is the finer
annotation.

```{r load-object}
library(scop)
library(ggplot2)

data(pancreas_sub)
pancreas_sub <- standard_scop(pancreas_sub, verbose = FALSE)

table(pancreas_sub$CellType)
table(pancreas_sub$SubCellType)
```

Use the same visual settings across the article so that heatmaps, statistical
summaries, and network plots can be compared without changing palettes or text
scale.

```{r visual-style}
ccc_cell_pal <- c(
  "Ductal" = "#4E79A7",
  "Endocrine" = "#E15759",
  "Ngn3-high-EP" = "#59A14F",
  "Ngn3-low-EP" = "#76B7B2",
  "Pre-endocrine" = "#F28E2B"
)

ccc_value_pal <- c("#2C5C85", "#8FBBD9", "#F7F7F7", "#F2B38F", "#B9473D")

ccc_theme_args <- list(base_size = 11)
```

Before communication analysis, check whether groups have enough cells or whether
some labels should be merged. CCC methods compare group-level expression, so
unstable labels usually become noisy sender-receiver edges. The bundled
`pancreas_sub` object is used as-is in this article.

```{r prepare-groups}
cell_counts <- sort(table(pancreas_sub$CellType), decreasing = TRUE)
cell_counts

pancreas_ccc <- pancreas_sub
pancreas_ccc$CellType <- droplevels(factor(pancreas_ccc$CellType))

table(pancreas_ccc$CellType)
```

## Run a CCC Screen

`RunCCC()` is the shortest path when the goal is to compare the same object with
several ligand-receptor scoring methods. In this article, it runs CellChat and
LIANA, then rebuilds the unified `srt@tools$CCC` tables for common downstream
plots.

```{r run-ccc}
pancreas_ccc <- RunCCC(
  pancreas_ccc,
  group.by = "CellType",
  methods = c("CellChat", "LIANA"),
  method_params = list(
    CellChat = list(species = "Mus_musculus", min.cells = 5),
    LIANA = list(
      method = c("natmi", "connectome"),
      resource = "Consensus",
      min_cells = 5
    )
  ),
  skip_failed = FALSE,
  thresh = 0.05,
  verbose = FALSE
)
```

Check the run status and standardized result tables before plotting. Use
`pancreas_ccc@tools$CCC$long_table` for ligand-receptor questions and
`pancreas_ccc@tools$CCC$pair_table` for sender-receiver summaries.
The long table is the most detailed layer: one row per method-specific
ligand-receptor sender-receiver event. The pair table aggregates those rows to
cell-type pairs, so it is useful for overview plots but less useful for naming
specific biology.

```{r inspect-run-status}
pancreas_ccc@tools$RunCCC$status
pancreas_ccc@tools$RunCCC$completed_methods

ccc_long <- pancreas_ccc@tools$CCC$long_table
ccc_pair <- pancreas_ccc@tools$CCC$pair_table

head(ccc_long)
head(ccc_pair)
```

```{r summarize-ccc-tables}
long_df <- pancreas_ccc@tools$CCC$long_table
pair_df <- pancreas_ccc@tools$CCC$pair_table

pair_df[order(pair_df$sum, decreasing = TRUE), ][1:6, ]

top_lr <- long_df[order(long_df$score, decreasing = TRUE), ]
top_lr[1:6, c("method", "sender", "receiver", "ligand", "receptor", "score", "pvalue")]
```

## Global Communication Maps

Start with the heatmap layer. A sender-receiver heatmap shows which cell-type
pairs have many or strong inferred interactions.
In aggregation views, `sum` summarizes the selected interaction score across
retained ligand-receptor events for a sender-receiver pair. Use it to rank cell
pairs, then return to the long table to inspect the individual ligands and
receptors behind the aggregate signal.

```{r ccc-heatmap, fig.width=6.8, fig.height=5.4}
CCCHeatmap(
  pancreas_ccc,
  method = "CCC",
  plot_type = "heatmap",
  display_by = "aggregation",
  value = "sum",
  top_n = 30,
  add_text = TRUE,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  value_palette = "RdBu",
  value_palcolor = ccc_value_pal,
  cell_palcolor = ccc_cell_pal,
  font.size = 10,
  theme_args = ccc_theme_args
)
```

A focused matrix keeps a single sender or receiver readable while preserving
interaction names. In static documentation, use small multiples or filters
instead of drawing every significant pair at once.

```{r ccc-ductal-dotplot, fig.width=7.6, fig.height=5.2}
CCCHeatmap(
  pancreas_ccc,
  method = "CCC",
  plot_type = "dot",
  display_by = "interaction",
  sender.use = "Ductal",
  top_n = 12,
  value_palette = "RdBu",
  value_palcolor = ccc_value_pal,
  cell_palcolor = ccc_cell_pal,
  font.size = 9,
  legend.position = "bottom",
  legend.direction = "horizontal",
  theme_args = ccc_theme_args
)
```

Compare the same interaction-level view from the receiver side. This is useful
when a sender-dominant dot plot hides receiver-specific specificity.

```{r ccc-endocrine-dotplot, fig.width=7.6, fig.height=5.2}
CCCHeatmap(
  pancreas_ccc,
  method = "CCC",
  plot_type = "dot",
  display_by = "interaction",
  receiver.use = "Endocrine",
  top_n = 12,
  value_palette = "RdBu",
  value_palcolor = ccc_value_pal,
  cell_palcolor = ccc_cell_pal,
  font.size = 9,
  legend.position = "bottom",
  legend.direction = "horizontal",
  theme_args = ccc_theme_args
)
```

For pathway-level questions, use pathway bubbles before drawing a network.
Keep this panel compact in the article because the pathway-specific interaction
labels are long.
Here a pathway such as `COLLAGEN` is a ligand-receptor signaling family from
the CCC method/resource, not a GO or KEGG enrichment term.

```{r ccc-pathway-bubble, fig.width=6.4, fig.height=3.8, out.width="78%"}
CCCHeatmap(
  pancreas_ccc,
  method = "CCC",
  plot_type = "pathway_bubble",
  display_by = "interaction",
  signaling = "COLLAGEN",
  top_n = 6,
  value_palette = "RdBu",
  value_palcolor = ccc_value_pal,
  cell_palcolor = ccc_cell_pal,
  font.size = 8,
  legend.position = "bottom",
  legend.direction = "horizontal",
  theme_args = ccc_theme_args
)
```

## Statistical Views

Use the statistical layer to decide whether a signal is dominated by a few
ligand-receptor pairs or spread across many pairs.
`pvalue` is method-specific and should be interpreted with the method and
resource used to generate it. When several CCC methods are combined, use the
standardized table for ranking and visualization, but keep method provenance in
the report.

```{r ccc-summary-plots, warning=FALSE, fig.width=8, fig.height=5.2}
CCCStatPlot(
  pancreas_ccc,
  method = "CCC",
  plot_type = "bar",
  display_by = "interaction",
  top_n = 14,
  cell_palcolor = ccc_cell_pal,
  link_palcolor = ccc_cell_pal,
  font.size = 10,
  legend.position = "bottom",
  legend.direction = "horizontal",
  theme_args = ccc_theme_args
)
```

A flow plot is a compact companion to the bar plot: it shows which senders,
ligand-receptor interactions, and receivers explain the top-ranked signal.
For the article, use a small interaction-level flow with direct labels instead
of a dense alluvial panel.

```{r ccc-sankey, warning=FALSE, fig.width=8, fig.height=4.8}
CCCNetworkPlot(
  pancreas_ccc,
  method = "CCC",
  plot_type = "sigmoid",
  display_by = "interaction",
  top_n = 5,
  edge_alpha = 0.5,
  edge_size = c(0.25, 1),
  node_size = 3.8,
  cell_palcolor = ccc_cell_pal,
  link_palcolor = ccc_cell_pal,
  label = TRUE,
  label.size = 3.2,
  label.fg = "black",
  label.bg = "white",
  label.bg.r = 0.12,
  font.size = 9,
  legend.position = "none",
  theme_args = ccc_theme_args
)
```

Distribution plots show whether each sender has consistent receiver-side scores
or only a few high-scoring outliers. The legend glyphs are simplified so the
legend reads as colored categories rather than mini violins.

```{r ccc-violin, warning=FALSE, fig.width=8.4, fig.height=5.4}
p_violin <- CCCStatPlot(
  pancreas_ccc,
  method = "CCC",
  plot_type = "violin",
  facet_by = "sender",
  top_n = 12,
  cell_palcolor = ccc_cell_pal,
  font.size = 9,
  legend.position = "right",
  theme_args = ccc_theme_args
)

p_violin +
  ggplot2::guides(
    color = ggplot2::guide_legend(
      override.aes = list(shape = 16, size = 4, linetype = 0, alpha = 1)
    ),
    fill = ggplot2::guide_legend(
      override.aes = list(shape = 22, size = 4, linetype = 0, alpha = 1)
    )
  ) +
  ggplot2::theme(
    legend.key.height = grid::unit(0.35, "cm"),
    legend.key.width = grid::unit(0.35, "cm")
  )
```

For a global sender-receiver distribution, the box plot is often more compact
than faceted violins.

```{r ccc-boxplot, warning=FALSE, fig.width=7.6, fig.height=4.8}
CCCStatPlot(
  pancreas_ccc,
  method = "CCC",
  plot_type = "box",
  top_n = 12,
  cell_palcolor = ccc_cell_pal,
  font.size = 9,
  legend.position = "bottom",
  legend.direction = "horizontal",
  theme_args = ccc_theme_args
)
```

## Network Views

Use the network layer after the global heatmap identifies the cell groups of
interest. Aggregate views summarize sender-receiver communication; interaction
views show ligand-receptor flow. The compact arrow view is usually easier to
read than a dense circle network in an article.
Network edge width encodes the selected summary score, and arrows encode sender
to receiver direction. The direction is inferred from ligand and receptor group
assignment, not from time-course causality.

```{r ccc-network-arrow, fig.width=7.5, fig.height=5.8}
CCCNetworkPlot(
  pancreas_ccc,
  method = "CCC",
  plot_type = "arrow",
  display_by = "aggregation",
  value = "sum",
  top_n = 12,
  directed = TRUE,
  edge_line = "curved",
  edge_curvature = 0.18,
  edge_alpha = 0.55,
  edge_size = c(0.25, 1.4),
  node_size = 8,
  cell_palcolor = ccc_cell_pal,
  link_palcolor = ccc_cell_pal,
  font.size = 10,
  theme_args = ccc_theme_args
)
```

The sigmoid flow view uses the same sender-receiver table but lays it out as
columns. It is usually cleaner than a circle when labels are long.

```{r ccc-network-sigmoid, fig.width=8.2, fig.height=5.2}
CCCNetworkPlot(
  pancreas_ccc,
  method = "CCC",
  plot_type = "sigmoid",
  display_by = "aggregation",
  value = "sum",
  top_n = 10,
  edge_alpha = 0.55,
  edge_size = c(0.25, 1.2),
  node_size = 4,
  cell_palcolor = ccc_cell_pal,
  link_palcolor = ccc_cell_pal,
  font.size = 10,
  legend.position = "bottom",
  legend.direction = "horizontal",
  theme_args = ccc_theme_args
)
```

The embedding network overlays sender-receiver links on the cell embedding.
This helps distinguish strong communication between nearby cell states from
long-range links across transcriptionally distant groups.

```{r ccc-embedding-network, fig.width=7.5, fig.height=5.8}
CCCNetworkPlot(
  pancreas_ccc,
  method = "CCC",
  plot_type = "embedding_network",
  group.by = "CellType",
  reduction = "Standardumap",
  display_by = "aggregation",
  value = "sum",
  top_n = 12,
  edge_alpha = 0.45,
  edge_size = c(0.15, 0.9),
  node_size = 6,
  cell_palcolor = ccc_cell_pal,
  link_palcolor = ccc_cell_pal,
  font.size = 10,
  label = TRUE,
  label_repel = TRUE,
  label.size = 3,
  theme_args = ccc_theme_args
)
```

For pathway-aware results, focus on one pathway at a time. This avoids the
label overlap that appears when all significant sender-receiver links are drawn
in one circle.

```{r ccc-pathway-network, fig.width=7.5, fig.height=5.8, fig.keep='first'}
CCCNetworkPlot(
  pancreas_ccc,
  method = "CCC",
  plot_type = "pathway",
  signaling = "COLLAGEN",
  display_by = "aggregation",
  value = "sum",
  top_n = 8,
  node_size = 7,
  cell_palcolor = ccc_cell_pal,
  link_palcolor = ccc_cell_pal,
  font.size = 9,
  label.size = 3,
  theme_args = ccc_theme_args
)
```

For a focused ligand, the bipartite plot shows the flow from sender groups to
the ligand, then to receptors and receiver groups. Keep `top_n` small when the
same ligand has many receiver contexts.

```{r ccc-bipartite, fig.width=8, fig.height=5.2}
CCCNetworkPlot(
  pancreas_ccc,
  method = "CCC",
  plot_type = "bipartite",
  ligand = "Igf2",
  top_n = 6,
  node_size = 4,
  cell_palcolor = ccc_cell_pal,
  link_palcolor = ccc_cell_pal,
  font.size = 8,
  legend.position = "bottom",
  legend.direction = "horizontal",
  theme_args = ccc_theme_args
)
```

Use a single ligand-receptor pair when the biological question is already
specific. The pair name should match `interaction_name`, `pair_lr`, or
`interaction_label` in `pancreas_ccc@tools$CCC$long_table`.

```{r ccc-individual-lr, fig.width=7.5, fig.height=5.8, fig.keep='first'}
CCCNetworkPlot(
  pancreas_ccc,
  method = "CCC",
  plot_type = "individual_lr",
  signaling = "IGF",
  pairLR.use = "Igf2 - Igf1r",
  display_by = "interaction",
  top_n = 8,
  node_size = 7,
  cell_palcolor = ccc_cell_pal,
  link_palcolor = ccc_cell_pal,
  font.size = 9,
  label.size = 3,
  theme_args = ccc_theme_args
)
```

The dense circle or chord variants are still useful for interactive inspection,
but they are intentionally not drawn here because the full `pancreas_sub`
network is too dense for a static article figure:

```r
CCCNetworkPlot(
  pancreas_ccc,
  method = "CCC",
  plot_type = "circle",
  display_by = "aggregation",
  value = "sum",
  top_n = 8
)

CCCNetworkPlot(
  pancreas_ccc,
  method = "CCC",
  plot_type = "chord",
  display_by = "aggregation",
  top_n = 8
)
```

## Method-Specific R Workflows

Use method-specific wrappers when you need parameters or outputs that are not
part of the unified scheduler. CellChat is useful for pathway-level
communication and pathway role plots:

```r
pancreas_cellchat <- RunCellChat(
  pancreas_ccc,
  group.by = "CellType",
  species = "Mus_musculus",
  min.cells = 10
)
```

LIANA is useful when you want several ligand-receptor scoring methods and a
consensus-style table:

```r
pancreas_liana <- RunLIANA(
  pancreas_ccc,
  group.by = "CellType",
  method = c("natmi", "connectome", "logfc", "cellphonedb"),
  resource = "Consensus",
  min_cells = 5
)

liana_tbl <- ccc_to_liana(pancreas_liana, method = "LIANA")
```

For condition comparisons, store the condition in metadata and run CellChat per
condition. The comparison output can then be used by differential heatmaps,
differential networks, and pathway ranking plots.

```r
pancreas_cmp <- RunCellChat(
  pancreas_ccc,
  group.by = "CellType",
  group_column = "Condition",
  group_cmp = list(c("ConditionA", "ConditionB")),
  species = "Mus_musculus"
)
```

## Link Ligands to Downstream Targets

Use NicheNet when the question is no longer "which cell types communicate?" but
"which sender ligands may explain receiver target-gene changes?". This requires
a receiver, sender groups, and either a condition comparison or a custom target
gene set.

```r
pancreas_nn <- RunNichenetr(
  pancreas_ccc,
  group.by = "CellType",
  receiver = "Ductal",
  sender = "all",
  condition.by = "Condition",
  condition_oi = "ConditionA",
  condition_reference = "ConditionB",
  mode = "aggregate_cluster_de",
  species = "Mus_musculus"
)
```

## Export R Results

Export a LIANA-style table from the unified CCC result when you want a portable
result table for downstream workflows without rebuilding method-specific
objects by hand.

```{r export-results}
liana_res <- ccc_to_liana(
  pancreas_ccc,
  method = "CCC",
  score_col = "score",
  pvalue_col = "pvalue"
)

head(liana_res)
```

## What to Report

A compact CCC report should include:

- the object, assay, expression layer, and sender-receiver annotation;
- the method or methods used, including species and ligand-receptor resource;
- the significance threshold and expression fraction cutoff;
- the top sender-receiver pairs by interaction count or score;
- the top ligand-receptor pairs, separated from aggregate cell-pair summaries;
- whether the result is steady-state communication or a condition comparison;
- downstream target interpretation only when using a method designed for that
  question, such as NicheNet.

Keep the first pass broad and conservative. Use focused ligand, receptor,
pathway, and downstream target analyses only after the global sender-receiver
structure is stable.
