---
title: "Cell annotation workflow"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.align = "center",
  dpi = 144,
  out.width = "92%"
)
```

This article shows how to annotate query cells in `scop` using reference
single-cell data, bulk-like reference profiles, and correlation-based summaries.
The goal is to make annotation evidence explicit instead of treating transferred
labels as final truth.

Use this workflow after the query object has passed basic preprocessing and
quality control.

Annotation in this article means evidence-backed label assignment. Projection,
nearest-neighbor prediction, and correlation heatmaps answer related but
different questions, so the transferred label should not be treated as final
until it agrees with marker expression and expected biology.

## Prepare Query and Reference Objects

The query is the bundled mouse pancreas subset. The reference is the bundled
human pancreas integration subset, which has a `celltype` annotation and an
integrated embedding.

```{r prepare-query-reference}
library(scop)

data(pancreas_sub)
data(panc8_sub)

pancreas_sub <- standard_scop(pancreas_sub, verbose = FALSE)
panc8_sub <- integration_scop(
  srt_merge = panc8_sub,
  batch = "tech",
  integration_method = "Harmony"
)
```

When query and reference use different gene-name conventions, standardize names
before mapping. Keep the name conversion explicit so that annotation results can
be traced later.
The example converts human reference symbols into a mouse-like capitalization
style because the query object is mouse. In real analyses, use the species and
ortholog mapping appropriate for the experiment rather than relying on simple
capitalization.

```{r rename-features}
genenames <- make.unique(
  thisutils::capitalize(
    rownames(panc8_sub[["RNA"]]),
    force_tolower = TRUE
  )
)
names(genenames) <- rownames(panc8_sub)

panc8_rename <- RenameFeatures(
  panc8_sub,
  newnames = genenames,
  assays = "RNA"
)
```

## Project Query Cells Onto a Reference

Use projection when the reference has a trusted embedding and you want to see
where query cells land in that reference coordinate system.

```{r projection}
srt_query <- RunKNNMap(
  srt_query = pancreas_sub,
  srt_ref = panc8_rename,
  ref_umap = "HarmonyUMAP2D"
)

ProjectionPlot(
  srt_query = srt_query,
  srt_ref = panc8_rename,
  query_group = "SubCellType",
  ref_group = "celltype",
  ref_param = list(
    xlab = "UMAP_1",
    ylab = "UMAP_2"
  )
)
```

Projection is evidence for similarity, not a replacement for marker validation.
Use it together with marker plots and group-level summaries.
Cells landing near a reference group indicate transcriptional similarity in the
reference embedding. They do not prove identity when the reference lacks the
state, contains batch effects, or uses a broader label vocabulary than the
query.

## Predict Labels From a Bulk-Like Reference

Use `RunKNNPredict()` with a bulk-like reference when reference profiles are
available but not individual reference cells.

```{r bulk-reference}
data(ref_scMCA)

pancreas_sub <- RunKNNPredict(
  srt_query = pancreas_sub,
  bulk_ref = ref_scMCA,
  filter_lowfreq = 20
)

CellDimPlot(
  pancreas_sub,
  group.by = "KNNPredict_classification",
  reduction = "UMAP",
  label = TRUE
)
```

Bulk-like references are useful for broad labels. They are less reliable for
substates that depend on local single-cell neighborhoods.
The `KNNPredict_classification` column is therefore best read as a suggested
label source. Keep it separate from curated annotations until discordant groups
have been reviewed.

## Predict Labels From a Single-Cell Reference

Use a single-cell reference when the reference annotation and query biology are
close enough for nearest-neighbor transfer.

```{r single-cell-reference}
pancreas_sub <- RunKNNPredict(
  srt_query = pancreas_sub,
  srt_ref = panc8_rename,
  ref_group = "celltype",
  filter_lowfreq = 20
)

CellDimPlot(
  pancreas_sub,
  group.by = "KNNPredict_classification",
  reduction = "UMAP",
  label = TRUE
)
```

Check group-level similarity before accepting labels. A correlation heatmap can
show whether query labels map cleanly to expected reference labels or split
across several alternatives.
Rows in this heatmap are query groups and columns are reference groups. High
correlation to one reference label supports a clear annotation; similar
correlation to multiple labels flags ambiguity.

```{r correlation-heatmap}
ht <- CellCorHeatmap(
  srt_query = pancreas_sub,
  srt_ref = panc8_rename,
  query_group = "SubCellType",
  ref_group = "celltype",
  nlabel = 3,
  label_by = "row",
  show_row_names = TRUE,
  show_column_names = TRUE
)
print(ht$plot)
```

## What to Report

A compact annotation report should include:

- the query object, reference object, and gene-name conversion strategy;
- the reference label column and embedding used for projection;
- whether labels came from bulk-like profiles or single-cell neighbors;
- filtering thresholds such as `filter_lowfreq`;
- projection plots, predicted label plots, and marker-gene checks;
- ambiguous groups where several labels have similar support.

Treat annotation as an evidence layer. Keep transferred labels separate from
manual labels until marker genes and biological context support the assignment.
