---
title: "Differential expression and enrichment workflow"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.align = "center",
  dpi = 144,
  out.width = "92%"
)
```

This article shows how to organize differential expression, marker visualization,
and pathway enrichment in `scop`. The main distinction is between testing genes,
summarizing gene sets, and interpreting enriched biological terms.

Use this workflow after cell identities and major covariates are stable.
Differential expression is gene-level evidence. Enrichment is a summary of a
gene list or ranked statistic against a named database. Keep those two evidence
layers separate in the text of a report.

## Prepare the Object

```{r prepare-object}
library(scop)

data(pancreas_sub)
pancreas_sub <- standard_scop(pancreas_sub, verbose = FALSE)

table(pancreas_sub$CellType)
```

## Run Differential Expression

`RunDEtest()` writes differential-expression tables under `srt@tools`. The
default single-cell marker workflow is appropriate for exploratory marker
finding. Use sample-aware or pseudobulk methods when biological replicates are
part of the question.

```{r run-detest}
pancreas_sub <- RunDEtest(
  pancreas_sub,
  group.by = "CellType",
  fc.threshold = 1,
  only.pos = FALSE
)

names(pancreas_sub@tools)
head(pancreas_sub@tools$DEtest_CellType$AllMarkers_wilcox)
```

Keep the contrast, grouping column, test method, and threshold in the report.
Those choices define the meaning of each marker table.
For `FindAllMarkers`-style results, `group1` names the group being compared
against the remaining cells unless a specific contrast is stated.

## Plot Differential Results

Use volcano, Manhattan, and ring views for different levels of detail. Volcano
plots are useful for one group or contrast; Manhattan and ring plots are useful
when scanning many groups.

```{r detest-plots}
DEtestPlot(
  pancreas_sub,
  group.by = "CellType",
  plot_type = "volcano",
  label.size = 2
)

DEtestPlot(
  pancreas_sub,
  group.by = "CellType",
  plot_type = "manhattan",
  label.size = 2
)

DEtestPlot(
  pancreas_sub,
  group.by = "CellType",
  plot_type = "ring",
  label.size = 2
)
```

When enrichment results are available, volcano plots can annotate enriched
terms to connect marker genes with pathway-level interpretation.
The enrichment overlay circles genes that belong to selected enriched terms.
The legend title is the analysis layer (`Enrichment`), and the legend item is
the database, for example `GO_BP`.

```{r annotated-volcano, fig.width=10, fig.height=8}
pancreas_sub <- RunEnrichment(
  pancreas_sub,
  group.by = "CellType",
  db = "GO_BP",
  species = "Mus_musculus"
)

DEtestPlot(
  pancreas_sub,
  group.by = "CellType",
  plot_type = "volcano",
  threshold_method = "hyperbolic",
  hyperbola_c = 6,
  annotate_enrichment = TRUE,
  enrich_from = "Enrichment",
  enrich_db = "GO_BP",
  enrich_top_terms = 3,
  enrich_nlabel = 15,
  label.size = 2
)
```

## Summarize Markers in Heatmaps

Feature heatmaps are useful after markers have been filtered. Keep the filter
explicit and avoid showing very large marker sets without grouping or splitting.
`TF` marks transcription-factor annotations and `CSPA` marks cell-surface
protein annotations. They help prioritize marker genes for follow-up, but they
do not change the DE statistics.

```{r feature-heatmap}
DEGs <- pancreas_sub@tools$DEtest_CellType$AllMarkers_wilcox
DEGs <- DEGs[with(DEGs, avg_log2FC > 1 & p_val_adj < 0.05), ]

pancreas_sub <- AnnotateFeatures(
  pancreas_sub,
  species = "Mus_musculus",
  db = c("TF", "CSPA")
)

ht <- FeatureHeatmap(
  pancreas_sub,
  group.by = "CellType",
  features = DEGs$gene,
  feature_split = DEGs$group1,
  exp_legend_title = "Z-score",
  species = "Mus_musculus",
  db = c("GO_BP", "KEGG", "WikiPathway"),
  anno_terms = TRUE,
  feature_annotation = c("TF", "CSPA")
)
print(ht$plot)
```

## Run Over-Representation Analysis

`RunEnrichment()` tests whether filtered marker sets are enriched for terms from
the selected database.
`GO_BP` is Gene Ontology Biological Process, `KEGG` is a curated pathway
database, and `WikiPathway` contains community-curated pathway sets. These are
database labels; the specific enriched terms are the rows shown by
`EnrichmentPlot()`.

```{r ora}
pancreas_sub <- RunEnrichment(
  pancreas_sub,
  group.by = "CellType",
  db = "GO_BP",
  species = "Mus_musculus",
  DE_threshold = "avg_log2FC > log2(1.5) & p_val_adj < 0.05",
  cores = 5
)

EnrichmentPlot(
  pancreas_sub,
  group.by = "CellType",
  group_use = c("Ductal", "Endocrine"),
  plot_type = "bar"
)
```

Use different plot types for different questions: bar or lollipop plots for
ranked terms, word clouds for broad themes, and networks or enrichmaps for term
relationships.

```{r enrichment-plots}
EnrichmentPlot(
  pancreas_sub,
  group.by = "CellType",
  group_use = "Ngn3-low-EP",
  plot_type = "network"
)

EnrichmentPlot(
  pancreas_sub,
  group.by = "CellType",
  plot_type = "comparison",
  topTerm = 3
)
```

## Run GSEA

Use GSEA when ranked gene statistics are more appropriate than a hard marker
cutoff.
ORA depends on a marker cutoff, while GSEA uses a ranked gene list. Use GSEA
when many genes shift modestly and a binary marker threshold would discard
useful signal.

```{r gsea}
pancreas_sub <- RunGSEA(
  pancreas_sub,
  group.by = "CellType",
  db = "GO_BP",
  species = "Mus_musculus",
  DE_threshold = "p_val_adj < 0.05",
  cores = 5
)

GSEAPlot(
  pancreas_sub,
  group.by = "CellType",
  group_use = "Endocrine",
  id_use = "GO:0007186"
)

GSEAPlot(
  pancreas_sub,
  group.by = "CellType",
  group_use = "Endocrine",
  plot_type = "bar",
  direction = "both",
  topTerm = 10
)
```

## What to Report

A compact differential and enrichment report should include:

- the grouping column, contrast, and DE method;
- whether the test is single-cell marker detection, sample-level pseudobulk, or
  bulk-style differential testing;
- thresholds for log fold change, adjusted p value, and marker filtering;
- the gene-set database, species, and enrichment method;
- the exact `srt@tools` result names used for downstream plots;
- pathway interpretations separated from the gene-level DE evidence.

Use enrichment to summarize marker lists or ranked statistics. Do not treat an
enriched term as proof that every gene in the pathway changes coherently.
