---
title: "Spatial framework bridges and Giotto compatibility"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

scop supports Giotto in two deliberately separate ways. The older workflow is
kept for compatibility, while the newer bridge functions exchange native
framework objects without wrapping every Giotto or SPATA2 analysis step.

## Choose the interface before starting

| Goal | Interface | Returned object | Plotting |
|---|---|---|---|
| Keep an existing scop Giotto pipeline | `SeuratToScopGiotto()`, `RunGiottoWorkflow()`, `GiottoPlot()`, `AddGiottoToSeurat()` | scop's legacy `giotto2` workflow wrapper | `GiottoPlot()` |
| Move one Seurat image into Giotto | `srt_to_giotto()` | Native Giotto object | Giotto's native plotting API |
| Move a native Giotto object into Seurat | `giotto_to_srt()` | New Seurat object | scop or Seurat plotting |
| Move between Seurat and SPATA2 | `srt_to_spata2()`, `spata2_to_srt()` | Native SPATA2 or Seurat object | Framework-native plotting |

Do not mix the two Giotto styles in one pipeline. In particular,
`RunGiottoWorkflow()` expects the legacy scop wrapper, whereas
`srt_to_giotto()` intentionally returns the native Giotto object inside that
wrapper.

## Giotto version boundary

Current Giotto Suite releases split analysis functions and object/conversion
helpers across the `Giotto` and `GiottoClass` namespaces. scop resolves these
optional functions at runtime and does not add Giotto to Imports, Depends, or
Suggests.

The new reverse bridge selects `giottoToSeuratV5()` for Seurat 5 and
`giottoToSeuratV4()` for Seurat 4. These converters are expected from a current
GiottoClass installation. If an older Giotto installation does not provide
them, update Giotto before using `giotto_to_srt()`.

The legacy workflow remains available for older scripts and is not migrated to
the new bridge result container. This compatibility promise applies to its
public arguments, `giotto2` return structure, stored result names, and
`GiottoPlot()` behavior.

## Single-image conversion

The new bridges process exactly one image at a time:

- a single image is selected automatically;
- a multi-image object requires an explicit `image`;
- an unknown image name is rejected;
- slices are never merged implicitly.

```{r}
gobject <- srt_to_giotto(spatial, image = "slice1")
spatial_roundtrip <- giotto_to_srt(gobject)
```

For exact image selection, `srt_to_giotto()` uses scop's controlled converter
by default and returns only its native Giotto object. Advanced users may pass
`use_official = TRUE`, but the controlled default is more stable across Giotto
versions and multi-image Seurat objects.

## SPATA2 bridge

SPATA2 receives the selected image explicitly through its `img_name`
interface. The bridge does not run trajectories, spatial screening, or SPATA2
plotting.

SPATA2 3.1.4 exports `asSeurat()` as a development stub rather than a complete
converter. scop still calls the official function first; when it does not
return a Seurat object, the compatibility fallback transfers counts, metadata,
and raw coordinates into a new Seurat object. It does not attempt to reproduce
SPATA2 analysis slots or framework plots.

```{r}
spata_object <- srt_to_spata2(spatial, image = "slice1")
spatial_roundtrip <- spata2_to_srt(spata_object)
```

Use SPATA2's native API after conversion. scop only owns image validation,
single-image isolation, dependency checks, and object exchange.
