---
title: "Spatial transcriptomics main workflow"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.align = "center",
  dpi = 144,
  out.width = "92%"
)
```

This article shows the recommended first-pass SCOP workflow for a spatial
transcriptomics object. It uses the bundled Visium pancreas subset and keeps the
main path short: inspect the object, run the spatial workflow, check summaries,
plot spatial features, and branch into optional backends only when the question
requires them.

It is not a catalogue of every spatial backend. Use method pages for backend
specific parameters and dependency notes.

Read the workflow as an evidence ladder: first confirm coordinates and image
orientation, then compute QC and spatially variable features, then add domains,
deconvolution, or neighborhood models only when they answer the biological
question.

## Load a Spatial Object

Start with a Seurat object that has expression, metadata, and spatial
coordinates. The bundled `visium_human_pancreas_sub` object has a `Spatial`
assay, a `slice1` image, and metadata coordinates.

```{r load-spatial}
library(scop)

data(visium_human_pancreas_sub)
spatial <- visium_human_pancreas_sub

SeuratObject::DefaultAssay(spatial) <- "Spatial"
SeuratObject::Images(spatial)
head(spatial@meta.data[, c("x", "y", "coda_label")])
```

Before running methods, make one direct map. This checks whether coordinates,
image orientation, labels, and the default spatial theme are sensible.
If this first map is rotated, mirrored, or shifted, downstream spatial domains
and neighborhood results will be difficult to interpret even if the model runs.

```{r inspect-spatial}
SpatialSpotPlot(
  spatial,
  group.by = "coda_label"
)

SpatialSpotPlot(
  spatial,
  features = rownames(spatial)[1:2],
  assay = "Spatial",
  layer = "counts"
)
```

## Run the Basic Spatial Workflow

`standard_scop(workflow = "spatial")` is the shortest entry point for a
first-pass spatial analysis. Keep optional domain clustering and deconvolution
off at first. This makes the first run fast and gives you QC and spatial
variable features before choosing heavier backends.

```{r standard-spatial}
spatial <- standard_scop(
  spatial,
  workflow = "spatial",
  assay = "Spatial",
  image = "slice1",
  coord.cols = c("x", "y"),
  do_spot_qc = TRUE,
  do_spatial_variable_features = TRUE,
  spatial_variable_features_params = list(
    method = "moran",
    nfeatures = 50
  ),
  do_spatial_cluster = FALSE,
  do_deconvolution = FALSE
)
```

After the run, look at the stable result locations before making more plots.
Most spatial wrappers write method parameters and summaries under `srt@tools`.
Spatial variable features are genes whose expression varies with spatial
position according to the selected method. They are candidates for spatial
patterning, not automatically domain markers.

```{r inspect-results}
names(spatial@tools)
spatial@tools$SpatialVariableFeatures$summary
head(spatial@misc$SpatialVariableFeatures)
```

## Plot the First Results

The default spatial maps hide axes and crop to the observed spots. Use
`show_axes = TRUE` only when debugging coordinates.
`SpotQC` summarizes spot-level quality-control status. Spatial feature surfaces
should be interpreted together with raw spot maps, because smoothing can make
sparse signal look broader than it is.

```{r plot-results}
SpatialSpotPlot(
  spatial,
  group.by = "SpotQC"
)

SpatialVariableFeaturePlot(
  spatial,
  plot_type = "summary"
)

SpatialVariableFeaturePlot(
  spatial,
  plot_type = "surface",
  nfeatures = 4
)
```

## Add Spatial Domains When Needed

Run a domain method only after the baseline object looks correct. BayesSpace is
a common Visium choice. BANKSY and SmoothClust are useful alternatives when
their assumptions match the data and optional dependencies are available.
Domain labels are unsupervised spatial clusters. They are useful for segmenting
tissue regions, but they need marker genes, histology, or reference labels
before being named biologically.

```r
spatial_bayes <- standard_scop(
  spatial,
  workflow = "spatial",
  assay = "Spatial",
  image = "slice1",
  coord.cols = c("x", "y"),
  do_spot_qc = FALSE,
  do_spatial_variable_features = FALSE,
  do_spatial_cluster = TRUE,
  spatial_cluster_method = "BayesSpace",
  spatial_q = 3
)

SpatialSpotPlot(
  spatial_bayes,
  group.by = "BayesSpace_cluster"
)
```

For other domain backends, keep the output contract in mind: cluster labels
should land in metadata and method details should stay under `srt@tools`.

```r
spatial <- RunBANKSY(
  spatial,
  assay = "Spatial",
  layer = "counts",
  coord.cols = c("x", "y"),
  cluster_colname = "BANKSY_cluster"
)
SpatialSpotPlot(spatial, group.by = "BANKSY_cluster")
```

## Add Cell Composition Only With a Reference

For spot-level data, deconvolution depends more on reference quality than on the
wrapper choice. Use a matched single-cell reference, then inspect dominant cell
types and maximum proportions before interpreting spatial biology.

```r
spatial <- RunSpatialDWLS(
  spatial,
  reference = reference,
  reference_label = "celltype",
  assay = "Spatial",
  reference_assay = "RNA",
  coord.cols = c("x", "y"),
  prefix = "SpatialDWLS"
)

spatial@tools$SpatialDWLS$summary

SpatialSpotPlot(
  spatial,
  group.by = "SpatialDWLS_dominant_type"
)

SpatialSpotPlot(
  spatial,
  group.by = "SpatialDWLS_dominant_type",
  plot_type = "pie"
)
```

`RCTD`, `SPOTlight`, `CARD`, and `STdeconvolve` follow the same practical rule:
first check the result summary, then visualize dominant labels, maximum
proportions, and selected proportions.
Deconvolution estimates cell-type composition per spot. It is constrained by
the reference, so missing reference cell types can be misassigned to the closest
available label.

## Add Neighborhood or Context Models Last

Neighborhood methods answer a different question from domain clustering. Use
them after labels are stable, either from metadata, clustering, or
deconvolution.

```{r spatial-neighborhood}
spatial <- RunSpatialNeighborhood(
  spatial,
  group.by = "coda_label",
  coord.cols = c("x", "y"),
  k = 6
)

spatial@tools$SpatialNeighborhood$summary

SpatialNeighborhoodPlot(spatial, plot_type = "heatmap")
SpatialNeighborhoodPlot(spatial, plot_type = "stat", top_n = 12)
```

```r
SpatialNeighborhoodPlot(spatial, plot_type = "network", top_n = 12)
```

Use `RunStatialKontextual()` when you have explicit `from`, `to`, and `parent`
cell populations. Use `RunMistyR()` when the question is feature-level local or
broader spatial context rather than pairwise label enrichment.
Neighborhood enrichment is label-level spatial association. Context models are
feature-level spatial association. Keep those interpretations separate in
reports.

Use the shared discovery APIs before choosing an optional method. They read one
registry, so the method list, backend diagnosis, result keys, coordinate
contract, and recommended plot cannot drift into separate hard-coded lists.

```r
ListSpatialMethods(task = "neighborhood")
SpatialBackendStatus(method = c("RunSpatialNeighborhood", "RunMistyR"))
SpatialResultInfo(spatial)
```

New small-method results use schema v1 while keeping method-specific fields at
their historical top-level paths. Read a normalized copy through
`GetSpatialResult()` instead of depending on the internal `@tools` layout.
Custom `tool_name` values remain discoverable. Legacy Giotto, SPATA2, and Semla
results are adapted only in the returned copy and are never migrated in place.

```r
coords <- SpatialCoordinates(spatial, image = "slice1", space = "raw")
network_result <- GetSpatialResult(spatial, tool_name = "SpatialNetwork")
network_graph <- GetSpatialGraph(
  res = network_result,
  format = "sparse",
  value = "weight"
)
SpatialResultInfo(spatial, detail = "graphs")
```

Distance-sensitive methods retain `coordinate_space = "legacy_display"` as
the compatibility default in this release. Set `coordinate_space = "raw"`
when radius or distance values must stay in original acquisition units. Plot
functions continue to map results to display coordinates by cell or spot ID.
Objects with multiple spatial images must always select one explicitly with
`image`; analysis, plotting, and framework conversion never silently use the
first image. A future release will make raw coordinates the default for all
distance-sensitive producers.

```r
spatial <- RunMistyR(
  spatial,
  assay = "Spatial",
  layer = "data",
  features = head(spatial@misc$SpatialVariableFeatures, 20),
  coord.cols = c("x", "y"),
  views = "para"
)
spatial@tools$MistyR$summary
MistyRPlot(spatial, type = "improvements")
```

For label-level contextual scores, use
`StatialKontextualPlot(spatial)` after `RunStatialKontextual()`. These dedicated
plots summarize stored backend output and never rerun the analysis.

## What to Report

A compact first-pass spatial report should include:

- the source object, assay, image, and coordinate columns;
- spot or cell counts after filtering;
- top spatial variable features and the method used;
- the domain method, cluster count, and domain sizes if clustering was run;
- deconvolution dominant labels and maximum proportions if composition was run;
- neighborhood or context summaries only after labels are stable.

Keep the main workflow short. Add backend-specific sections only when they
answer a concrete biological question for the current dataset.
