---
title: "Spatial framework bridges and Giotto compatibility"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

scop supports Giotto in two deliberately separate ways. The older workflow is
kept for compatibility, while the newer bridge functions exchange native
framework objects without wrapping every Giotto or SPATA2 analysis step.

## Choose the interface before starting

| Goal | Interface | Returned object | Plotting |
|---|---|---|---|
| Keep an existing scop Giotto pipeline | `SeuratToScopGiotto()`, `RunGiottoWorkflow()`, `GiottoPlot()`, `AddGiottoToSeurat()` | scop's legacy `giotto2` workflow wrapper | `GiottoPlot()` |
| Move one Seurat image into Giotto | `srt_to_giotto()` | Native Giotto object | Giotto's native plotting API |
| Move a native Giotto object into Seurat | `giotto_to_srt()` | New Seurat object | scop or Seurat plotting |
| Move between Seurat and SPATA2 | `srt_to_spata2()`, `spata2_to_srt()` | Native SPATA2 or Seurat object | Framework-native plotting |

Do not mix the two Giotto styles in one pipeline. In particular,
`RunGiottoWorkflow()` expects the legacy scop wrapper, whereas
`srt_to_giotto()` intentionally returns a native Giotto object instead of that
wrapper.

## Giotto version boundary

Current Giotto Suite releases split analysis functions and object/conversion
helpers across the `Giotto` and `GiottoClass` namespaces. scop resolves these
optional functions at runtime and does not add Giotto to Imports, Depends, or
Suggests.

The new reverse bridge selects `giottoToSeuratV5()` for Seurat 5 and
`giottoToSeuratV4()` for Seurat 4. These converters are expected from a current
GiottoClass installation. If an older Giotto installation does not provide
them, update Giotto before using `giotto_to_srt()`.

The legacy workflow remains available for older scripts and is not migrated to
the new bridge result container. This compatibility promise applies to its
public arguments, `giotto2` return structure, stored result names, and
`GiottoPlot()` behavior.

Use `SpatialBackendStatus()` to distinguish installation from API
compatibility. `Giotto` and `GiottoClass` are reported separately because the
runtime dispatcher resolves their symbols separately from the same registry.

```r
SpatialBackendStatus(backend = c("giotto", "giotto_class"), api_check = TRUE)
```

An `available` row means the package and required symbols for that route were
found. `api_incompatible` means the package exists but a required current or
legacy symbol is absent; install the matching Giotto generation rather than
adding it to scop's DESCRIPTION dependencies.

## GitHub documentation build contract

The pkgdown site documents both generations of the Giotto interface. Legacy
topics remain under **Spatial Framework Bridges and Giotto Workflows**, next to
the native bridge topics; they are not aliases for one another and neither set
is removed when the other is updated.

Giotto, GiottoClass, and SPATA2 remain runtime-optional backends. The examples
in this article are deliberately shown but not evaluated during the GitHub
pkgdown build, so publishing the documentation does not select or install a
particular Giotto generation. Reference pages are still parsed and indexed,
and backend/version checks run when a user calls a legacy workflow or a new
bridge. This keeps the site build independent from optional framework releases
while preserving visible documentation for both interfaces.

On pull requests, the GitHub workflow validates the complete reference index.
After a merge to `main`, changes that include source, reference, or pkgdown
configuration files trigger a full site build; an article-only change can use
the narrower article build. The scheduled full build remains the compatibility
check against the then-current dependency ecosystem.

Ordinary package and pkgdown checks remain independent of a user's Giotto
installation. A Giotto compatibility failure must be classified as a scop
regression, an upstream Giotto API regression, or a runner failure. Current API
failures block a spatial release. Legacy API failures do not block unrelated
pull requests, but must be resolved or documented before the spatial feature
freeze is lifted. This keeps old Giotto support visible without making every
documentation build permanently red.

## Single-image conversion

The new bridges process exactly one image at a time:

- a single image is selected automatically;
- a multi-image object requires an explicit `image`;
- an unknown image name is rejected;
- slices are never merged implicitly.

```{r}
gobject <- srt_to_giotto(spatial, image = "slice1")
spatial_roundtrip <- giotto_to_srt(gobject)
```

For exact image selection, `srt_to_giotto()` uses scop's controlled converter
by default and returns only its native Giotto object. Advanced users may pass
`use_official = TRUE`, but the controlled default is more stable across Giotto
versions and multi-image Seurat objects.

## SPATA2 bridge

SPATA2 receives the selected image explicitly through its `img_name`
interface. The bridge does not run trajectories, spatial screening, or SPATA2
plotting.

SPATA2 3.1.4 exports `asSeurat()` as a development stub rather than a complete
converter. scop still calls the official function first; when it does not
return a Seurat object, the compatibility fallback transfers counts, metadata,
and raw coordinates into a new Seurat object. It does not attempt to reproduce
SPATA2 analysis slots or framework plots.

```{r}
spata_object <- srt_to_spata2(spatial, image = "slice1")
spatial_roundtrip <- spata2_to_srt(spata_object)
```

Use SPATA2's native API after conversion. scop only owns image validation,
single-image isolation, dependency checks, and object exchange.
